### 2023-06-06 README.Rmd
initial commit

---
title: "Supplementary materials for: How many speech sounds are there in the world's languages?"
author: "Steven Moran & Nicholas A. Lester\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
      toc: true
bibliography: 'bibliography.bib'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=F, warning=F, message=F, cache=T)
```

# Overview

All code is written in the R the programming language [@R]. This report uses the following R libraries  [@tidyverse;@lmerTest;@testthat;@ggpubr].

```{r}
# Load the libraries
library(tidyverse)
library(lmerTest)
library(testthat)
library(ggpubr)
```

This report is organized as follows:

* First, we describe the data, which consists of two (nested) databases of phonological inventories: UPSID [@Maddieson1984;@MaddiesonPrecoda1990] and PHOIBLE [@MoranMcCloy2019] -- UPSID is included in PHOIBLE.

* Second, we show that both the UPSID and PHOIBLE databases' segment type distributions follow Zipf's law [@Zipf1936;@Zipf1949]. This step helps us to motivate a connection to Herdan-Heaps' law (described next).

* Third, we evaluate the segment distributions in each database against Herdan-Heaps' law [@Herdan1964;@Heaps1978] (replacing the original standard document count with language count, and word type count with segment type count).

* Fourth, we apply a different estimator of type counts, which we refer to as the Chao-Shen estimator [@ChaoChiu2016], which was developed to estimate the richness of species from limited sample sizes.

* Fifth, we check how each estimator performs when predicting our largest known sample (PHOIBLE) from subsets of languages.

* Sixth, we extrapolate to unseen numbers of languages using both Herdan-Heaps' and the Chao-Shen estimator to estimate how many segment types we might expect in increasingly larger samples of languages.

* Seventh, we compare the results from Herdan-Heaps' and the Chao-Shen estimator and show that the evidence suggests that there is no upper bound to the number segment types (phonemes) we may expect to find. That is, ever increasing linguistic diversity leads to an ever increasing number of phonemes in light of linguistic laws.

# Data

For this study, we use the latest [PHOIBLE data](https://phoible.org) [@MoranMcCloy2019], which includes 3020 phonological inventories from 2186 distinct languages. The dataset is a convenience sample of languages, in which some geographic regions and language families have greater or lesser coverage. Included in PHOIBLE is UPSID-451 [@Maddieson1984;@MaddiesonPrecoda1990], a genealogically balanced sample of languages in which each data point was selected as a representative of its language family. We use both UPSID and PHOIBLE as baselines to evaluate our models.

We read the PHOIBLE and UPSID data from the [GitHub repository](https://github.com/phoible/dev).

```{r}
#phoible <-  read_csv(url("https://github.com/phoible/dev/blob/646f5e4f64bfefb7868bf4a3b65bcd1da243976a/data/phoible.csv?raw=true"), col_types = c(InventoryID = "i", Marginal = "l", .default = "c"))

phoible = read.csv("./phoible.csv")
```

Our central question concerns how the number of spoken languages that are documented relate to the number of unique phonemes that are observed. This is a question of cumulative growth. It can be rephrased in particulars:

* "If I have seen *x* languages with *y* unique phonemes, how many unique phonemes do I expect to find in *x* + *i* languages?"

Here is a function to get cumulative segment type counts.

```{r}
# Arguments 
# - df: dataframe of segments (i.e., PHOIBLE or UPSID)
get_cumulative_segment_counts_df <- function(df) {
  results <- numeric()
  encountered <- character()
  for (i in unique(df$InventoryID)) {
    tmp <- df %>% filter(InventoryID == i)
    encountered <- union(encountered, tmp$Phoneme)
    results <- c(results, length(encountered))
  }
  results <- as.data.frame(results)
  results <- results %>% mutate(id = row_number())
  return(results)
}
```

Here is code to randomly sample from PHOIBLE by [ISO 639-3](https://iso639-3.sil.org) code. This step allows us to avoid having multiple inventories for a single language because whereas [Glottocodes](https://glottolog.org/glottolog/glottologinformation) in PHOIBLE provide detail on [doculects](https://phoible.org/faq#inventories-language-codes-doculects-and-sources) [@CysouwGood2013] and ISO 639-3 codes, the latter uniquely identify languages. Language code mappings are available [online](https://github.com/phoible/dev/blob/master/mappings/InventoryID-LanguageCodes.csv).

```{r}
set.seed(1)

nas <- phoible %>%
        distinct(InventoryID, ISO6393) %>%
        filter(is.na(ISO6393)) %>%
        pull(InventoryID)

phoible %>%
  distinct(InventoryID, ISO6393) %>%
  drop_na(ISO6393) %>%
  group_by(ISO6393) %>%
  sample_n(1) %>%
  pull(InventoryID) ->
inventory_ids_sampled_one_per_isocode

inventory_ids_sampled_one_per_isocode <- c(inventory_ids_sampled_one_per_isocode, nas)

phoible %>%
  filter(InventoryID %in% inventory_ids_sampled_one_per_isocode) ->
my_sample
```

Here is code to generate the data for the cumulative segment type data sample.

```{r}
# Extract only the necessary columns
phoible_long <- my_sample %>% select(InventoryID, Phoneme)

# Randomize the order of inventory IDs
set.seed(1)
ids <- unique(phoible_long$InventoryID) %>% sample()

# Reorder the database using the new order of inventory IDs
phoible_long <- phoible_long %>% arrange(match(InventoryID, ids))

# Get the cumulative frequency estimates
phoible_cumulative <- get_cumulative_segment_counts_df(phoible_long)
```

What does it look like for just UPSID-451, which is a genealogically-balanced sample? First we extract just the UPSID inventories.

```{r}
upsid <- phoible %>% filter(Source == "upsid")
expect_equal(length(unique(upsid$InventoryID)), 451)
```

Next we generate the UPSID-451 dataframe for plotting.

```{r}
set.seed(1)

# Extract only the necessary columns
upsid_long <- upsid %>% select(InventoryID, Phoneme)

# Randomize the order of inventory IDs
ids <- unique(upsid_long$InventoryID) %>% sample()

# Reorder the database using the new order of inventory IDs
upsid_long <- upsid_long %>% arrange(match(InventoryID, ids))

# Get the cumulative frequency estimates
upsid_cumulative <- get_cumulative_segment_counts_df(upsid_long)
```

Plot it.

```{r cumulative_plot_upsid}
qplot(id, results, data = upsid_cumulative, xlab="Number of languages (random order)", ylab="Cumulative number of segment types") + theme_bw()
```

Now we plot the full PHOIBLE language sample. Note that the shape is similar, but the x-axis contains many more languages.

```{r, cumulative_plot_phoible}
qplot(id, results, data = phoible_cumulative, xlab="Number of languages (random order)", ylab="Cumulative number of segment types") + theme_bw()
```

The jump in y-axis on both plots is [!Xoo](https://phoible.org/inventories/view/1379), a language reported in UPSID that contains many click consonants unique to this language and not found in any other language in the sample.^[See @Naumann2013 for a reassessment of the number of phonemes in !Xoo.]

We can also plot the estimates against each other.

```{r, cumulative_plot_phoible_upsid}
# Join the dataframes
pho.and.ups <- bind_rows(phoible_cumulative %>% 
                         mutate(Database = as.factor(rep("phoible", nrow(.)))),
                         upsid_cumulative %>%
                         mutate(Database = as.factor(rep("upsid", nrow(.)))))

ggplot(pho.and.ups, aes(x = id, y = results, color = Database)) +
    geom_point() +
    xlab("Number of languages (random order)") +
    ylab("Cumulative number of segment types") +
    theme_bw() + 
    scale_color_discrete(labels = c("PHOIBLE (OIPL)", "UPSID-451"))
```

It looks like the curve for UPSID is slightly shallower than that for PHOIBLE. The curves appear to diverge exponentially. The more diverse sample -- PHOIBLE -- not only produces higher type counts; it also suggests a generally more pronounced rate of increase.

Crucially, however, all the curves plotted above resemble Herdan-Heaps' curve -- a function describing the relationship between number of samples and the number of distinct types discovered in those samples.

Herdan-Heaps' law can be derived from the Mandelbrot distribution (of which Zipf is a special case). As a first step, we test whether the frequency of segments in our databases follows a Zipfian distribution.

# Analyses

## Zipf's law

Some code in this selection is based on the [report](https://rstudio-pubs-static.s3.amazonaws.com/215309_736f5cc00eea4bb9be5a8c566da2beb6.html) by @BastrakovaGarcia2016.

Does the segment distribution in the UPSID-451 and PHOIBLE sample of languages follow [Zipf's law](https://en.wikipedia.org/wiki/Zipf%27s_law)?

First we extract the segments and their frequencies.

```{r}
segments_upsid <- phoible %>%
  filter(Source == "upsid") %>%
  group_by(Phoneme) %>%
  summarize(Frequency = n()) %>%
  arrange(desc(Frequency))

segments_phoible <- phoible %>%
  group_by(Phoneme) %>%
  summarize(Frequency = n()) %>%
  arrange(desc(Frequency))
```

Next we add their rank.

```{r}
segments_upsid$Rank <- 1:nrow(segments_upsid)
segments_phoible$Rank <- 1:nrow(segments_phoible)
```

And we plot the segment type frequency distributions.

```{r}
plot(segments_upsid$Rank, segments_upsid$Frequency, xlab = "Rank", ylab = "Frequency", main = "UPSID-451 segment distribution")
plot(segments_phoible$Rank, segments_phoible$Frequency, xlab = "Rank", ylab = "Frequency", main = "PHOIBLE phoneme distribution")
```

We can also remove all segments with a frequency of one and look more closely at the distribution.

```{r}
segments_gt_1_upsid <- segments_upsid %>% filter(Frequency > 1)
plot(segments_gt_1_upsid$Rank, segments_gt_1_upsid$Frequency, xlab = "Rank", ylab = "Frequency", main = "UPSID segment distribution")

segments_gt_1_phoible <- segments_phoible %>% filter(Frequency > 1)
plot(segments_gt_1_phoible$Rank, segments_gt_1_phoible$Frequency, xlab = "Rank", ylab = "Frequency", main = "PHOIBLE segment distribution")
```

Next we plot log(rank) against log(frequency) and we see that the segment frequency distribution nicely follows Zipf's law.

```{r, zipf_distribution_upsid}
x <- log(segments_gt_1_upsid$Rank)
y <- log(segments_gt_1_upsid$Frequency)
df <- data.frame(x, y)

x_rank <- segments_gt_1_upsid$Rank

plot(x, y, xlab = "Rank (log)", ylab = "Frequency (log)", main = "Zipf's law on UPSID-451")

zipf <- function(rank, a, b, c) {
  return(log(c) + a * log(rank + b))
}

# Choosing the best parameters visually
# lines(x, zipf(x_rank, -1.3, 18, 1000000), col = "green", lwd = 3)

# Estimating the parameters -- TODO have set c lower
fitted <- nls(y ~ zipf(x_rank, a, b, c), data = df, start = list(a = -1.3, b = 18, c = 100000), trace = T)

# Plotting the Zipf's law with the parameters estimated
lines(x, predict(fitted), lty = 1, col = "lightblue", lwd = 4)

Zipfs_coef <- coef(fitted)
```

```{r, zipf_distribution_phoible}
x <- log(segments_gt_1_phoible$Rank)
y <- log(segments_gt_1_phoible$Frequency)
df <- data.frame(x, y)

x_rank <- segments_gt_1_phoible$Rank

plot(x, y, xlab = "Rank (log)", ylab = "Frequency (log)", main = "Zipf's law on PHOIBLE (OIPL)")

zipf <- function(rank, a, b, c) {
  return(log(c) + a * log(rank + b))
}

# Choosing the best parameters visually
# lines(x, zipf(x_rank, -1.3, 18, 1000000), col = "green", lwd = 3)

# Estimating the parameters
fitted <- nls(y ~ zipf(x_rank, a, b, c), data = df, start = list(a = -1.3, b = 18, c = 1000000), trace = T)

# Plotting the Zipf's law with the parameters estimated
lines(x, predict(fitted), lty = 1, col = "lightblue", lwd = 4)

Zipfs_coef <- coef(fitted)
```

The distributions of phonemes reported in both databases are strongly Zipfian.

We now test whether the distribution likewise follows Herdan-Heaps' law.

## Herdan-Heaps' law

What about the UPSID-451 and PHOIBLE segments and whether their distributions follows [Herdan–Heaps' law](https://en.wikipedia.org/wiki/Heaps%27_law)?

Here is a function to get the cumulative number of segment types versus tokens.

```{r}
# Arguments:
#
# df:   long-form phoneme dataframe 
#       (Inventory ID | Phoneme)
# seed: (optional) set.seed value for random 
#       reshuffling of inventories

get_heaps_df <- function(df, seed = 1) {

  inv <- unique(df$InventoryID)

  # Shuffle inventories
  set.seed(seed)
  rand.inv <- sample(inv)

  # Reorder the df    
  df <- df %>% arrange(match(InventoryID, rand.inv))

  # Get the inventory sizes of each language
  inv.sizes <- df %>%
               group_by(InventoryID) %>%
               summarize(inv.size = n()) %>%
               arrange(match(InventoryID, rand.inv)) %>%
               select(inv.size) %>%
               pull()

  # Compute the cumulative sums of the inventory sizes
  cum.tokens <- cumsum(inv.sizes)

  # Get the number of unseen (so far) types per inventory
  types <- df %>%
           mutate(index = seq(1, n())) %>%
           mutate(phon.seen = ifelse(index %in% match(unique(Phoneme), Phoneme),
                                     1, 0)) %>%
           group_by(InventoryID) %>%
           summarize(new.types = sum(phon.seen)) %>%
           arrange(match(InventoryID, rand.inv)) %>%
           select(new.types) %>%
           pull()

  # Compute the cumulative sum of type counts
  cum.types <- cumsum(types)

  # Combine the results
  output <- bind_cols(index = seq(1, length(cum.types)),
                      token_count = cum.tokens,
                      type_count = cum.types)

  return(output)
}
```

The equation for Herdan–Heaps' law is: `d = k * n ^B`. In our case `d` is the number of distinct segment types in the sample of languages and `n` is the total number of segment tokens.

```{r}
heaps <- function(K, n, B) {
  K * n^B
}

# Get the total number of observations from PHOIBLE
n <- sum(segments_phoible$Frequency)

# Settings of K and B based on trial and error -- not the optimal values, but the result is reasonably close to the true value. Therefore, if we do any model tuning, the values we arrive at should be close to these; otherwise, we most likely have a bad fit of Herdan-Heaps' to the data. 
heaps(8, n, .5)
```

An important feature of the Herdan-Heaps' equation is that it can be expressed in the form of a standard linear equation. Simply take the logarithm of each term:

<br><center><span style="font-family:'Times New Roman'">log <i>d</i> = log <i>K</i> + <i>B</i>&sdot; log <i>n</i></span></center></br>

which has the familiar form:

<br><center><span style="font-family:'Times New Roman'"><i>y</i> = <i>&Beta;</i><sub>0</sub> +  <i>&Beta;</i><sub>1</sub> &sdot; <i>x</i></span></center>

The upshot is that we can fit a (log-transformed) linear regression model to the data and:

- The resulting intercept is the optimal *K* (with some back-transformation: *K* = *e*<sup>*K*</sup>).

- The coefficient &beta; expresses *B*.

With these values, we can estimate the type count for any sample size *n* using the standard Herdan-Heaps' equation.

In the following, we apply this function to PHOIBLE and UPSID in order to estimate the number of phonemes for any sample size of *n* languages.

First, recall that we have already generated the filtered and sampled long data formats of counts for both UPSID-451 and PHOIBLE.

```{r}
head(upsid_long)
```

Let's get the cumulative counts of the segment types and tokens for several randomized orders of inventories.

First, a function to randomly permute the order of inventories before collecting the input for the Herdan-Heaps' analysis.

```{r}
rand.heaps.data <- function(df, k){
  data.list <- vector(mode="list", length = k)
  for(i in 1:k){
    curr.data <- get_heaps_df(df, i)
    curr.data$iteration <- as.factor(rep(i, nrow(curr.data)))
    data.list[[i]] <- curr.data
  }

  output = do.call(rbind, data.list)

  return(output)
}
```

Now apply the function *k* times. Here, we set *k*=10.

```{r}
heaps_results_upsid <- rand.heaps.data(upsid_long, 10)
heaps_results_phoible <- rand.heaps.data(phoible_long, 10)
```

### UPSID

Now we fit linear regressions for UPSID-451 and PHOIBLE to generate (empirically) optimal values for *B* and *K* (our two free parameters in the Herdan-Heaps' equation).

```{r}
fitHeaps.upsid <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data = heaps_results_upsid)

logK.upsid <- fixef(fitHeaps.upsid)[1]
B.upsid <- fixef(fitHeaps.upsid)[2]
K.upsid <- exp(logK.upsid)

# Get predicted values
hyp.data <- data.frame(token_count = seq(1:nrow(upsid_long)),
                      iteration = rep(1, nrow(upsid_long)))

preds.hyp <- exp(predict(fitHeaps.upsid, hyp.data, re.form=NA))
```

Here we plot the predicted values against the true values.

```{r heaps_upsid}
plot.dat <- bind_cols(hyp.data, Predicted = preds.hyp)

heaps_upsid_p = ggplot(heaps_results_upsid, 
                       aes(x=token_count,
                           y=type_count,
                           color=iteration)) +
                        geom_point(alpha=0.07) +
                        geom_line(data=plot.dat, 
                                  aes(x=token_count,
                                      y=Predicted),
                                  color="black", lwd=1) +
                        ylab("Type count") +
                        xlab("Token count") +
                        theme_bw() +
                        ggtitle("UPSID-451") +
                        theme(legend.position="none",
                              plot.title = element_text(hjust=0.5))

heaps_upsid_p
```

The UPSID model fits quite nicely, suggesting that the distribution follows Herdan-Heaps' law.

### PHOIBLE (one-inventory-per-language)

Now for PHOIBLE, we begin with the "one-inventory-per-language" version of the database.

```{r}
fitHeaps.phoible <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data = heaps_results_phoible)

logK.phoible<- fixef(fitHeaps.phoible)[1]
B.phoible <- fixef(fitHeaps.phoible)[2]
K.phoible <- exp(logK.phoible)

# Get predicted values
hyp.data.pho <- data.frame(token_count = seq(1:nrow(phoible_long)),
                      iteration = rep(1, nrow(phoible_long)))

preds.hyp.pho <- exp(predict(fitHeaps.phoible, hyp.data.pho, re.form=NA))
```

And plot the predicted values against the true values.

```{r heaps_phoible_oipl}
plot.dat.pho <- bind_cols(hyp.data.pho, Predicted = preds.hyp.pho)

heaps_phoible_p = ggplot(heaps_results_phoible, 
                       aes(x=token_count,
                           y=type_count,
                           color=iteration)) +
                        geom_point(alpha=.01) +
                        geom_line(data=plot.dat.pho, 
                                  aes(x=token_count,
                                      y=Predicted),
                                  color="black", lwd=1) +
                        ylab("Type count") +
                        xlab("Token count") +
                        theme_bw() +
                        ggtitle("PHOIBLE (OIPL)") +
                        theme(legend.position="none",
                              plot.title = element_text(hjust=0.5))

heaps_phoible_p
```

Again, we find a very nice fit. Even though the observed cumulative distributions of UPSID-451 and PHOIBLE have different shapes, the Herdan-Heaps' equation produces quality predictions for both.

### PHOIBLE (all)

Let's try the same with the entirety of the PHOIBLE language sample, including multiple inventories for some languages.

```{r, message=F}
# Get the segment frequencies from full PHOIBLE
freq <- phoible %>%
  group_by(Phoneme) %>%
  summarize(Frequency = n()) %>%
  arrange(desc(Frequency))

num_langs <- nrow(phoible %>% select(InventoryID) %>% distinct())
```

Generate the long-form table.

```{r}
set.seed(1)

# Select the relevant columns
phoible_long_all <- phoible %>% select(InventoryID, Phoneme)

# Randomize the order of inventories
ids <- unique(phoible_long_all$InventoryID) %>% sample()

# Reorder the database according to the new random order
phoible_long_all <- phoible_long_all %>% arrange(match(InventoryID, ids))

# Generate 10 such orderings 
heaps_results_phoible_all <- rand.heaps.data(phoible_long_all, 10)
```

Apply the Herdan-Heaps' model.

```{r}
# Fit the model
fitHeaps.phoible.all <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data = heaps_results_phoible_all)

# Get the values for our parameters
logK.phoible.all <- fixef(fitHeaps.phoible.all)[1]
B.phoible.all <- fixef(fitHeaps.phoible.all)[2]
K.phoible.all <- exp(logK.phoible.all)

# Get predicted values
hyp.data.pho.all <- data.frame(token_count = seq(1:nrow(phoible_long_all)),
                      iteration = rep(1, nrow(phoible_long_all)))

preds.hyp.pho.all <- exp(predict(fitHeaps.phoible.all, hyp.data.pho.all, re.form=NA))
```

Plot the predicted values against the true values.

```{r heaps_phoible_all}
plot.dat.pho.all <- bind_cols(hyp.data.pho.all, Predicted = preds.hyp.pho.all)

heaps_phoible_all_p = ggplot(heaps_results_phoible_all, 
                       aes(x=token_count,
                           y=type_count,
                           color=iteration)) +
                        geom_point(alpha=0.009) +
                        geom_line(data=plot.dat.pho.all, 
                                  aes(x=token_count,
                                      y=Predicted),
                                  color="black", lwd=1) +
                        ylab("Type count") +
                        xlab("Token count") +
                        theme_bw() +
                        ggtitle("PHOIBLE (all)") +
                        theme(legend.position="none",
                              plot.title = element_text(hjust=0.5))

heaps_phoible_all_p
```

This plot replicates the other PHOIBLE-based plot, but on the entire dataset.

### Comparing model parameters across databases

Let's compare the parameter estimates across the databases.

```{r}
params = data.frame(Database = c("UPSID-451",
                                 "PHOIBLE (OIPL)",
                                 "PHOIBLE (all)"),
                    B = c(B.upsid, B.phoible, B.phoible.all),
                    K = (c(K.upsid, K.phoible, K.phoible.all)))

params

```

The values are relatively similar. Interestingly, UPSID-451 and PHOIBLE (all) have nearly identical values for *B*. They differ in the scalar *K*, which explains the sharper increases in values for PHOIBLE.

If we combine the above plots on the same scale, we get a picture of how these differences in parameters affect the shape of our results.

```{r compare_models_all}
plot.dat.all <- bind_rows(plot.dat %>% 
                            mutate(Source = "UPSID-451"), 
                          plot.dat.pho %>% 
                            mutate(Source = "PHOIBLE (OIPL)"), 
                          plot.dat.pho.all %>% 
                            mutate(Source = "PHOIBLE (all)"))

heaps_comp_p = ggplot(data=plot.dat.all, 
                       aes(x=token_count,
                           y=Predicted, 
                           color=Source)) +
                        geom_line(lwd=0.8) +
                        ylab("Type count") +
                        xlab("Token count") +
                        theme_bw()

heaps_comp_p
```

The PHOIBLE estimates behave unsurprisingly similarly. The growth curve for UPSID-451 is again shallower; however, it does not appear to be too far off.

Now only plotting OIPL vs. UPSID-451 (for the manuscript)

```{r compare_models}
plot.dat.oipl <- plot.dat.all %>%
                 filter(Source != "PHOIBLE (all)")

heaps_comp_oipl = ggplot(data=plot.dat.oipl, 
                       aes(x=token_count,
                           y=Predicted, 
                           color=Source)) +
                        geom_line(lwd=0.8) +
                        ylab("Type count") +
                        xlab("Token count") +
                        theme_bw()

heaps_comp_oipl
```

We now explore an alternative, generally more flexible estimator of type counts: the Chao-Shen estimator [@Chao_etal2009;@ChaoChiu2016]. If the results converge, then we have some foundation for trusting extrapolated values.

## The Chao-Shen estimator

The central equation for the Chao-Shen estimator is given below [@Chao_etal2009]. The central parameter for extrapolation is *r*, which expresses the number of additional "sampling units" (in our case, languages) we expect to encounter.

![Chao et al., 2009](figures/image.png)

The Chao-Shen estimator takes an **incidence matrix** as input. An incidence matrix is a record of the presence of absence of some type of thing in each of a set of sampling units. Imagine that you are looking to estimate the number of types of beetles in a forest. You set up a number of traps around the forest to get a (hopefully) representative sample of the overall ecosystem. The traps are the sampling units. The species of beetles are the types. You visit each trap, and for each species of beetle you find there, you record a 1 for that species in that trap.

The same logic can be applied to languages and phonemes. Languages are our beetle traps and phonemes are the beetles. For each sound out of all known sounds, we record a 1 if it occurs in a given language and a 0 otherwise. Here is an incidence matrix for an invented dataset consisting of 150 total observations of <i>S</i><sub>obs</sub> phonemes from *R* languages:

*Example incidence matrix for invented dataset*

|Phoneme|Lang<sub>1</sub>|Lang<sub>2</sub>|&middot;&middot;&middot;|Lang<sub>*R*</sub>|&sum;<sub>phoneme</sub>|
|:---:|---|---|:---:|---|---|
|**Phon<sub>1</sub>**|1|0|&middot;&middot;&middot;|0|1|
|**Phon<sub>2</sub>**|0|1|&middot;&middot;&middot;|1|2|
|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|
|**Phon<sub>*S*<sub>obs</sub></sub>**|1|1|&middot;&middot;&middot;|1|30|
|**&sum;<sub>language</sub>**|15|23|&middot;&middot;&middot;|6|150|

- ***R*** is the number of sampling units, in this case the number of languages in our sample. 
- ***S*<sub>obs</sub>** is the number of types that *have* been observed (i.e., the number of sounds which were observed in at least one language).
- **&sum;<sub>language</sub>** is the phonological inventory size per language.
- **&sum;<sub>phoneme</sub>** is the frequency of occurrence of the sounds across languages.

Note that Phon<sub>1</sub> appears only once in the sample; it is a **singleton**. The total number singleton phonemes is ***Q*<sub>1</sub>**. Similarly, Phon<sub>2</sub> appears only twice; it is a **duplicate** (in the terminology of Gotelli and Chao, 2013). The total number of duplicate phonemes is ***Q*<sub>2</sub>**.

@GotelliChao2013 functions coded in R by Nicholas Lester:

```{r}
# args (hold the same for both equations)
# - vector = frequency vector, cells = PHOIBLE frequency of a sound
# - R = number of sampling sites (i.e., # of languages in the sample)
# - r = number of hypothetical additional sampling sites (languages)

# Function to get extrapolated estimates of types
Extrapolate <- function(vector, R, r) {
  Sest <- S_Chao2(vector, R)
  Sobs <- sum(vector > 0)
  Q1 <- sum(vector == 1)
  Q0 <- Sest - Sobs
  if (Q0 > 0) {
    return(Sobs + Q0 * (1.0 - (1.0 - Q1 / (Q1 + R * Q0))**r))
  }
  else {
    return(Sobs)
  }
}

# Function to estimate the number of unseen types from a proposed assemblage (used to compute our estimate of Q0)
S_Chao2 <- function(vector, R) {
  fd <- vector[vector > 0]
  Q1 <- sum(vector == 1)
  Q2 <- sum(vector == 2)
  Sobs <- length(fd)
  if (Q2 > 0.0) {
    res <- Sobs + ((R - 1.0) / R) * ((Q1**2.0) / (2.0 * Q2))
  }
  else {
    res <- Sobs + ((R - 1.0) / R) * Q1 * (Q1 - 1) / 2.0
  }
  return(res)
}
```

The *vector* argument should be a vector of frequencies (in this case, a vector of the frequencies of phonemes, where each cell represents the frequency of an individual phoneme in PHOIBLE). Note that this amounts to the row-wise sums of the incidence matrices described above. *R* is the number of languages in the sample. *r* is a free parameter. It refers to the number of additional **hypothetical** draws that we want to allow our estimator. The output is the expected number of types in the (fake) larger sample.

### Relationship between Chao-Shen estimates and *r*

Here we examine how the Chao-Shen estimate grows in response to increase in *r* by:

- Picking a number of inventories.
- Computing the frequency vector.
- Computing R.
- Computing r.

Here is a function to predict type count based on iterative increases of the number of observed languages *s*, where this process is repeated *k* times.

```{r, message=F}
# Arguments
#
# df:       long-form database (e.g., ny_sample)
# r:        how many additional languages to observe 
# k:        how many times to shuffle and the data and perform
#           the iterative estimates
# step:     how many languages to sample at a time
iterative.chao.estimator <- function(df, r, k, step){
  # vector of the unique inventories in the sample
  invs <- unique(df$InventoryID)
  # number of unique inventories in the sample
  num.invs <- length(invs)
  # list to store iterations
  iter.list <- vector(mode="list", length = k)
  # Iterate the process k times
  for(i in 1:k){
    set.seed(i)
    # randomly reorder the inventories
    rand.inds <- sample(invs)
    # reorder the dataframe by the inventories
    df <- df %>%
          arrange(match(InventoryID, rand.inds))
    # The target inventory sizes (what we want estimates for)
    target.sizes <- c(seq(1, num.invs, step), num.invs) 
    # Vector to store the Chao estimates
    iter.chao <- vector(length = length(target.sizes))
    # Vector to store Rs
    Rs <- vector(length = length(target.sizes))
    # Vector to store the total number of 
    # observations (i.e., n from the Herdan-Heaps equation)
    num.obs <- vector(length = length(target.sizes))
    # Vector to store the number of observed types
    num.typ.obs <- vector(length = length(target.sizes))
    # for each target size of languages to estimate
    # type counts for... (note that we can only go
    # to num.invs - r languages to generate estimates
    # in order to get a true baseline).
    for(j in 1:length(target.sizes)){
      # Get R
      R <- target.sizes[j]
      # Get the inventories to use as basis for estimate
      basis.invs <- rand.inds[1:R]
      obs.invs <- rand.inds[1:(R+r)]
      vec <- df %>%
             filter(InventoryID %in% basis.invs) %>%
             group_by(Phoneme) %>%
             summarize(frequency = n(), .groups = 'drop')
      obs.typs <- df %>%
                  filter(InventoryID %in% obs.invs) %>%
                  select(Phoneme) %>%
                  distinct() %>%
                  summarize(type_count = n(), .groups = 'drop') %>%
                  pull()
      num.obs[j] <- sum(vec$frequency)
      num.typ.obs[j] <- obs.typs
      iter.chao[j]  <- Extrapolate(vec$frequency, R, r)
      Rs[j] <- R
    }
    chao.data <- bind_cols(Estimate = iter.chao, 
                            Sample.Size = Rs, 
                            Obs.Size = num.obs,
                            True.Type.Count = num.typ.obs,
                            Iteration = i,
                            r = r)

    diffs <- ifelse((chao.data$Sample.Size + chao.data$r) < num.invs,
                    chao.data$Estimate - chao.data$True.Type.Count,
                    NA)

    chao.data$Difference <- diffs

    chao.data$Iteration <- as.factor(chao.data$Iteration)

    iter.list[[i]] <- chao.data
  }
  output <- do.call(rbind, iter.list)

  return(output)
}

chao.data.10.10 <- iterative.chao.estimator(my_sample, 10, 10, 10)
chao.data.50.10 <- iterative.chao.estimator(my_sample, 50, 10, 10)
chao.data.100.10 <- iterative.chao.estimator(my_sample, 100, 10, 10)
chao.data.500.10 <- iterative.chao.estimator(my_sample, 500, 10, 10)
chao.data.1000.10 <- iterative.chao.estimator(my_sample, 1000, 10, 10)
chao.data.1500.10 <- iterative.chao.estimator(my_sample, 1500, 10, 10)
chao.data.2000.10 <- iterative.chao.estimator(my_sample, 2000, 10, 10)

```

Next we plot the results.

First, we get mean estimates of (a) number of observations, (b) estimated type count, and (c) mean difference (where available) between estimated and observed type count.

```{r}
# function to generate mean estimates, etc.
# Arguments
#
# df:   dataframe of estimates Chao-Shen values (output of 
#       iterative.chao.estimator())
mean.obs.fnc <- function(df){
  new.df <- df %>%
            group_by(Sample.Size) %>%
            summarize(mean.typ = mean(True.Type.Count),
                      mean.est = mean(Estimate),
                      mean.diff = mean(Difference, na.rm=T),
                      mean.obs = mean(Obs.Size),
                      r = r) %>%
            distinct()

  return(new.df)

}

# Get means for each of the sizes of r
mean.obs.df.10 <- mean.obs.fnc(chao.data.10.10) 
mean.obs.df.50 <- mean.obs.fnc(chao.data.50.10) 
mean.obs.df.100 <- mean.obs.fnc(chao.data.100.10) 
mean.obs.df.500 <- mean.obs.fnc(chao.data.500.10) 
mean.obs.df.1000 <- mean.obs.fnc(chao.data.1000.10) 
mean.obs.df.1500 <- mean.obs.fnc(chao.data.1500.10) 
mean.obs.df.2000 <- mean.obs.fnc(chao.data.2000.10) 
```

Next the plots.

```{r}
# Function to plot the iterative Chao-Shen estimates
# Arguments
#
# chao.df:    dataframe containing iterative Chao-Shen estimates
#             (output of iterative.chao.estimator())
# mean.df:    dataframe with mean estimates of Chao-Shen
#             across iterations
# title:      plot title

choa.iter.plotter <- function(choa.df, mean.df, title){
  p <- ggplot(choa.df, aes(x=Sample.Size, 
                           y=Estimate, 
                           color=Iteration)) +
                  geom_line() +
                  scale_color_manual(values = rep("grey70", 10)) + 
                  geom_line(data = na.omit(mean.df), aes(y=mean.typ, 
                                                x=Sample.Size), 
                            col="red", lwd=1) +
                  geom_line(data = mean.df, aes(y=mean.est, 
                                                x=Sample.Size),
                            col="dodgerblue", lwd=1) +
                  ylab("Estimated number of types") + 
                  xlab("Number of languages") +
                  theme_bw() +
                  ggtitle(title) +
                  theme(legend.position="none",
                        plot.title = element_text(hjust=0.5))

  return(p)

}

# r = 10
chao.iter.plot.10 <- choa.iter.plotter(chao.data.10.10, mean.obs.df.10, "r = 10") + ylim(0,4000)

chao.iter.plot.10

# r = 50
chao.iter.plot.50 <- choa.iter.plotter(chao.data.50.10, mean.obs.df.50, "r = 50") + ylim(0,4000)

chao.iter.plot.50

# r = 100
chao.iter.plot.100 <- choa.iter.plotter(chao.data.100.10, mean.obs.df.100, "r = 100") + ylim(0,4000)

chao.iter.plot.100

# r = 500
chao.iter.plot.500 <- choa.iter.plotter(chao.data.500.10, mean.obs.df.500, "r = 500") + ylim(0,4000)

chao.iter.plot.500

# r = 1000
chao.iter.plot.1000 <- choa.iter.plotter(chao.data.1000.10, mean.obs.df.1000, "r = 1000") + ylim(0,4000)

chao.iter.plot.1000

# r = 1500
chao.iter.plot.1500 <- choa.iter.plotter(chao.data.1500.10, mean.obs.df.1500, "r = 1500") + ylim(0,4000)

chao.iter.plot.1500

# r = 2000
chao.iter.plot.2000 <- choa.iter.plotter(chao.data.2000.10, mean.obs.df.2000, "r = 2000") + ylim(0,4000)
chao.iter.plot.2000
```

```{r, chao_lang_types}
increm.chao.plots <- ggarrange(chao.iter.plot.10 +
                                 ylab("") +
                                 xlab(element_blank()),
                               chao.iter.plot.50 + 
                                 ylab("") +
                                 xlab(element_blank()),
                               chao.iter.plot.100 + 
                                 ylab("") +
                                 xlab(element_blank()),
                               chao.iter.plot.500 +
                                 ylab("") +
                                 xlab(element_blank()),
                               chao.iter.plot.1000 + 
                                 ylab("") + 
                                 xlab(element_blank()), 
                               chao.iter.plot.1500 + 
                                 ylab("") + 
                                 xlab(element_blank()),
                               chao.iter.plot.2000 +
                                 ylab("") + 
                                 xlab(element_blank()), 
                               ncol=3,
                               nrow=3)

annotate_figure(increm.chao.plots, bottom = "Number of languages (R)", left = "Estimated number of types")
```

In the above plots, the red lines represent the mean of the true number of observed types per number of languages *R*, across iterations. The blue lines represent the same taken over the Chao-Shen estimates. Grey lines represent the actual Chao-Shen estimates for each iteration. The panels reflect the different values of *r* tested here.

(Note that the red lines become shorter as *r* increases; this is because we have fewer true values to compare against when we look further ahead per calculation).

The results are highly consistent, with near-perfect prediction for lower values of *r*.  Moreover, the variability in the estimates remains stable across lower sample sizes, but decreases gradually as the ratio of *R* to *r* increases. The most stable estimates for these sample sizes begin to surface around 1000 languages. This appears to be consistent across values of *r*.

As *r* increases (holding *R* constant), so does the degree of underestimation.  This discrepancy appears as the increasing difference between red and blue lines as *r* increases, particularly for smaller values of *R*.

Unfortunately, we are constrained by the sample size. Our true values of *r* will be much larger than those we are able to examine empirically (at least twice what our highest value is here).

But if we look carefully (and do some imaginative extrapolation), it looks like sample sizes of ~1000 languages show very close estimates even as *r* increases. Put another way, after 1000 languages, increases in *r* affect the degree of underestimation less.

We can also plot the proximity of the estimates to the true values by substituting the difference between estimated and observed values.

### Plotting error of Chao-Shen estimates in PHOIBLE

```{r}
# Function to plot differences between estimated and true values
#
# Arguments
# - df:       output of iterative.chao.estimator()
# - mean.df:  output of mean.obs.fnc()
# - title:    (string) title for the plot
diff.plotter <- function(df, mean.df, title){
    p <- ggplot(df, 
                aes(x=Sample.Size, 
                    y=Difference, 
                    color=Iteration)) +
                geom_line() +
                geom_line(data=mean.df,
                          aes(x=Sample.Size,
                              y=mean.diff),
                              col="dodgerblue",
                              lwd=0.9) +
                ylab("Difference in type counts (Est. - Obs)") + 
                xlab("Number of languages") +
                theme_bw() +
                ggtitle(title) +
                scale_color_manual(values = rep("grey80", 10)) + 
                theme(legend.position="none",
                      plot.title = element_text(hjust=0.5))

    return(p)
}

chao.iter.diff.plot.10 <- diff.plotter(chao.data.10.10, mean.obs.df.10, "r=10") + ylim(-2500, 400)

chao.iter.diff.plot.10

chao.iter.diff.plot.50 <- diff.plotter(chao.data.50.10, mean.obs.df.50,"r=50") + ylim(-2500, 400)

chao.iter.diff.plot.50

chao.iter.diff.plot.100 <- diff.plotter(chao.data.100.10, mean.obs.df.100,"r=100") + ylim(-2500, 400)

chao.iter.diff.plot.100

chao.iter.diff.plot.500 <- diff.plotter(chao.data.500.10, mean.obs.df.500,"r=500") + ylim(-2500, 400)

chao.iter.diff.plot.500

chao.iter.diff.plot.1000 <- diff.plotter(chao.data.1000.10, mean.obs.df.1000,"r=1000") + ylim(-2500, 400)

chao.iter.diff.plot.1000

chao.iter.diff.plot.1500 <- diff.plotter(chao.data.1500.10, mean.obs.df.1500,"r=1500") + ylim(-2500, 400)

chao.iter.diff.plot.1500

chao.iter.diff.plot.2000 <- diff.plotter(chao.data.2000.10, mean.obs.df.2000,"r=2000")

chao.iter.diff.plot.2000
```

```{r, chao_error_estimates}
chao.diff.plots <- ggarrange(chao.iter.diff.plot.10 + 
                               xlab(element_blank()) +
                               ylab(element_blank()), 
                             chao.iter.diff.plot.50 + 
                               xlab(element_blank()) +
                               ylab(element_blank()), 
                             chao.iter.diff.plot.100 + 
                               xlab(element_blank()) +
                               ylab(element_blank()),
                             chao.iter.diff.plot.500 + 
                               xlab(element_blank()) +
                               ylab(element_blank()), 
                             chao.iter.diff.plot.1000 + 
                               xlab(element_blank()) +
                               ylab(element_blank()), 
                             chao.iter.diff.plot.1500 + 
                               xlab(element_blank()) +
                               ylab(element_blank()),                     
                             chao.iter.diff.plot.2000 + 
                               xlab(element_blank()) +
                               ylab(element_blank()), 
                             ncol=3,
                             nrow=3)

annotate_figure(chao.diff.plots, 
                bottom = "Number of languages",
                left = "Difference (est - obs)")
```

Again, we see a stronger tendency for underestimation. However, convergence on 0 by n=1000 languages or sooner for all values of *r* < 1000.

The time to convergence appears to be inversely related to the ratio of *R*:*r*. Estimates converge on the observed values earlier on average when *R* is larger relative to *r*. Simply put, we do better when we have more data and predict values closer to our observed sample.

Another factor to consider is that the variability in estimates across iterations (grey lines) increases with *r* and decreases with *R*. Our estimates are noisier when we look further ahead and have smaller samples.

We are making estimates with a far smaller ratio of *R*:*r*, something closer to 0.42:1, at the least:

<center>
||||
|---:|:---:|:---|
|*r*|=|7151-2099|
||=|5052|
||||
|*R*<sub>prop</sub>|=|2099 / 5052|
||=|0.42|
</center>

## Predicting known segment type diversity in PHOIBLE

We can now see how models trained on a smaller sample (e.g., 1500 random languages from PHOIBLE) perform in predicting the known value.

From PHOIBLE, we select a random 1500 languages -- an arbitrary number that includes roughly half of the number of languages in the total sample.

### Herdan-Heaps' law

First, we sample 1500 languages randomly from PHOIBLE.

```{r}
set.seed(1)
rand.ids <- sample(unique(phoible_long$InventoryID), 1500, replace=F)

rand.pho.1k <- phoible_long %>%
               filter(InventoryID %in% rand.ids)

set.seed(1)
rand.pho.1k.results <- rand.heaps.data(rand.pho.1k, 10)
```

Next, we model the random sample from PHOIBLE. These functions extrapolate Herdan-Heaps' values from random samples.

```{r}
# Arguments
# df:   database (long format)
# n:    number of languages to sample
# k:    to specify number of iterations
rand.samp <- function(df, n, k){
  exp.values <- vector()
  for(i in 1:k){
    set.seed(i)
    count.df <- df %>%
                group_by(InventoryID) %>%
                summarize(inventory.size = n())
    # Note that we are sampling with replacement
    counts <- sample(count.df$inventory.size, n, replace=T)
    new.size <- sum(counts)
    exp.values <- c(exp.values, new.size)
  }
  return(exp.values)
}

# Arguments
# new.n:   vector of expected values n
# K:       estimate of K from model
# B:       estimate of B from model
# mod:     the HH model for the given dataset
Heaps.extrapolate <- function(new.n, K, B, mod){
                     heaps.df <- data.frame(token_count = new.n,
                                            iteration = 1:length(n))
                     new.heaps <- heaps(K, 
                                        new.n, 
                                        B)
                     heaps.mod <- exp(predict(mod, heaps.df))
                     output.df <- data.frame(new.heaps, heaps.mod)

                     return(output.df)  
                     }
```

Next, we fit the model.

```{r}
# Fit the model
fitHeaps.phoible.rand.1k <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data=rand.pho.1k.results)

# Get the coefficients
logK.phoible.1k <- fixef(fitHeaps.phoible.rand.1k)[1]
B.phoible.1k <- fixef(fitHeaps.phoible.rand.1k)[2]
K.phoible.1k <- exp(logK.phoible.1k)

# Determine how many tokens we need
new.n <- rand.samp(phoible_long, 2099, 10)

# Get the expected type counts
ex.heaps.pho.2k <- Heaps.extrapolate(new.n, K.phoible.1k, B.phoible.1k, fitHeaps.phoible.rand.1k) %>%
                  mutate(diff.from.true.heaps = length(unique(phoible_long$Phoneme))- new.heaps,
                         diff.from.true.heaps.mod = length(unique(phoible_long$Phoneme))- heaps.mod)

ex.heaps.pho.2k
summary(ex.heaps.pho.2k)
```

And visualize the result.

```{r, herdan_heaps_ci}
# Function taken from https://tinyurl.com/8awbaxpe
confidence_interval <- function(vector, interval) {
  # Standard deviation of sample
  vec_sd <- sd(vector)
  # Sample size
  n <- length(vector)
  # Mean of sample
  vec_mean <- mean(vector)
  # Error according to t distribution
  error <- qt((interval + 1)/2, df = n - 1) * vec_sd / sqrt(n)
  # Confidence interval as a vector
  result <- c("lower" = vec_mean - error, "upper" = vec_mean + error)
  return(result)
}

mean.val <- mean(ex.heaps.pho.2k$heaps.mod)
cis <- confidence_interval(ex.heaps.pho.2k$heaps.mod, .975)
upper <-cis[2]
lower <- cis[1]
true.val <- length(unique(phoible_long$Phoneme))

heaps.2k.plot <- ggplot(ex.heaps.pho.2k, aes(x=heaps.mod)) +
  geom_density(fill="dodgerblue", alpha=0.5) +
  geom_rect(aes(xmin=lower, 
                xmax=upper, 
                ymin=-Inf, 
                ymax=Inf), 
                fill="grey90",
                alpha=0.09) +
  geom_vline(xintercept = true.val, lty=2,  col="red") +
  geom_vline(xintercept = mean.val, lty=1) +
  geom_vline(xintercept = upper, lty=2) +
  geom_vline(xintercept = lower, lty=2) +
  xlab("Number of phoneme types") + 
  ylab("Probability density") + 
  annotate("text", x=mean.val+24, y=.02, label= paste(floor(mean.val), "\n(est.)", collapse="")) +
  annotate("text", x=true.val-20, y=.02, label= paste(true.val, "\n(true)", collapse=""), col="red") +
  theme_bw() +
  xlim(2550, 2750)

heaps.2k.plot

abs(true.val-mean.val)
```

The 97.5% confidence intervals (CIs) are plotted around the estimate. The difference between the estimated and true values was ~18, indicating a slight overestimate.

### Chao-Shen estimator

Get a random sample.

```{r}
get_random_sample <- function() {
  random_1000 <- phoible_long %>%
    distinct(InventoryID) %>%
    sample_n(1500)
  sample_1000 <- phoible_long %>%
    filter(InventoryID %in% random_1000$InventoryID)
  freq_1000 <- sample_1000 %>%
    group_by(Phoneme) %>%
    summarize(Frequency = n()) %>%
    arrange(desc(Frequency))
}
```

Predict type count with *r* = 2099-1500 = 599 and *R* at 1000.

```{r, message=F}
set.seed(1)
random_sample <- get_random_sample()
r <- length(unique(phoible_long$InventoryID)) - 1500
vector <- random_sample$Frequency
chao.est.1k <- Extrapolate(vector, 1500, r)

chao.est.1k
```

As we saw before, the true value is 2596, which makes the difference in this estimate:

```{r}
num.phons <- length(unique(phoible_long$Phoneme))

abs(num.phons - chao.est.1k)
```

The number of types is overestimated by ~41 phonemes, indicating worse performance than the HH model.

Let's do it *k* times and look at the distribution of outcomes.

Here, we set *k* to 100 iterations.

```{r, message=F}
set.seed(1)
results <- sapply(1:100, function(x) {
  random_sample <- get_random_sample()
  r <- 2099 - 1500
  vector <- random_sample$Frequency
  x <- Extrapolate(vector, 1500, r)
})
```

Plot the results.

```{r, chao_ci}
mean.val <- mean(results)

cis <- confidence_interval(results, .975)
upper <-cis[2]
lower <- cis[1]
true.val <- length(unique(phoible_long$Phoneme))

plot.dat <- data.frame(results)

chao.2k.plot <- ggplot(plot.dat, aes(x=results)) +
  geom_density(fill="dodgerblue", alpha=0.5) +
  geom_rect(aes(xmin=lower, 
                xmax=upper, 
                ymin=-Inf, 
                ymax=Inf), 
                fill="grey90",
                alpha=0.013) +
  geom_vline(xintercept = mean.val, lty=1) +
  geom_vline(xintercept = upper, lty=2) +
  geom_vline(xintercept = lower, lty=2) +
  geom_vline(xintercept=true.val, lty=2,  col="red") +
  xlab("Number of phoneme types") + 
  ylab("Probability density") + 
  annotate("text", x=mean.val-50, y=.001, label= paste(floor(mean.val), "\n(est.)", collapse="")) +
  annotate("text", x=true.val+50, y=.001, label= paste(true.val, "\n(true)", collapse=""), col="red") +
  theme_bw()

chao.2k.plot

abs(mean.val-true.val)
```

Estimated and true values are given, along with .975 confidence interval.

The Chao-Shen estimator produced a mean almost identical to the true number of phonemes.

## Extrapolating estimates to unseen numbers of observations

### Herdan-Heaps' law

Above, we extracted estimates for the two free parameters in Herdan-Heaps' equation by performing a (log) linear regression. These parameters can also be used to extrapolate to unseen numbers of observations. All we need to know is how many observations (*n*) we expect to see in any (arbitrarily large) sample.

However, since we do not have access to the hypothetical larger sample, we face a problem. Our goal is to guess how many sound types we would expect to find in a sample of *x* languages. But Herdan-Heaps' equation requires us to know the *number of observations* we would make, i.e., the sum of all of the inventory sizes of languages in the sample. We do not know how many sounds each language would contribute in an unseen population, so we cannot directly derive the requisite number of observations from the number of languages alone. Therefore, we need some way to estimate this figure.

#### Estimating *n*

We approach this problem in two ways.

Approach 1: Random re-sampling with replacement from real data. The steps are:

1. Randomly sample *L* languages from the existing database with replacement.
2. Sum the inventory sizes of all languages *L*.
3. Add this figure to the known number of observations.
4. Recompute Herdan-Heaps' estimate.

NB: We want a single estimate of the number of additional phonemes we should expect to see. Therefore, we rely on the larger database (PHOIBLE) to generate the estimate, and use the resulting value for both the UPSID-41 (genealogically balanced sample) and PHOIBLE (convenience sample) models.

Approach 2: Regression modeling of sample size. The steps are (1--3 same as above):

1. Randomly sample *L* languages from the existing database with replacement.
2. Sum the inventory sizes of all languages *L*.
3. Add this figure to the known number of observations.
4. Fit a linear regression: number of observations ~ number of languages.
5. Use the model to predict expected number of observations for any number of languages.

##### Approach 1: Random resampling

First, a function to extrapolate values.

```{r rr_funcs}
# Arguments
# df:   database (long format)
# n:    number of languages to sample
# k:    to specify number of iterations
rand.samp <- function(df, n, k){
  exp.values <- vector()
  for(i in 1:k){
    set.seed(i)
    count.df <- df %>%
                group_by(InventoryID) %>%
                summarize(inventory.size = n())
    counts <- sample(count.df$inventory.size, n, replace=T)
    new.size <- sum(counts)
    exp.values <- c(exp.values, new.size)
  }
  return(exp.values)
}

# Arguments
# new.n:   vector of expected values n
# K:       estimate of K from model
# B:       estimate of B from model
# mod:     the HH model for the given dataset
Heaps.extrapolate <- function(new.n, K, B, mod){
                     heaps.df <- data.frame(token_count = new.n,
                                            iteration = 1:length(n))
                     new.heaps <- heaps(K, 
                                        new.n, 
                                        B)
                     heaps.mod <- exp(predict(mod, heaps.df, re.form = NA))
                     output.df <- data.frame(new.heaps, heaps.mod)

                     return(output.df)  
                     }
```

Next we extrapolate to 7151, 10K, 50k, and 100k languages, respectively, in UPSID-451 and PHOIBLE.

```{R}
# Get estimates of the observations we need
exp.obs.7k <- rand.samp(phoible_long, 7151, 10)
exp.obs.10k <- rand.samp(phoible_long, 10000, 10)
exp.obs.50k <- rand.samp(phoible_long, 50000, 10)
exp.obs.100k <- rand.samp(phoible_long, 100000, 10)

# Get the means
mean.obs.7k <- mean(exp.obs.7k)
mean.obs.10k <- mean(exp.obs.10k)
mean.obs.50k <- mean(exp.obs.50k)
mean.obs.100k <- mean(exp.obs.100k)

# UPSID
# 7k
ex.heaps.upsid.7k <- Heaps.extrapolate(exp.obs.7k, K.upsid, B.upsid, fitHeaps.upsid)

ex.heaps.upsid.7k
summary(ex.heaps.upsid.7k)

# UPSID
# 10k
ex.heaps.upsid.10k <- Heaps.extrapolate(exp.obs.10k, K.upsid, B.upsid, fitHeaps.upsid)

ex.heaps.upsid.10k
summary(ex.heaps.upsid.10k)

# 50k
ex.heaps.upsid.50k <- Heaps.extrapolate(exp.obs.50k, K.upsid, B.upsid, fitHeaps.upsid)

ex.heaps.upsid.50k
summary(ex.heaps.upsid.50k)

# 100k
ex.heaps.upsid.100k <- Heaps.extrapolate(exp.obs.100k, K.upsid, B.upsid, fitHeaps.upsid)

ex.heaps.upsid.100k
summary(ex.heaps.upsid.100k)

# PHOIBLE
# 7k
ex.heaps.pho.7k <- Heaps.extrapolate(exp.obs.7k, K.phoible, B.phoible, fitHeaps.phoible.all)

ex.heaps.pho.7k
summary(ex.heaps.pho.7k)

# 10k
ex.heaps.pho.10k <- Heaps.extrapolate(exp.obs.10k, K.phoible, B.phoible, fitHeaps.phoible.all)

ex.heaps.pho.10k
summary(ex.heaps.pho.10k)

# 50k
ex.heaps.pho.50k <- Heaps.extrapolate(exp.obs.50k, K.phoible, B.phoible, fitHeaps.phoible.all)

ex.heaps.pho.50k
summary(ex.heaps.pho.50k)

# 100k
ex.heaps.pho.100k <- Heaps.extrapolate(exp.obs.100k, K.phoible, B.phoible, fitHeaps.phoible.all)

ex.heaps.pho.100k
summary(ex.heaps.pho.100k)
```

We can plot these distributions for comparison.

```{r}
extrap.plot.data <- bind_rows(ex.heaps.pho.7k %>%
                                mutate(Database = as.factor(rep("phoible",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("7151",
                                                                nrow(.)))),
                              ex.heaps.pho.10k %>%
                                mutate(Database = as.factor(rep("phoible",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("10000",
                                                                nrow(.)))),
                              ex.heaps.pho.50k %>%
                                mutate(Database = as.factor(rep("phoible",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("50000",
                                                                nrow(.)))),
                              ex.heaps.pho.100k %>%
                                mutate(Database = as.factor(rep("phoible",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("100000",
                                                                nrow(.)))),
                              ex.heaps.upsid.7k %>%
                                mutate(Database = as.factor(rep("upsid",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("7151",
                                                                nrow(.)))),
                              ex.heaps.upsid.10k %>%
                                mutate(Database = as.factor(rep("upsid",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("10000",
                                                                nrow(.)))),
                              ex.heaps.upsid.50k %>%
                                mutate(Database = as.factor(rep("upsid",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("50000",
                                                                nrow(.)))),
                              ex.heaps.upsid.100k %>%
                                mutate(Database = as.factor(rep("upsid",
                                                                nrow(.))),
                                       Estimate = as.factor(rep("100000",
                                                                nrow(.)))))

extrap.p <- ggplot(extrap.plot.data, aes(x = Database, y = heaps.mod, fill = Estimate)) +
  geom_boxplot() + 
  theme_bw()

extrap.p
```

The PHOIBLE-based model predicts more sounds than the UPSID-based model, and the difference appear to get more extreme as *n* increases. This pattern makes sense given that UPSID-451 showed a shallower curve than PHOIBLE.

The final mean estimates for Approach 1 are as follows:
```{r}
mean.extrap.heaps <- extrap.plot.data %>%
                     group_by(Database, Estimate) %>%
                     summarize(mean.heaps = mean(heaps.mod))

pho.mean.extrap <- mean.extrap.heaps %>%
                   filter(Database == "phoible") %>%
                   pull(mean.heaps)

ups.mean.extrap <- mean.extrap.heaps %>%
                   filter(Database == "upsid") %>%
                   pull(mean.heaps)

cat("PHOIBLE\n")
names(pho.mean.extrap) = c("7151 langs", "10000 langs", "50000 langs", "100000 langs")
pho.mean.extrap

cat("\nUPSID\n")
names(ups.mean.extrap) = c("7151 langs", "10000 langs", "50000 langs", "100000 langs")
ups.mean.extrap

cat("\nDifferences\n")
diff.extrap <- pho.mean.extrap-ups.mean.extrap
names(diff.extrap) = c("7151 langs", "10000 langs", "50000 langs", "100000 langs")
diff.extrap
```

The means confirm the visual impressions from the above plot. PHOIBLE produces higher estimates, and it grows at a faster rate than UPSID-451, as evidenced by the growth in the size for the differences between the estimates as a function of sample size.

##### Approach 2: Regression modeling

Here we explore an additional way to estimate sample size. We create a model to predict number of observations from the number of languages.

```{r data_to_estimate_sample_size, message=F}
# Function to sample iteratively from PHOIBLE at incremental sample sizes,
# without replacement. The goal is to eliminate the arbitrary influence of
# language order when computing cumulative frequency

# args:
# df = dataframe (e.g., my_sample or phoible_long)
iter.sampling <- function(df, k){
    output <- vector(mode = "list")
    ids <- unique(df$InventoryID)
    phoible.size <- length(ids)
    inv.sizes <- df %>% 
                  group_by(InventoryID) %>% 
                  summarize(inv.size = n())
    j <- 1
    for(i in 1:k){
        ct.list <- vector(mode = "list", length = phoible.size) 
        # Increments of 10 as we increase sample size
        for(n.inv in seq(1, phoible.size, 10)){
            set.seed(n.inv)
            SampleFrequency <- inv.sizes %>%
                               sample_n(., n.inv) %>%
                               pull(inv.size) %>%
                               sum()

            curr.row <- c(SampleFrequency, n.inv, i)
            output[[j]] <- curr.row
            j <- j+1
        }
    }
    output <- as.data.frame(do.call(rbind, output))
    colnames(output) <- c("SampleFrequency", "InventoryCt", "Iteration")
    return(output)
}

sample.counts <- iter.sampling(my_sample, 10)
```

To extrapolate to the values we want, we create a regression-based estimator.

```{r modeling_sample_size}
##########
# Modeling
##########
# Try with lmer
mod.lmer = lmer(SampleFrequency ~ InventoryCt + (1|Iteration), data=sample.counts)
summary(mod.lmer) # near perfect fit

# Try just lm (technically, no one of the pulls in any iteration depend on the others)
mod.lm = lm(SampleFrequency ~ InventoryCt, data=sample.counts)
summary(mod.lm) # Still, nearly perfect.
```

The model fits are nearly perfect, but that makes sense. It means that most languages have a very similar number of segments to contribute. Therefore, we need not worry about overfitting in applying these models to new data. For simplicity, we apply the `lm` model.

```{r applying_models}
#####################
# Applying the models
#####################
hyp.data = data.frame(InventoryCt = seq(1:100000), Iteration = rep("test", 10000))

# Produce estimates
hyp.preds.with = predict(mod.lm, newdata = hyp.data)

## True (72,733 is correct)
n.2099 <- hyp.preds.with[2099]

## 7151
n.7k <- hyp.preds.with[7151]

## 10000
n.10k <- hyp.preds.with[10000]

## 50000
n.50k <- hyp.preds.with[50000]

## 100000
n.100k <- hyp.preds.with[100000]

c(n.2099, n.7k, n.10k, n.50k, n.100k)

```

Finally, we can feed these values into our Herdan-Heaps' estimator to get specific predictions for the expected number of phonemes given the larger sample sizes.

```{r predict_from_heaps}
hyp.data.heaps <- data.frame(token_count = c(n.7k, n.10k, n.50k, n.100k), iteration = rep(1, 4))

predictions <- exp(predict(fitHeaps.phoible.all, newdata=hyp.data.heaps, re.form=NA))
names(predictions) = c("7151 langs", "10000 langs", "50000 langs", "100000 langs")
predictions

cat("\nDifferences\n")
predictions-pho.mean.extrap
```

The PHOIBLE estimates are generally higher than those based on Approach 1. The differences are perhaps non-trivial (on the order of 100s of segment types). Notice also that the differences increase with sample size. This suggests that our randomly resampling technique (Approach 1) produces estimates that grow at a slower pace than the linear model would predict.

### Chao-Shen estimator

Prepare the data.

```{r, message=F}
freq <- phoible_long %>%
        group_by(Phoneme) %>%
        summarize(Frequency = n()) %>%
        arrange(desc(Frequency))
```

Compute *r* for different target sample sizes.

```{r}
r.7k <- 7151 - length(unique(phoible_long$InventoryID))
r.10k <- 10000 - length(unique(phoible_long$InventoryID))
r.50k <- 50000 - length(unique(phoible_long$InventoryID))
r.100k <- 100000 - length(unique(phoible_long$InventoryID))
```

#### Chao-Shen extrapolation

Results from PHOIBLE (2099 distinct languages) for *r* = {7151, 10k, 50k, 100k}.

```{r}
num_langs <- nrow(my_sample %>% select(InventoryID) %>% distinct())

vector <- freq$Frequency
chao.7k <- Extrapolate(vector, num_langs, r.7k)
chao.10k <- Extrapolate(vector, num_langs, r.10k)
chao.50k <- Extrapolate(vector, num_langs, r.50k)
chao.100k <- Extrapolate(vector, num_langs, r.100k)

chao.ext <- c(chao.7k, chao.10k, chao.50k, chao.100k)
names(chao.ext) = c("7151 langs", "10000 langs", "50000 langs", "100000 langs")

chao.ext

cat("\nDifference from PHOIBLE (Approach 1)\n")
pho.mean.extrap-chao.ext

cat("\nDifference from PHOIBLE (Approach 2)\n")
predictions-chao.ext

cat("\nDifference from UPSID\n")
ups.mean.extrap-chao.ext

```

Plot these results
```{r diff_plots_ext}
plot.diff.dat <- data.frame(Type.ct.est = c(pho.mean.extrap, predictions, chao.ext), Estimator = factor(rep(c("HH Approach 1", "HH Approach 2", "Chao-Shen"), each = 4)), Sample.Size = factor(rep(c("7151", "10K", "50K", "100K"), 3)))

plot.diff.dat$Sample.Size = factor(plot.diff.dat$Sample.Size, levels = levels(plot.diff.dat$Sample.Size)[c(4, 2, 3, 1)])

p = ggplot(plot.diff.dat, aes(y = Type.ct.est, x = Estimator, fill = Sample.Size), group = Estimator) +
    geom_bar(stat="identity", position="dodge") + 
    ylab("Type count (est.)") +
    labs(fill=expression(italic("r"))) + 
    theme_bw() +
    theme(legend.title.align = 0.5)

p

```

The Chao-Shen estimator produces estimates much lower than Herdan-Heaps' equation for the PHOIBLE-based models. For UPSID-451, it predicts substantially higher values for lower *r*, but much lower values for higher *r* (50k+). 
To this point, the Chao-Shen estimator becomes increasingly conservative as *r* increases and *R* is held constant. Under more uncertain conditions with less data to support estimation, increases in type count are small for even very large increases in hypothetical sample size.

Given the difference in results, we now compare the two estimators on known data, i.e., PHOIBLE.

## Comparing Herdan-Heaps' to Chao-Shen estimator

We can compare the iterative performance of the Chao-Shen estimator to that of Herdan-Heaps' estimator.

```{r, comp_plot}
# Combine the mean tables
all.mean.obs <- bind_rows(mean.obs.df.10,
                          mean.obs.df.50,
                          mean.obs.df.100,
                          mean.obs.df.500,
                          mean.obs.df.1000,
                          mean.obs.df.1500,
                          mean.obs.df.2000) %>%
                mutate(target.val = Sample.Size + r)

# Need to shift the expected number of observations to match
# the estimate (because with Chao-Shen, our estimate for R given r 
# is for the value R + r)
shifted.estimates <- left_join(all.mean.obs,
                               all.mean.obs %>% 
                               select(mean.est, 
                                      mean.obs, 
                                      Sample.Size,
                                      r) %>%
                                rename(target.val = Sample.Size,
                                       target.est = mean.est,
                                       target.obs = mean.obs),
                               by=c("r", "target.val"))

# Plotting against HH curve and observed values from PHOIBLE
comp.plot <- ggplot(heaps_results_phoible, 
                    aes(x=token_count,
                        y=type_count)) +
                    geom_point(alpha=0.007, color="grey70") +
                    geom_line(data=plot.dat.pho, 
                              aes(x=token_count,
                                  y=Predicted),
                              color="black", lwd=2) +
                    geom_line(data=shifted.estimates,
                              aes(x=target.obs,
                                  y=mean.est, color=as.factor(r)),
                              lwd=1) +
                    geom_hline(yintercept = length(unique(my_sample$Phoneme)), 
                               lty=2, col="red") +
                    theme_bw() +
                    ylim(0, 3000) +
                    labs(color="r") +
                    xlab("Number of tokens") +
                    ylab("Number of types")

comp.plot
```

The black line represents PHOIBLE-based Herdan-Heaps' estimates. The colored lines represent Chao-Shen estimates for different *r*. The dotted red line indicates the known number of types. Grey points reflect true cumulative type counts from 10 reshuffled versions of PHOIBLE.

The fits for Herdan-Heaps' and Chao-Shen for the small-scale *r* values are almost identical. Higher values of *r* show Chao-Shen converging relatively quickly on the trend described by Herdan-Heaps'. This suggests that if Herdan-Heaps' produces higher estimates than Chao-Shen, we might have reason to trust the former when dealing with small *R*:*r* ratios like the ones we have here (as discussed above).

Importantly, the Chao-Shen estimator seems to be "seeking out" the trend expressed by Herdan-Heaps'. However, it needs to be fed the right ratio of data to *r* in order to converge quickly on the true values. This fact provides further, independent support for the suitability of Herdan-Heaps' as an estimator of type counts.

A final point: none of these trends shows signs of approaching an asymptote. The evidence therefore suggests that there is no upper bound to the number of unique phonemes we should expect to find by documenting and describing the phonological systems of more and more languages. In other words, infinite linguistic diversity seems to breed infinite possible phonemes.

# References

### 2023-06-06 paper.Rmd
initial commit

---
title: "How many speech sounds are there in the world's languages?"
author: "Steven Moran & Nicholas A. Lester\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    number_sections: true
bibliography: 'bibliography.bib'
biblio-style: "apalike"
csl: unified-style-sheet-for-linguistics.csl
link-citations: true
header-includes:
  \usepackage{fontspec}
  \setmainfont{Times New Roman}
  \setmonofont{Arial Unicode MS}
  \usepackage{lineno}
  \linenumbers
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(out.width="80%")
knitr::opts_chunk$set(echo = TRUE, error=F, warning=F, message=F, cache=T)
```

<!--
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
-->

# Abstract {-}

How many speech sounds are there in the world's languages? To answer this question, we extend two formal models of lexical diversity to segmental phonology. Zipf's law states that most word types are rare; Herdan-Heaps' law states that vocabulary size grows indefinitely with increasingly new text. We find that both models produce accurate estimates of the diversity of contrastive speech sounds (phonemes) in existing typological databases. Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) documenting more phonological repertoires always yield more phoneme types cross-linguistically. As a sanity check, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates align with Herdan-Heaps' law. We use both models to extrapolate phoneme counts to undocumented languages, resulting in an estimate of 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k), which is roughly 1.5x more than what is currently documented in the world's languages (~35%). By extrapolating to hypothetical sample sizes, e.g., the number of languages estimated to have been spoken through time, and beyond, we find no indication of convergence: infinite languages mean infinite phonemes.

<!-- How many speech sounds are there in the world's languages? To answer this question, we extend two formal models of lexical diversity -- Zipf's law and Herdan-Heaps' law -- to segmental phonology. Zipf's law states that most word types are rare; Herdan-Heaps' law states that vocabulary size grows indefinitely with increasing amounts of new text. We find that both models produce accurate estimates of the diversity of contrastive speech sounds, aka *phonemes*, in existing typology databases. Our findings suggest that (a) most speech sounds are difficult to detect because they are rare, and (b) documenting more phonological repertoires always yield more phonemes cross-linguistically. As a sanity check, we apply an unrelated estimator of species diversity from biology, Chao-Shen, which combines information about rare types and sample size. We find that the number of phonemes predicted by Chao-Shen align with those predicted by Herdan-Heaps' law. We then use both models to extrapolate phoneme counts to undocumented languages. Our results suggest that there are between 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k). This figure is roughly 1.5 times the number of currently documented speech sounds found in the broadest typological databases available (~35\% of the world's languages). By extrapolating to hypothetical sample sizes, e.g., the number of languages estimated to have been spoken through time, and beyond, we find no indication of convergence: infinite languages mean infinite speech sounds. -->

*Keywords*: speech, phonetics, phonology, language, Zipf's law, Herdan-Heaps' law

***

# Introduction

Under the umbrella of linguistic typology, studies of language diversity aim to catalog the range of variation in comparable structures across the world's languages. For some areas of variation, the problem is moot, i.e., the full range of options is known *a priori*. For example in grammar, the relative ordering of a verb (V) and its direct object (O) allows for exactly three possible types: OV, VO, and OV/VO. Thus, every language falls into one of these categories. Other linguistic domains are not logically constrained in this way because the true diversity of types is inherently unknown.

Consider the set of possible speech sounds in the world's languages. Phonemes are distinguished according to their articulatory, acoustic, and perceptual properties. However, these properties exist in a continuous temporal and physical space, which can be carved up in infinitely many ways. For example, at suitably fine-grained levels of measurement, no two realizations of the same phoneme will ever be identical in every dimension of variation.

Thus, each new word presents a non-zero probability of containing a previously unheard sound, and by extension, new phoneme. Complicating matters, our current pool of "consensus phonemes" -- however defined -- is subject to re-evaluation in response to advances in data collection, methods, tools, theories, and experience that inform language documentation and analysis.^[In the last 20 years or so, grammars and phonological descriptions (e.g., *Illustrations of the IPA*) increasingly include detailed data and acoustic phonetic analysis using software, such as Praat [@Boersma2001]. This makes phonetic interpretation open to a wider audience.]

<!-- Physiological differences between speakers' vocal tracts means that equivalance does not exist at the acoustic level. As Papcun notes, "Although various scholars have claimed that phonetic quality is particularly a function of one or another of these domains, this is an incorrect notion. Rather, phonetic quality must be defined in all its domains and mappings must be found between them." -->

The problem is therefore to determine whether there is some sample size of languages after which no amount of additional phonological inventories will yield an unseen phoneme type. The absence of a limit would raise important theoretical issues for the description of phonemes, including questions concerning the role of phonetics in phonology [@Whalen2019] and the importance of cross-linguistic variation [@Hockett1963;@EvansLevinson2009].

The present study seeks to determine whether such a limit exists. To meet this goal, we propose several methods for modeling the relationship between sample size and number of linguistic types. We apply these methods to ask three core questions. First, how many phonemes are there in the world today (based on a limited sample)? Second, how many phonemes have there been in the history of human language? And lastly, how many phonemes are possible in human language? In Section 2, we motivate our research questions. In section 3, we describe the data and in Section 4 we present our methods and results. In Section 5, we discuss our results and avenues for future research.

<!-- In sum, we show that segment type distributions in typological databases follow Zipf's law, which motivates a connection to Herdan-Heaps' law. We then evaluate these distributions against Herdan-Heaps' law and find a tight fit. We then evaluate these results against an independent type estimator from biology, the Chao-Shen estimator [@ChaoChiu2016], and compare how they both perform on various subsets of known languages' phonological inventories. Lastly, we use both estimators to predict the number of segment types in increasingly larger and undocumented samples of languages. To conclude, we discuss why we expect that ever increasing linguistic diversity leads to an ever increasing number of phonemes, and the broad applications that these methods have for inferring the number of linguistic types more generally.-->

# Motivation

The question of whether there are limits on the number of distinct speech sounds that can appear in spoken languages has been recognized for some time. Alexander J. Ellis (1814--1890) stated as early as 1848 that the number of different speech sounds, i.e., candidate phonemes, is in principle infinite. At that time, teachers and philologists aimed to develop a transcription standard that could discriminate between all manners of speech sounds with orthographic symbols for purposes that included (particularly English) spelling reform and orthography development for newly documented and unwritten languages. But as practitioners increasingly converted the articulations of more and more languages into a universal phonetic alphabet, the need for new symbols also seemingly increased without limit [@Hockett1995-phoneme, pg. 63].^[This early work later became the basis for the International Phonetic Association (and its phonetic alphabet; both abbreviated IPA). First published in 1888, today the IPA provides and maintains a speech transcriptional device that encodes aspects of articulation, and acoustics, for both narrow (phonetic) and broad (phonemic) transcription. It is comprised of a set of alphabetic symbols and diacritics with the aim of representing all possible sounds in the world's spoken languages (there are extensions for disordered speech and speech pathology). The IPA provides a summary of linguists' phonetic knowledge. This system uses a set of "base" symbols, e.g., letters or combination of letters, that aim to encode phonemes, and the use of diacritics on base glyphs to provide phonetic detail. Its design also affords infinite transcription possibilities.] Finer and finer distinctions in phonetic detail created more and more space for novel phonemes to emerge. Thus, if the phonetic distinctions to be drawn are infinite, then tautologically so are the number of possible phonemes.

If there was any doubt about the infinite nature of the speech signal, instrumental phonetics put the nail in the coffin. Jean-Pierre Rousselot (1846--1924) studied acoustic properties of speech with the kymograph, originally developed to measure physiological patterns such as blood pressure. The later advent of the phonograph by Thomas Edison (1847--1931) in 1877 provided a means to record and reproduce sound. Then during WWII the sound spectogram was developed in secret to analyze ciphony by making sounds visible. The work on visible speech was released publicly after the war [@Potter1945] and the spectrogram became the staple of acoustic phonetic analysis beginning with @Joos1948 (Joos was the first linguist to have extensive access to the acoustic spectrograph). For speech, it measured fluctuations in air pressure, which capture duration, pitch, and intensity. The recorded speech waveforms showed visually that speech is dynamic and continuous, even if listeners perceive it as discrete, e.g., phonemes, syllables, words.^[This relationship between continuous and discrete is at the core of the debate about the interface between phonetics and phonology [@Whalen2019]. The division goes back to at least the late 19th century when phoneticians and dialectologists including Henry Sweet (1845--1912; 1877) and Jost Winteler (1846--1929; 1876) began making a distinction in language descriptions between sounds that change the meaning of words and those that do not.]

Aside from the acoustic-phonetic argument, evidence for infinite phonemes comes from studies on the cross-linguistic frequencies of speech sounds. Early evidence is found in a large cross-linguistic sample of phoneme inventories, where inventory is understood as the set of distinctive sounds attested in a language. This sample is described in the UCLA Phonological Segment Inventory Database (UPSID; @Maddieson1984, @MaddiesonPrecoda1990). @Maddieson1984 created a genealogically-balanced sample of the world's languages (*n* = 317) to investigate cross-linguistic properties of phonological systems and the relationship between aspects such as inventory size, contents, and structure. Maddieson noticed that certain segments are more or less likely to be present in inventories regardless of their size. That is, certain sounds are almost always present and others are almost never present. In other words, no segment occurs in every language and some very frequent segments occur in most languages. Moreover, nearly half of the reported segment types are singletons in the sample, i.e., one-off language-specific occurrences.^[Whether rare segments occur more or less frequently *within* a language is a language-by-language research question. See for example [@Everett2018].] While Maddieson did not explicitly test for it, his observations fit a well-attested pattern known as Zipf's law [@Zipf1936;@Zipf1949]. Zipf's law states that the frequency distribution of words in a language is inversely proportional to their frequency rank. The relationship is log-linear: high frequency words are few in number but account for most of the tokens in a text, and low frequency words are many, but occur rarely or even just once.^[Here is a plot showing Zipf's law in the first 10 million words in the Wikipedias from 30 different languages: https://en.wikipedia.org/wiki/Zipf%27s_law#/media/File:Zipf_30wiki_en_labels.png.] , Zipf's law states that the frequency of an element $f(r)$ is proportional to the inverse frequency rank of that element (modulated by an exponent $\alpha$) [@Piantadosi2014]:^[@Piantadosi2014 discusses in detail that word frequency distributions are not actually as simple as they seem according to Zipf's law. For example, semantics affects word frequency distributions. Moreover, observations of Zipf's law have been obscured by word frequency distribution plots that fail to address correlated errors. Zipfian distributions are not unique to language. They appear in various aspects of human society, such as music and aspects of the internet, as well as in other biological and physical systems. Strikingly, power law distributions can arise from nearly nothing [@Piantadosi2014, p. 1121].]

$$
f(r) \propto \frac{1}{r^\alpha}
$$

The Zipfian distribution characterizes many aspects of language. Two features make it interesting for our purposes. First, it describes systems for which most types are rare, hence difficult to observe. Second, under certain conditions ($\alpha \ge 1$), the Zipf function is non-asymptotic: more types can always be added at the rank for singletons, and the expected type count computed. The fact that Maddieson's observations are consistent with the Zipfian distribution suggests that phoneme types are indeed difficult to discover (particularly at the fringe) and that there could be some room left for additional phonemes.

@Moran2012 revisited the frequency distribution of segments in a larger sample of the world's languages (*n* = 1780, including UPSID) and showed that the relationship between cross-linguistic segment frequency follows a similar pattern as in the @Maddieson1984 (*n* = 317) and @MaddiesonPrecoda1990 (*n* = 451) UPSID language samples. That is, some phonemes are very frequent across the world's languages, and half of all documented phonemes are language-specific. @Moran2012 also did not explicitly test whether cross-linguistic phoneme frequency adheres to a Zipfian distribution,^[@Moran2012, p. 212 does provide a log-log plot of segment frequencies that is Zipfian-esque.] but he does show that the cumulative number of segment types seems to be bounded to infinity; the curve appears quadratic and showing no sign of a slowing towards an asymptote.

In Figure 1, we reproduce these plots using the genealogically balanced UPSID-451 language sample [@MaddiesonPrecoda1990] and compare it to the data available in the latest version of PHOIBLE Online [@MoranMcCloy2019], which contains 2186 languages.^[Note that although nearly half of the segments in the language samples occur in particular languages, some languages contain more than one segment not found in any other languages in the sample. This is most apparent from the rich phonological inventory of Juǀʼhoan (ISO 639-3: ktz; Glottocode: juho1239; aka !Xu, ǃXun, Southeastern ǃKung), which has been described as having 141 phonemes -- 66 of which occur in no other documented language [@Snyman1970;@Snyman1975;@MaddiesonPrecoda1990].]

<!--
```{r upsid, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}Cumulative segments in UPSID-451"}
knitr::include_graphics("README_files/figure-gfm/cumulative_plot_upsid-1.png")
```

```{r phoible, echo=FALSE, fig.align="center", fig.cap="Cumulative segments in PHOIBLE 2.0"}
knitr::include_graphics("README_files/figure-gfm/cumulative_plot_phoible-1.png")
```
-->

```{r phoible_upsid, echo=FALSE, fig.align="center", fig.cap="Cumulative segments in PHOIBLE 2.0 and UPSID-451"}
knitr::include_graphics("README_files/figure-gfm/cumulative_plot_phoible_upsid-1.png")
```

These plots resemble the distribution of another, less-commonly known linguistic law: Herdan's and/or Heaps' law. Originally proposed by Gustav Herdan [-@Herdan1964], but frequently attributed to Harold Heaps [-@Heaps1978] in the information retrieval literature [@Egghe2007], Herdan-Heaps' law describes the number of distinct words within a document as a function of the document's (or combined documents') length in terms of total word count. More specifically, Herdan–Heaps' law states that, as more of a text is consumed, the rate of never-before-seen words per unit diminishes (but does not reach 0). This makes intuitive sense given what we know from Zipf's law: most words are hard to find, and most of the words that you can find, you have seen before. In fact, Herdan–Heaps' and Zipf's laws have been shown to be asymptotically equivalent with respect to individual word frequencies within corpora [@BaezaNavarro2000;@vanLeijenhorst2005]. Both of these laws are instances of the same fundamental property: token counts can be inferred from type counts and vice versa [@Milivcka2009].

Herdan-Heaps' is expressed  as:

$$
V_{R}(n)=Kn^{\beta }
$$

where $V_{R}$ is the number of distinct words in a text of size _n_ and _K_ and $\beta$ are the free parameters that are determined empirically. An illustration of a typical Herdan-Heaps' curve for English text is given in Figure 2 (compare to Figure 1 above).

```{r heaps, echo=FALSE, fig.align="center", fig.cap="A typical Herdan-Heaps plot, where x-axis denotes text size and the y-axis the number of distinct vocaulary items (Source: Wikipedia)."}
knitr::include_graphics("figures/Heaps_law_plot.png")
```

Herdan–Heaps' law has been validated using large corpora of naturally occurring discourse. For example, in a corpus of 50 million words, @Kornai2002 found that iteratively adding more text always led to the discovery of more unique words. @BrantsFranz2006, using a custom one-trillion-word corpus of English, showed that even with nearly 14 million word types, the growth in vocabulary showed no sign of stopping.^[Note that "words" at the trillion-word scale corpus include all segmentable substrings, including alternate spellings for the same target word. These alternatives can be numerous. For example, Google's spell correction system has detected nearly 600 different spellings for the query *britney spears*: http://archive.google.com/jobs/britney.html.]

We can apply the same approach to phonemes nested within inventories. By analogy to words and documents, phonemes are the words, and the concatenated inventories of a sample of languages (*L*) is a document. The "document" is expanded by considering new, entire phoneme inventories one language at a time. Any new phonemes are added to the existing set and the new count recorded (as in Figure 2 above). If the phoneme type counts predicted by Herdan-Heaps' law match those of the known sample of inventories, we can then extrapolate the trend to larger number of inventories. For example, how many phonemes potentially exist in the full (but only partially documented) set of the world's 7151 spoken languages.^[This estimate is taken from Ethnologue ( https://www.ethnologue.com/guides/how-many-languages?).] What about even more languages? How many phonemes may have existed throughout time? And what is the upper bound on the number of distinct phonemes? Is there one? Or does Herdan–Heaps' law suggest that the number of phonemes in languages is potentially infinite?

<!-- The outline of this research report is as follows. In the next section, we describe the data and then present our methods and results. First, we show that the segment type distributions in both the UPSID and PHOIBLE language samples follow Zipf's law. This helps us motivate a connection to Herdan-Heaps' law and provides initial support for infinite phonemes. Second we evaluate the segment distributions in each database against Herdan-Heaps' law. Third, as a point of independent evidence to evaluate the findings of Herdan-Heaps' law, we apply a second estimator of type counts, which we refer to as the Chao-Shen estimator [@ChaoChiu2016]. The Chao-Shen estimator was developed to estimate the richness of species in biology from limited sample sizes, but we apply it to phonological inventories. Fourth, we compare how each estimator performs when predicting the number of phonemes in subsets of the language samples in PHOIBLE. This step allows us to evaluate the various count estimates against known data. Fifth, we compare the results from Herdan-Heaps' law and the Chao-Shen estimator and show that (a) the Chao-Shen estimator "seeks out" a curve described by the Herdan-Heaps' equation and (b) evidence from both estimators suggests that there is no upper bound on the number of segment types. Lastly, we use both Herdan-Heaps' and the Chao-Shen estimator to predict the number of segment types in increasingly larger -- and undocumented -- samples of languages. To conclude, we discuss why we expect that ever increasing linguistic diversity leads to an ever increasing number of phonemes and how this work has broad implications for inferring the number of types of other linguistic elements more generally. -->

# Data

For this study, we use the latest PHOIBLE data, which includes 3020 phonological inventories from 2186 distinct languages, covering 3183 distinct phonemes [@MoranMcCloy2019]. The dataset is a convenience sample of languages, in which some geographic regions and language families have greater or lesser coverage. Included in PHOIBLE is UPSID-451 [@Maddieson1984;@MaddiesonPrecoda1990], a genealogically balanced sample of languages in which each data point was selected as a representative of its language family. Therefore, we use both UPSID and PHOIBLE as baselines to evaluate our models.

Our central question concerns how the number of spoken languages that are documented relate to the number of unique phonemes that are observed. This is a question of cumulative growth. It can be rephrased as: if we have seen *x* languages with *y* unique phonemes, how many unique phonemes do we expect to find in $x + i$ languages?

# Methods and results

## Zipf's law

First, we test whether the segment distribution in the databases follow [Zipf's law](https://en.wikipedia.org/wiki/Zipf%27s_law). Figures 3 & 4 plot log frequency (y-axis) by log frequency rank (x-axis) for UPSID and PHOIBLE, respectively. The circles represent the cumulative distributional counts of each phoneme. The blue dotted line represents predictions based on Zipf's distribution using the following equation:

$$
\begin{aligned}
  f(rank;a;b;c) & = \log{\frac{c}{(rank+b)^a}} \\ 
  &= \log{(c) + a \cdot\log{(rank + b)}}
\end{aligned}
$$

where $a$, $b$, and $c$ are tunable parameters. In this case, we determined the optimal parameters using non-linear least squares regression as implemented in the *nls* function from the *stats* package (v. 3.6.2) for the programming environment *R* (*R* Core Team, 2019).

```{r upsid_zipf, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}Cumulative segments in UPSID-451 follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/zipf_distribution_upsid-1.png")
```

```{r phoible_zipf, echo=FALSE, fig.align="center", fig.cap="Cumulative segments in PHOIBLE (OIPL) follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/zipf_distribution_phoible-1.png")
```

The close fit between the observed curve and the predictions drawn from a Zipfian distribution suggests that global phoneme distributions are Zipf-distributed.^[We note the debates in the literature on what the underlying distributions are of phoneme inventory size (e.g., @Lehfeldt1975, @JustesonStephens1984, @Maddieson2008) and of phonemes frequencies in text [@Martindale_etal1996;@TambovtsevMartindale2007;@Macklin-CordesRound2020]. But here we are specifically looking at presence of phonemes in phonological inventories cross-linguistically.] Note that this close approximation holds for both the genealogically balanced and large-scale databases.

## Herdan–Heaps' law

The distribution of phonemes in both language samples are clearly Zipfian. We next test whether their distributions likewise follow Herdan–Heaps' law. Does the number of yet-to-be-seen segment types decrease as the number of phonological inventories observed increases by way of a power law relation?

An important feature of the Herdan-Heaps' equation is that it can be expressed in the form of a standard linear equation by simply taking the logarithm of each term, or:

$$
\begin{aligned}
  \log d & = \log (K + n^B)\\
  & = \log K + B \cdot \log n
\end{aligned}  
$$

which has the familiar form:

$$y = \beta_0 + \beta_1 \cdot x$$

This gives us the added benefit of being able to fit a log-transformed linear regression model to the data, such that:

* The resulting intercept is the optimal $K$ (with some back-transformation: $K = e^k$).
* The coefficient $\beta_1$ expresses $B$.

With these values, we can estimate the type count for any sample size $n$ using the standard Herdan-Heaps' equation. As proof of concept, we estimate the number of phonemes for both UPSID-451 and PHOIBLE. Some languages in PHOIBLE are associated with more than one doculect. To avoid an overrepresentation bias, we randomly sample one doculect from every language or take the single doculect where only one exists (one inventory per language, or OIPL. Results of a preliminary analysis indicated that the full and OIPL versions of PHOIBLE behaved similarly. We therefore focus on the results from the OIPL database here (see the supplementary materials for comparisons between the full and OIPL versions).

We begin by randomly permuting the order of inventories and computing the cumulative number of unique segment types for each new language observed. Then we repeat this process $k$ times (we set k=10 trials) for each database. We fit a log-linear regression to the resulting sets of estimates to generate empirically optimal values for $B$ and $K$ -- our two free parameters in the Herdan-Heaps' equation.

Figures 5, 6 & 7 show the predicted values against the true values of the number of segment types (phoneme type count) versus segments recorded in phonological inventories in the samples (segment token count). Model fits for the UPSID-451 and PHOIBLE language samples suggest that the phoneme type versus token count closely follows Herdan-Heaps' law.^[In the supplementary materials, we also fit a "one inventory per language" (OIPL) model of PHOIBLE to show that the language sample's genealogical and areal biases, e.g., the sample is more representative of African languages than other world areas, are not affected by simply sampling the whole database. The outcomes between a stratified random resampling and testing the segment type versus token on the entire convenience sample are nearly the same.]

```{r upsid_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}Cumulative segments in UPSID-451 follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/zipf_distribution_upsid-1.png")
```

```{r phoible_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}: Cumulative segments in PHOIBLE (OIPL) follow Zipf's law"}
knitr::include_graphics("README_files/figure-gfm/zipf_distribution_phoible-1.png")
```

Next, we compare the model parameters across the two databases, UPSID-451 and PHOIBLE. The values for $B$ and $K$ are relatively similar. UPSID-451 and PHOIBLE (all) have nearly identical values for B, but they differ in the scalar K, which explains the sharper increases in segment type counts for PHOIBLE.

| Database | B | K |
|----------|---|---|
| UPSID | 0.4994352 | 8.087208 |
| PHOIBLE (OIPL) | 0.5442040 | 5.870403 |

We illustrate these parameter differences in Figure 7 by combining the plots on the same scale. The curve describing Herdan-Heaps' estimates based on UPSID-451 has a shallower trajectory than the curve based on PHOIBLE. Nevertheless, both curves are highly similar. This similarity indicates that the known areal and genetic biases present in PHOIBLE do not have much of an effect on the modeling of phoneme diversity. Based on this finding, we focus on the larger and more complete database, PHOIBLE, in the remaining analyses.

```{r compare_models, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}: Cumulative segments in UPSID-451 and PHOIBLE (OIPL) follow Herdan-Heaps' law"}
knitr::include_graphics("README_files/figure-gfm/compare_models-1.png")
```

Next, we implement a general and flexible estimator for type counts called the Chao-Shen estimator [@Chao_etal2009;@ChaoChiu2016]. If the results from the two estimators (Chao-Shen and Herdan-Heaps' law) converge, then we have a more solid foundation for extrapolating phoneme counts for larger, hypothetical sample sizes.

## The Chao-Shen estimator

The Chao-Shen estimator was developed to determine the diversity of species within an ecosystem given only limited (spatial and temporal) sampling [@Chao_etal2009;@ChaoChiu2016]. It allows us to extrapolate the number of types in some population based on an incomplete sample using the number of rare types and an estimate of the number of unobserved types.

This estimator takes as input an incidence matrix and returns the expected number of types in a larger (imaginary) sample. An incidence matrix is a record of the presence or absence of the types being sampled from a set of sampling units. For example, imagine that an entomologist is interested in identifying how many types of beetles are in a forest. She knows from prior research that there are 1000 documented species of beetles in the area. In order to get a representative sample of the diversity of beetle species in the overall ecosystem, she sets up 50 beetle traps, regularly spaced across the forest. In this example, the traps are the sampling units and the species of beetles are the types. The entomologist prepares a 50 X 1000-cell spreadsheet in which the rows and columns correspond to the traps and known beetle species, respectively. She then visits each trap and for each of the 1000 known species of beetle she finds, she records a 1 for that species in appropriate cell. All unseen species are given a value of 0. The result after all traps have been visited is a 50 X 1000 incidence matrix. The information contained in this matrix can then be fed into the Chao-Shen estimator to provide an estimate of the true diversity of beetle species in the forest.

The same approach can be extended to languages and their phonemes. On analogy with the beetle example, the forest is the sum of the human-populated areas of Earth; the beetle traps are all documented human languages; and the beetles are the phonemes which appear in those languages. The incidence matrix is of shape *l* X *s*, where *l* is the number of documented languages and *s* is the number of types of segments that have been documented as phonemes in at least one language. By definition, if a segment appears as a phoneme in a given language, the corresponding cell in the incidence matrix contains a 1; otherwise, it contains a 0.

Table 1 shows a hypothetical example of an incidence matrix for 150 observed segments from a set of $S_{obs}$ phonemes in $R$ languages. Marginal sums are given per language ($\sum_{language}$, or *inventory size*) and per segment ($\sum_{phoneme}$, or *cross-linguistic prevalence*)

|Phoneme|Lang<sub>1</sub>|Lang<sub>2</sub>|&middot;&middot;&middot;|Lang<sub>*R*</sub>|&sum;<sub>phoneme</sub>|
|:---:|---|---|:---:|---|---|
|**Phon<sub>1</sub>**|1|0|&middot;&middot;&middot;|0|1|
|**Phon<sub>2</sub>**|0|1|&middot;&middot;&middot;|1|2|
|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|&middot;&middot;&middot;|
|**Phon<sub>*S*<sub>obs</sub></sub>**|1|1|&middot;&middot;&middot;|1|30|
|**&sum;<sub>language</sub>**|15|23|&middot;&middot;&middot;|6|150|

Table: Hypothetical sample of an incidence matrix

Using the definitions provided above, the Chao-Shen estimator is expressed formally as follows:

$$ \hat{S}_{\text{Chao-Shen}} = S_{\text{obs}} + \hat{Q}_0 \left[1 - \left(1 - \frac{Q_1}{Q_1 + R \cdot \hat{Q}_0}\right)^r\right] $$

$S_{obs}$ is the number of segment types that have been observed overall. $R$ is the number of languages in the observed sample. $Q_1$ is the number of segment types that have only been observed once (*singletons*), e.g., Phon<sub>1</sub> in Table 1). The hyperparameter $r$ expresses the number of new languages that we wish to add (hypothetically) to our existing sample. Finally, $\hat{Q}_0$ is the estimated number of unobserved types that were nevertheless "truly" present. We do not have empirical access to this figure -- we cannot directly measure how much we haven't seen. Therefore, it must be inferred. $\hat{Q}_0$ can be derived in many ways. Here, we use the following equation.

$$ \hat{Q}_0 = \begin{cases} S_{\text{obs}} + \left[\frac{R-1}{R}\right] \frac{Q^2_1}{2 \cdot Q_2} & \text{if }Q_2 > 0 \\ S_{\text{obs}} + \left[\frac{R-1}{R} \right] \cdot Q_1 \cdot \frac{Q_1-1}{2} & \text{if }Q_2 = 0\end{cases}$$

In this equation, $Q_2$ stands for number of segment types which were observed exactly twice, or *duplicates* (e.g., Phon<sub>2</sub> in Table 1).

Similar to the Herdan-Heaps' equation, we can test how well the estimates of the model match known values based on increasing sample sizes. However, unlike Herdan-Heaps', the Chao-Shen estimator requires a hyperparameter that stipulates how many *additional* observations it should consider in making its estimate. It looks ahead to larger samples, rather than at the current sample size. We must therefore approach this problem differently. Our solution is to sample repeatedly from our set of languages, each time increasing the number of languages $R$ by an increment of $d$. For each value of $R$, we compute the Chao-Shen estimates for several values of $r$, each time increasing $r$ by $c$ additional observations.

The entire procedure is thus:

1. Pick a random set of phonological inventories of size $R_s$
2. Compute the frequency vector of the segment types
3. Apply the Chao-Shen estimator for $r$ with $R = R_s$
  3a. Increase $r$ by $c$ 
  3b. Repeat 3a until maximum $r_{\text{max}}$ is reached
5. Repeat steps 1-3 $k$ times
  5a. Increase $R_s$ by $d$

Figure 8 shows the results of applying this procedure to PHOIBLE. $k$ is arbitrarily set to 10 iterations. The relatively small degree of variation across iterations suggests that this value is suitably high. *d* was set to 10. *r* was manually adjusted to exemplify behavior at different scales (within the limits afforded by the overall sample size). Grey lines represent the Chao-Shen estimates for each iteration, and blue lines give the means. Red lines represent the mean of the true number of observed types across iterations. Red lines become shorter as *r* increases because we have fewer true values to compare against when we look further ahead from any position (i.e., fewer values of *R* for which we can look *r* steps ahead).

```{r chao_lang_types, echo=FALSE, fig.align="center", fig.cap="\\label{fig:chao_lang_types}Estimated segment types by number of languages"}
knitr::include_graphics("README_files/figure-gfm/chao_lang_types-1.png")
```

The first thing to notice about these plots is that they closely resemble the Herdan-Heaps' curve that we generated from PHOIBLE. These plots thus also point to slowing but non-asymptotic growth of type counts.

Regarding the behavior of the parameters, the Chao-Shen estimator shows near-perfect prediction across values of *R* when *r* is relatively small. That is, the model is very good at predicting known segment type counts for samples with relatively fewer additional languages (small *r*), regardless of the number of observed languages (all *R*). If we compare the panels against one another, particularly at lower values of *R*, we see that the differences between the true values (red lines) and the estimated values (blue lines) increase with *r*. As *r* increases, so does the degree of underestimation. In other words, the model is more conservative in its guesses when data are sparse and the distances of extrapolation are large. The underestimation appears to dissipate around *R* = 1000. Figure 9 plots this relationship directly.

```{r chao_error_estimates, echo=FALSE, fig.align="center", fig.cap="\\label{fig:chao_error_estimates}Chao error estimates in PHOIBLE"}
knitr::include_graphics("README_files/figure-gfm/chao_error_estimates-1.png")
```

Variability across iterations increases with *r*, particularly for sample sizes of ~1700 languages or fewer. This pattern indicates that larger values of *r* should be compensated for by larger values of *R*. Otherwise, the results may not generalize well to other samples (even if the other samples are drawn from the same underlying pool, as we have shown here!).

To summarize, estimates converge on the observed values earlier on average when *R* is larger relative to *r*. Thus, we do better when we have more data and predict values closer to our observed sample. In addition, the variability in estimates across iterations (grey lines) increases with *r* and decreases with *R*. Thus, our estimates are noisier when we look further ahead and have smaller samples. Together these findings suggest two things: (a) the ratio of observed sample size to sample size increase $\frac{R}{r}$ should be maximized to reduce sample-induced noise, and (b) *R* should be maximized to fight the underestimation bias.

Unfortunately, PHOIBLE offers a proportionally small *R* given some of the values of *r* that we would like to investigate. For example, to account for the estimated number of current languages, we would need *r* = 7151 - 2099 = 5052, a value almost 2.5 times as large as our observed *R* (2099). Extrapolating from our known sample to the estimated number of world languages therefore requires an undesirably small proportion $\frac{R}{r}$ -- around 0.42. That said, our data suggest that sample sizes of ~1000 languages produce estimates that closely approximate the true segment type counts, even as *r* increases. In other words, after 1000 languages, increases in *R* do less to mitigate the underestimation bias. Given that PHOIBLE contains more than double that number of languages, we have no reason to suspect significant underestimation for the relatively large values of *r* that we investigate here. However, because of the small ratio of *R*:*r*, we should expect our estimate to come with greater uncertainty.

We have so far described the distributional properties of our databases, associated their behavior with two linguistic laws, and introduced two methods for inferring type counts from an observed frequency distribution. We now explore in more detail how well the two type estimators -- Herdan-Heaps' and Chao-Shen -- perform when predicting the total segment type diversity of a database from an incomplete subsample.

## Predicting known segment type diversity

To establish whether we can reliably estimate the number of segment types in a known sample, we evaluate the accuracy of each estimator in predicting the known type count given a fixed-size random subsample. From PHOIBLE (*n* = 2099), we randomly select 1500 languages – an arbitrary number that includes roughly three-quarters the number of languages in the total sample. Note that this value was set above the threshold of *R* = 1000 (reported above for the Chao-Shen estimator) to help support prediction accuracy.  We then extrapolate our model parameters from this random sample and use them to predict total segment type diversity in PHOIBLE. We iterate this process *k* = 100 times.

The Herdan-Heaps' estimates are visualized in Figure 10. The mean is plotted as a solid black line. The 97.5% confidence intervals (CIs) are plotted as dotted lines around the mean. The red dotted line indicates the true number of phoneme types. Blue shading indicates the density function of the estimated values from the 100 iterations.

For this run of 100 iterations, the Herdan-Heaps' estimator overestimated the true value by ~18 segment types on average (~2613 estimated vs. 2596 observed).

```{r hh_ci, echo=FALSE, fig.align="center", fig.cap="\\label{fig:herdan_heaps_ci}Herdan-Heaps' confidence intervals for a sample of 1500 languages"}
knitr::include_graphics("README_files/figure-gfm/herdan_heaps_ci-1.png")
```

Next we test the Chao-Shen estimator. We again sample 1500 languages 100 times from PHOIBLE. The results are plotted in Figure 11.

```{r chao_ci, echo=FALSE, fig.align="center", fig.cap="\\label{fig:chao_ci}Chao confidence intervals on a sample for 1500 languages"}
knitr::include_graphics("README_files/figure-gfm/chao_ci-1.png")
```

The Chao-Shen estimator overestimates the true value by only 1 type (the estimate is actually closer: 0.56 types). It therefore provides an almost perfect fit. Note that $\frac{R}{r} = \frac{1000}{1099} = 0.91$ (three times the size of what we face with our extrapolation to the world's 7151 known languages).

The Chao-Shen estimator therefore seems to outperform Herdan-Heaps' law in predicting type counts in PHOIBLE. However, we stress that predictions from both models fall remarkably close to the true value.

As a final point of comparison, we plot Herdan-Heaps estimates against the corresponding Chao-Shen estimates for increasing sample size and values of *r* (Figure 12).

```{r comp_plot, echo=FALSE, fig.align="center", fig.cap="\\label{fig:comp_plot}Comparing the Herdan-Heaps' and Chao esimtator"}
knitr::include_graphics("README_files/figure-gfm/comp_plot-1.png")
```

The black line represents PHOIBLE-based Herdan-Heaps estimates. Grey points reflect true cumulative type counts from 10 reshuffled versions of PHOIBLE. The horizontal dotted red line indicates the total known number of segment types in PHOIBLE. The solid colored lines represent Chao-Shen estimates for different *r* (indicated in the legend). Note that these lines have been shifted along the x-axis to reflect that Chao-Shen estimates look ahead, such that the $\hat{R}$ being estimated is actually *R* + *r*.

This plot demonstrates several of the points raised above. The Chao-Shen estimates at small *R* and large *r* are gross underestimates of the true value. However, as *R* increases, the estimated values climb quickly towards the true values. What is striking about this plot is that the Chao-Shen estimates seem to "seek out" the Herdan-Heaps curve. For these data, the Chao-Shen estimator has decided that segment type counts grow according to a rate predicted by Herdan-Heaps' law. This need not have been the case; the Chao-Shen estimator is capable of predicting values produced by many different kinds of functions. Finally, and most crucially, neither estimator shows any indication of leveling out: both appear to be non-asymptotic.

## Extrapolating estimates to unseen numbers of observations

We have so far established that (a) both the Herdan-Heaps' estimator and the Chao-Shen estimator predict unseen type counts accurately and (b) the two ultimately agree on these estimates, though Chao-Shen needs more data to "catch up." We have also shown that the Chao-Shen estimator requires two things to perform well: large ratios of *R* to *r* and generally large *R*. With this in mind, we now apply both estimators to unseen sample sizes. We start with the number of known world languages (7151), then proceed to larger values (10k, 50k, and 100k).

### Herdan-Heaps' law

Recall that the Herdan-Heaps' equation requires that we know three things: two free parameters, *K* and *B*, and the number of observations *n*. As before, the parameters are optimized using a log-transformed linear regression over cumulative type counts from PHOIBLE. However, determining the third variable -- number of observations -- for an arbitrarily large sample poses a problem. Suppose you have a sample of 100 languages with 1000 total segments (tokens, not types), and you want to guess how many total segments you would find in a sample of 150 languages. Is it simply a linear increase, i.e., 1500? No, or rather we cannot know. We do not know how many sounds each language would contribute in an unseen population, so we cannot directly derive the requisite number of observations from the number of languages alone. Therefore, we need some way to estimate this figure.

We approach this problem in two ways. First, we sample inventories randomly with replacement from the observed dataset up to a desired number of languages. The steps are as follows:

1. Randomly sample *L* languages from the existing database with replacement
2. Sum the inventory sizes of the resulting sample of languages.
3. Repeat steps 1-2 *k* times

*L* corresponds to the desired number of languages. For example, setting *L* to 7151 will produce the number of segments (tokens) we should expect to see if we had access to the inventories of all known human languages.

Second, we regress the number of total phonemes on sample size using the observed dataset. The steps are as follows:

1. Randomly sample *l* languages from the existing database with replacement.
2. Sum the inventory sizes of all languages in the resulting sample.
3. Increase *l* by increment *s*
4. Repeat steps 1-3 until desired number of languages (*L*) has been sampled.
5. Repeat steps 1-4 *k* times
6. Fit a regression to the sequences created
7. Use the regression to predict token counts for desired *L*

The result of steps 1-5 is *k* sequences of increasing token counts (by steps of *s*) for independent, random samples *l* languages up to the desired number of languages *L*. Steps 6-7 fit a linear regression to the set of \textit{k} sequences (with random intercept adjustments for the iterations). The resulting model is then used to predict the number of observations for any number of languages.

We apply both approaches to PHOIBLE for samples sizes of 7151, 10k, 50k, and 100k. The mean estimates for Approach 1 and 2 are presented in Table 2.

||7151|10K|50K|100K|
|---:|:---:|:---:|:---:|:---:|
|Approach 1|247623|346234|1731657|3465640|
|Approach 2|247620|346257|1731115|3462189|
|Difference|3|-23|542|3451|

Table: Mean estimates for Approach 1 & 2.

As sample size increases, Approach 1 produces increasingly larger estimates relative to Approach 2. At lower sample sizes, the values are comparable.

The values from Approach 1 and 2 can now be entered into the standard Herdan-Heaps' equation with *K* and *B* taken from the full-PHOIBLE regression model. Results are given in Table 3.

||7151|10K|50K|100K|
|---:|:---:|:---:|:---:|:---:|
|Approach 1|4850|5770|13282|19025|
|Approach 2|4850|5770|13280|19016|
|Difference|0.02|-0.20|2.15|9.81|

Table: Phoneme type counts for different sample sizes.

Both approaches behave similarly, even producing the exact same estimate for 7151 and 10000 tokens.^[The Supplementary Materials contain an analysis which suggests that most languages tend to settle on a similar number of phonemes, meaning that the overall number of observations can be neatly derived from the number of languages.]

### Chao-Shen estimator
Recall that the Chao-Shen estimator only requires a single parameter *r*, i.e., the number of additional languages that we wish to sample. To derive *r* for a target sample size, e.g., 7151, we need only subtract our observed *R*, or 2099, from that number. The resulting figure is then plugged into the estimator to get the expected type count $\hat{R}$. The full set of *r* values and corresponding type count estimates are given in Table 4.

||7151|10K|50K|100K|
|---:|:---:|:---:|:---:|:---:|
|$r$|5052|7901|47901|97901|
|$\hat{R}$|4592|5037|5530.86|5530.92|

Table: *r* values and corresponding type count estimates

Predicted type counts are generally lower than those produced by Herdan-Heaps' law, and this difference increases with *r*. Unlike the Herdan-Heaps' estimator, the Chao-Shen estimator does appear to reach an asymptote. Indeed, increasing *r* beyond 100K always results in the same value: 5530.92. The presence of this asymptote can be explained by the function of *r* in the Chao-Shen equation. *r* serves to exponentiate a term that falls in the range {0,1}. Therefore, as *r* increases, the full term shrinks to 0. The result is subtracted from 1, meaning that as *r* increases, the compound term $1-x^r$ approaches 1. This value controls how much of the estimated number of unobserved types $\hat{Q_0}$ is added to the number of observed types $S_{\text{obs}}$. Therefore, as the compound term approaches 1 per unit increase in *r*, the resulting type count estimates get closer and closer together. Looking too far ahead slows the growth of type count estimates to a crawl. In turn, this slowdown produces increasingly more severe underestimates (as outlined above). The apparent asymptote can therefore be interpreted as a mechanical feature of the Chao-Shen estimator rather than a substantial property of global phonological distributions.

Figure 13 summarizes the segment type estimates of the two Herdan-Heaps' methods and the Chao-Shen estimator.

```{r diff_plot_ext, echo=FALSE, fig.align="center", fig.cap="\\label{fig:diff_plot_ext}Type count estimates from the Chap and Herdan-Heaps' estimators for four sample sizes"}
knitr::include_graphics("README_files/figure-gfm/diff_plots_ext-1.png")
```

The two approaches to estimating *n* for Herdan-Heaps are nearly identical. Furthermore, we see diminishing rates of increase of the estimates (going from 10K to 50K results in a greater increase than going from 50K to 100K). For smaller *r*, the Chao-Shen and Herdan-Heaps' estimators behave similarly. As *r* increases, the degree of difference between the two estimators increases (as predicted based on the underestimation bias reported above for Chao-Shen).

To summarize our main findings: (a) Herdan-Heaps' law fits the known phonological diversity well, (b) a different strategy for estimating type count performs best when it mimics Herdan-Heaps' law, (c) we therefore have no reason to believe that there is a limit to the number of phonemes to be observed in human languages, and (d) even if limited, the true number is most likely much larger than what we have documented so far.

# Discussion

The present study asks a seemingly straightforward question: how many different phonemes are possible in human language? We cannot determine this number empirically from existing resources because the vast majority of languages, known or unknown, are undocumented, extinct, or both. Moreover, differences in documentary practice (theoretical and/or methodological) mean that even our existing phonological descriptions are provisional and objects of debate. For example, PHOIBLE contains approximately 1.4 different inventories per language.^[Different inventories per "language" mean they do not agree on the number or exact repertoire of phonemes. However, this should be expected given that phonological descriptions typically represent different doculects, i.e., different instances of linguistic behavior at different places, times, or with different speakers [@GoodCysouw2013] and that phoneme analysis is a non-deterministic process [@Chao1934;@Hockett1963].]

To address these issues, we modeled the distributional properties of documented phonemes. First, we found that frequencies of phoneme types in existing cross-linguistic databases are well described by Zipf's law, regardless of whether samples were controlled for areal and genetic biases. Systems that obey Zipf's law, under mild assumptions, also follow Herdan-Heaps' law [@BaezaNavarro2000;@vanLeijenhorst2005], which predicts type count from the number of tokens with only two tunable parameters. Herdan-Heaps' law was originally intended to describe how vocabulary size increases as more words are observed in a document or set of documents. Extending this approach, we demonstrate that the number of phoneme types in a sample of languages also grows according to Herdan-Heaps' law. To our knowledge, we are the first to apply Herdan-Heaps' law to the domain of segmental phonology by reformulating it as estimator of the number of phonemes in groups of languages. Finally, we verify the goodness-of-fit of Herdan-Heaps' by applying an additional and formally unrelated tool borrowed from Ecology, which we refer to as the Chao-Shen estimator. The predictions of this estimator confirm that the relationship between sample size and segment type count is best described by a curve that approximates Herdan-Heaps' law. Most importantly, the optimal Herdan-Heaps' model did not converge on any single number of phoneme types as number of languages increased.

Assuming a target of 7151 known languages,^[https://www.ethnologue.com/guides/how-many-languages?] our models predict a range of total segment types from 4592 to 4850 (mean = 4721). Based on PHOIBLE, our current best guess of the diversity of documented phoneme types is 3183. Therefore, the models argue for the existence of approximately 1500 living but undocumented phoneme types. As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30k-500k [@Pagel2000] or 64k-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19K. The Chao-Shen model produced much more conservative estimates for samples of size 10K and above, even appearing to settle on an asymptote of roughly 5500 phonemes for samples of 100K languages or more. However, when we compare the predictions of Herdan-Heaps and Chao-Shen  against known type counts in a closed dataset, the source of the discrepancy emerged: as the sample grows, Chao-Shen lags behind Herdan-Heaps in approximating the true value. Thus, given a larger pool of training data, we expect Chao-Shen to approximate more closely the estimates of Herdan-Heaps.

We therefore provisionally conclude three things: (1) adding languages to a sample will always carry a non-zero probability of producing a novel phoneme; (2) in accordance with Herdan-Heaps' law, the rate at which new phonemes are discovered per language will always decrease; and (3) the decrease in the rate of new observations will never reach 0. Thus, based on the evidence presented here, we have no reason to believe that there is an upper bound on the number of unique phonemes possible in spoken languages. Infinite linguistic diversity seems to breed infinite possible phonemes. But what would it mean if there were indeed infinitely many potential phonemes?

If one adopts a feature-based perspective on the phonetic targets of phonemes, then there is an *a priori* upper bound of $2^F$ contrastive sounds, where $F$ is the number of feature categories and 2 reflects the choice between presence (+) or absence (-) of that feature.^[In some feature sets, some values may be 0, i.e., undefined, underspecified, or ``for which a segment can be said 'not to care'" [@Hayes2009, pg. 91]. For example, speech sounds that are -coronal are underspecified for the features +/-anterior and +/-distributed, which capture coronal distinctions for places of articulation involving (inter-)dentals and alveolars vs palatal alveolars and retroflexes (anterior), and laminal vs apical coronals (distributed).] @Clements2009 coined this calculation *Feature Bounding*. The expected number of unique phonemes is controlled by the number of features, where additional features increase the pool of possible distinctive segments (nearly) exponentially. For example, given a feature set such as in PHOIBLE with 37 distinctive features, an upper bound on the number of segments under a strictly binary formalization is over 137 billion (2^37) feature vectors.

<!-- This type of feature based constraint limits the upper bound on the number of contrastive sounds in spoken language, yet @Clements2009 also defines an upper limit on the number of possible phonemic contrasts. That is, phonemic contrast is set by the number of features, such that C = (S * (S - 1)) / 2 [@Clements2009, pg. 25]. Thus, given a maximum number of speech sounds (e.g., 2^37), ... -->

However, most widely accepted feature sets contain features that are specified at different levels of phonetic detail: some features are more abstract (i.e., general similarities such as +/- syllabic), while others are more concrete (i.e., closer to the articulatory reality, e.g., advanced tongue root). Some are grounded in pure articulatory or acoustic phonetics, while others involve phonological properties of language. And looming in the background of these discussions, the debate of whether distinctive features are innate and universal or rather learned and emergent continues. Nevertheless, in general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors ($2^24$). Current consensus aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature. Making finer and finer contrasts may always produce more and more features. Thus, the exponent in the Feature Bounding calculation can itself (by hypothesis) be infinite, in line with the predictions of the models presented here, and therefore with the notion of infinite possible phonemes. That said, contemporary feature sets can capture phonological classes for place and manner of articulation, which neatly classify the phonological patterns and processes in and across languages (cf. @Mielke2008).

<!-- Perhaps the enterprise of distinctive features and their mapping to cognitive or universal categories is... -->

<!-- It is phonological contrast, i.e., the distinction within a language that convey its different lexical and grammatical meanings. In terms of universalitity, all languages -- spoken or signed -- rest on this seemingly simple notion of contrast. Thus, linguistic contrast is universal, and whichever sets of features defined to capture these distinguishable articulatory and/or acoustic -->

We have so far seen that the feature space is a discrete re-coding of the continuous variability of the physical acoustic and articulatory spaces -- and so it should by way of finer and finer specification, be able to support an infinitely diverse set of phonemes. However, one objection to this account is that humans are subject to physiological limits, both in production and perception. In speech production, articulatory gestures are executed with some degree of imprecision [@Ladefoged_etal1972;@Lindblom1990]. If this stochastic variability is greater than the difference between two or more phonetic targets, then no reliable distinction can be sustained (cf. @Stevens1972). In comparison, if the range of articulatory gestures is greater than their acoustic (or perceptual) reception, there is a "many-to-one" mapping between articulation and perception. In perception, the ability to discriminate acoustic stimuli is also limited both anatomically and cognitively. Differences in the signal that fall below the various thresholds for discrimination are thus ignored.

In response to these physical constraints, languages tend to spread their phonological inventories across the articulatory and acoustic spaces, thereby maintaining workable discriminability. A stark example comes from Central Rotokas (North Bougainville; Papua New Guinea), which has only 6 consonants (9 counting allophones) from three easily discriminated places of articulation: labials `[p, b]`, alveolars `[t, d]`, and velars `[k, g]` [@FirchowFirchow1969]. Compare this to a hypothetical less-optimal language that contrasts only `[t, d]`, `[tʰ, dʰ]`, and `[t̪, d̪]`. Given this strategy, it is unlikely that ultra-fine-grained phonetic contrasts would ever arise in a single language. Rather, the contrasts are most likely to arise between languages. In other words, the infinity of phonemes predicted by the analyses reported here plays out in the realm of cross-linguistic variation of phonetic targets, even if those differences cannot be reliably articulated by a single tongue or perceived by a single ear.

Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond whatever the phonetics-phonology interface may be. Thus, we believe the approach presented here has broad applications to the study of other areas of grammar and the cross-linguistic study of languages.

# Data accessibility

Data and code required to reproduce all results presented in this paper are available as electronic supplementary material at: https://github.com/bambooforest/heaps.

# Competing interests

The authors declare no competing interests.

# Acknowledgements

Steven Moran was funded by the Swiss National Science Foundation (Grant No. PCEFP1_186841).

# References

### 2023-06-06 paper.tex
initial commit

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fontspec} \setmainfont{Times New Roman} \setmonofont{Arial Unicode MS} \usepackage{lineno} \linenumbers
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={How many speech sounds are there in the world's languages?},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{How many speech sounds are there in the world's languages?}
\author{Steven Moran \& Nicholas A. Lester}
\date{06 June, 2023}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

How many speech sounds are there in the world's languages? To answer
this question, we extend two formal models of lexical diversity to
segmental phonology. Zipf's law states that most word types are rare;
Herdan-Heaps' law states that vocabulary size grows indefinitely with
increasingly new text. We find that both models produce accurate
estimates of the diversity of contrastive speech sounds (phonemes) in
existing typological databases. Our findings suggest that (a) most
phonemes are difficult to detect because they are rare, and (b)
documenting more phonological repertoires always yield more phoneme
types cross-linguistically. As a sanity check, we apply an unrelated
estimator of species diversity from biology, Chao-Shen, and show that
its estimates align with Herdan-Heaps' law. We use both models to
extrapolate phoneme counts to undocumented languages, resulting in an
estimate of 4600-4900 distinct phonemes in the full set of languages
thought to be spoken today (\textasciitilde7-8k), which is roughly 1.5x
more than what is currently documented in the world's languages
(\textasciitilde35\%). By extrapolating to hypothetical sample sizes,
e.g., the number of languages estimated to have been spoken through
time, and beyond, we find no indication of convergence: infinite
languages mean infinite phonemes.

\emph{Keywords}: speech, phonetics, phonology, language, Zipf's law,
Herdan-Heaps' law

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Under the umbrella of linguistic typology, studies of language diversity
aim to catalog the range of variation in comparable structures across
the world's languages. For some areas of variation, the problem is moot,
i.e., the full range of options is known \emph{a priori}. For example in
grammar, the relative ordering of a verb (V) and its direct object (O)
allows for exactly three possible types: OV, VO, and OV/VO. Thus, every
language falls into one of these categories. Other linguistic domains
are not logically constrained in this way because the true diversity of
types is inherently unknown.

Consider the set of possible speech sounds in the world's languages.
Phonemes are distinguished according to their articulatory, acoustic,
and perceptual properties. However, these properties exist in a
continuous temporal and physical space, which can be carved up in
infinitely many ways. For example, at suitably fine-grained levels of
measurement, no two realizations of the same phoneme will ever be
identical in every dimension of variation.

Thus, each new word presents a non-zero probability of containing a
previously unheard sound, and by extension, new phoneme. Complicating
matters, our current pool of ``consensus phonemes'' -- however defined
-- is subject to re-evaluation in response to advances in data
collection, methods, tools, theories, and experience that inform
language documentation and analysis.\footnote{In the last 20 years or
  so, grammars and phonological descriptions (e.g., \emph{Illustrations
  of the IPA}) increasingly include detailed data and acoustic phonetic
  analysis using software, such as Praat
  (\protect\hyperlink{ref-Boersma2001}{Boersma 2001}). This makes
  phonetic interpretation open to a wider audience.}

The problem is therefore to determine whether there is some sample size
of languages after which no amount of additional phonological
inventories will yield an unseen phoneme type. The absence of a limit
would raise important theoretical issues for the description of
phonemes, including questions concerning the role of phonetics in
phonology (\protect\hyperlink{ref-Whalen2019}{Whalen 2016}) and the
importance of cross-linguistic variation
(\protect\hyperlink{ref-Hockett1963}{Hockett 1963};
\protect\hyperlink{ref-EvansLevinson2009}{Evans \& Levinson 2009}).

The present study seeks to determine whether such a limit exists. To
meet this goal, we propose several methods for modeling the relationship
between sample size and number of linguistic types. We apply these
methods to ask three core questions. First, how many phonemes are there
in the world today (based on a limited sample)? Second, how many
phonemes have there been in the history of human language? And lastly,
how many phonemes are possible in human language? In Section 2, we
motivate our research questions. In section 3, we describe the data and
in Section 4 we present our methods and results. In Section 5, we
discuss our results and avenues for future research.

\hypertarget{motivation}{%
\section{Motivation}\label{motivation}}

The question of whether there are limits on the number of distinct
speech sounds that can appear in spoken languages has been recognized
for some time. Alexander J. Ellis (1814--1890) stated as early as 1848
that the number of different speech sounds, i.e., candidate phonemes, is
in principle infinite. At that time, teachers and philologists aimed to
develop a transcription standard that could discriminate between all
manners of speech sounds with orthographic symbols for purposes that
included (particularly English) spelling reform and orthography
development for newly documented and unwritten languages. But as
practitioners increasingly converted the articulations of more and more
languages into a universal phonetic alphabet, the need for new symbols
also seemingly increased without limit
(\protect\hyperlink{ref-Hockett1995-phoneme}{Hockett 1995}, pg.
63).\footnote{This early work later became the basis for the
  International Phonetic Association (and its phonetic alphabet; both
  abbreviated IPA). First published in 1888, today the IPA provides and
  maintains a speech transcriptional device that encodes aspects of
  articulation, and acoustics, for both narrow (phonetic) and broad
  (phonemic) transcription. It is comprised of a set of alphabetic
  symbols and diacritics with the aim of representing all possible
  sounds in the world's spoken languages (there are extensions for
  disordered speech and speech pathology). The IPA provides a summary of
  linguists' phonetic knowledge. This system uses a set of ``base''
  symbols, e.g., letters or combination of letters, that aim to encode
  phonemes, and the use of diacritics on base glyphs to provide phonetic
  detail. Its design also affords infinite transcription possibilities.}
Finer and finer distinctions in phonetic detail created more and more
space for novel phonemes to emerge. Thus, if the phonetic distinctions
to be drawn are infinite, then tautologically so are the number of
possible phonemes.

If there was any doubt about the infinite nature of the speech signal,
instrumental phonetics put the nail in the coffin. Jean-Pierre Rousselot
(1846--1924) studied acoustic properties of speech with the kymograph,
originally developed to measure physiological patterns such as blood
pressure. The later advent of the phonograph by Thomas Edison
(1847--1931) in 1877 provided a means to record and reproduce sound.
Then during WWII the sound spectogram was developed in secret to analyze
ciphony by making sounds visible. The work on visible speech was
released publicly after the war
(\protect\hyperlink{ref-Potter1945}{Potter 1945}) and the spectrogram
became the staple of acoustic phonetic analysis beginning with Joos
(\protect\hyperlink{ref-Joos1948}{1948}) (Joos was the first linguist to
have extensive access to the acoustic spectrograph). For speech, it
measured fluctuations in air pressure, which capture duration, pitch,
and intensity. The recorded speech waveforms showed visually that speech
is dynamic and continuous, even if listeners perceive it as discrete,
e.g., phonemes, syllables, words.\footnote{This relationship between
  continuous and discrete is at the core of the debate about the
  interface between phonetics and phonology
  (\protect\hyperlink{ref-Whalen2019}{Whalen 2016}). The division goes
  back to at least the late 19th century when phoneticians and
  dialectologists including Henry Sweet (1845--1912; 1877) and Jost
  Winteler (1846--1929; 1876) began making a distinction in language
  descriptions between sounds that change the meaning of words and those
  that do not.}

Aside from the acoustic-phonetic argument, evidence for infinite
phonemes comes from studies on the cross-linguistic frequencies of
speech sounds. Early evidence is found in a large cross-linguistic
sample of phoneme inventories, where inventory is understood as the set
of distinctive sounds attested in a language. This sample is described
in the UCLA Phonological Segment Inventory Database (UPSID; Maddieson
(\protect\hyperlink{ref-Maddieson1984}{1984}), Maddieson \& Precoda
(\protect\hyperlink{ref-MaddiesonPrecoda1990}{1990})). Maddieson
(\protect\hyperlink{ref-Maddieson1984}{1984}) created a
genealogically-balanced sample of the world's languages (\emph{n} = 317)
to investigate cross-linguistic properties of phonological systems and
the relationship between aspects such as inventory size, contents, and
structure. Maddieson noticed that certain segments are more or less
likely to be present in inventories regardless of their size. That is,
certain sounds are almost always present and others are almost never
present. In other words, no segment occurs in every language and some
very frequent segments occur in most languages. Moreover, nearly half of
the reported segment types are singletons in the sample, i.e., one-off
language-specific occurrences.\footnote{Whether rare segments occur more
  or less frequently \emph{within} a language is a language-by-language
  research question. See for example
  (\protect\hyperlink{ref-Everett2018}{Everett 2018}).} While Maddieson
did not explicitly test for it, his observations fit a well-attested
pattern known as Zipf's law (\protect\hyperlink{ref-Zipf1936}{Zipf
1936}; \protect\hyperlink{ref-Zipf1949}{Zipf 1949}). Zipf's law states
that the frequency distribution of words in a language is inversely
proportional to their frequency rank. The relationship is log-linear:
high frequency words are few in number but account for most of the
tokens in a text, and low frequency words are many, but occur rarely or
even just once.\footnote{Here is a plot showing Zipf's law in the first
  10 million words in the Wikipedias from 30 different languages:
  \url{https://en.wikipedia.org/wiki/Zipf\%27s_law\#/media/File:Zipf_30wiki_en_labels.png}.}
, Zipf's law states that the frequency of an element \(f(r)\) is
proportional to the inverse frequency rank of that element (modulated by
an exponent \(\alpha\))
(\protect\hyperlink{ref-Piantadosi2014}{Piantadosi 2014}):\footnote{Piantadosi
  (\protect\hyperlink{ref-Piantadosi2014}{2014}) discusses in detail
  that word frequency distributions are not actually as simple as they
  seem according to Zipf's law. For example, semantics affects word
  frequency distributions. Moreover, observations of Zipf's law have
  been obscured by word frequency distribution plots that fail to
  address correlated errors. Zipfian distributions are not unique to
  language. They appear in various aspects of human society, such as
  music and aspects of the internet, as well as in other biological and
  physical systems. Strikingly, power law distributions can arise from
  nearly nothing (\protect\hyperlink{ref-Piantadosi2014}{Piantadosi
  2014: 1121}).}

\[
f(r) \propto \frac{1}{r^\alpha}
\]

The Zipfian distribution characterizes many aspects of language. Two
features make it interesting for our purposes. First, it describes
systems for which most types are rare, hence difficult to observe.
Second, under certain conditions (\(\alpha \ge 1\)), the Zipf function
is non-asymptotic: more types can always be added at the rank for
singletons, and the expected type count computed. The fact that
Maddieson's observations are consistent with the Zipfian distribution
suggests that phoneme types are indeed difficult to discover
(particularly at the fringe) and that there could be some room left for
additional phonemes.

Moran (\protect\hyperlink{ref-Moran2012}{2012}) revisited the frequency
distribution of segments in a larger sample of the world's languages
(\emph{n} = 1780, including UPSID) and showed that the relationship
between cross-linguistic segment frequency follows a similar pattern as
in the Maddieson (\protect\hyperlink{ref-Maddieson1984}{1984}) (\emph{n}
= 317) and Maddieson \& Precoda
(\protect\hyperlink{ref-MaddiesonPrecoda1990}{1990}) (\emph{n} = 451)
UPSID language samples. That is, some phonemes are very frequent across
the world's languages, and half of all documented phonemes are
language-specific. Moran (\protect\hyperlink{ref-Moran2012}{2012}) also
did not explicitly test whether cross-linguistic phoneme frequency
adheres to a Zipfian distribution,\footnote{Moran
  (\protect\hyperlink{ref-Moran2012}{2012}), p.~212 does provide a
  log-log plot of segment frequencies that is Zipfian-esque.} but he
does show that the cumulative number of segment types seems to be
bounded to infinity; the curve appears quadratic and showing no sign of
a slowing towards an asymptote.

In Figure 1, we reproduce these plots using the genealogically balanced
UPSID-451 language sample
(\protect\hyperlink{ref-MaddiesonPrecoda1990}{Maddieson \& Precoda
1990}) and compare it to the data available in the latest version of
PHOIBLE Online (\protect\hyperlink{ref-MoranMcCloy2019}{Moran \& McCloy
2019}), which contains 2186 languages.\footnote{Note that although
  nearly half of the segments in the language samples occur in
  particular languages, some languages contain more than one segment not
  found in any other languages in the sample. This is most apparent from
  the rich phonological inventory of Juǀʼhoan (ISO 639-3: ktz;
  Glottocode: juho1239; aka !Xu, ǃXun, Southeastern ǃKung), which has
  been described as having 141 phonemes -- 66 of which occur in no other
  documented language (\protect\hyperlink{ref-Snyman1970}{Snyman 1970};
  \protect\hyperlink{ref-Snyman1975}{Snyman 1975};
  \protect\hyperlink{ref-MaddiesonPrecoda1990}{Maddieson \& Precoda
  1990}).}

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/cumulative_plot_phoible_upsid-1}

}

\caption{Cumulative segments in PHOIBLE 2.0 and UPSID-451}\label{fig:phoible_upsid}
\end{figure}

These plots resemble the distribution of another, less-commonly known
linguistic law: Herdan's and/or Heaps' law. Originally proposed by
Gustav Herdan (\protect\hyperlink{ref-Herdan1964}{1964}), but frequently
attributed to Harold Heaps (\protect\hyperlink{ref-Heaps1978}{1978}) in
the information retrieval literature
(\protect\hyperlink{ref-Egghe2007}{Egghe 2007}), Herdan-Heaps' law
describes the number of distinct words within a document as a function
of the document's (or combined documents') length in terms of total word
count. More specifically, Herdan--Heaps' law states that, as more of a
text is consumed, the rate of never-before-seen words per unit
diminishes (but does not reach 0). This makes intuitive sense given what
we know from Zipf's law: most words are hard to find, and most of the
words that you can find, you have seen before. In fact, Herdan--Heaps'
and Zipf's laws have been shown to be asymptotically equivalent with
respect to individual word frequencies within corpora
(\protect\hyperlink{ref-BaezaNavarro2000}{Baeza--Yates \& Navarro 2000};
\protect\hyperlink{ref-vanLeijenhorst2005}{Leijenhorst \& Van der Weide
2005}). Both of these laws are instances of the same fundamental
property: token counts can be inferred from type counts and vice versa
(\protect\hyperlink{ref-Milivcka2009}{Milička 2009}).

Herdan-Heaps' is expressed as:

\[
V_{R}(n)=Kn^{\beta }
\]

where \(V_{R}\) is the number of distinct words in a text of size
\emph{n} and \emph{K} and \(\beta\) are the free parameters that are
determined empirically. An illustration of a typical Herdan-Heaps' curve
for English text is given in Figure 2 (compare to Figure 1 above).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{figures/Heaps_law_plot}

}

\caption{A typical Herdan-Heaps plot, where x-axis denotes text size and the y-axis the number of distinct vocaulary items (Source: Wikipedia).}\label{fig:heaps}
\end{figure}

Herdan--Heaps' law has been validated using large corpora of naturally
occurring discourse. For example, in a corpus of 50 million words,
Kornai (\protect\hyperlink{ref-Kornai2002}{2002}) found that iteratively
adding more text always led to the discovery of more unique words.
Brants \& Franz (\protect\hyperlink{ref-BrantsFranz2006}{2006}), using a
custom one-trillion-word corpus of English, showed that even with nearly
14 million word types, the growth in vocabulary showed no sign of
stopping.\footnote{Note that ``words'' at the trillion-word scale corpus
  include all segmentable substrings, including alternate spellings for
  the same target word. These alternatives can be numerous. For example,
  Google's spell correction system has detected nearly 600 different
  spellings for the query \emph{britney spears}:
  \url{http://archive.google.com/jobs/britney.html}.}

We can apply the same approach to phonemes nested within inventories. By
analogy to words and documents, phonemes are the words, and the
concatenated inventories of a sample of languages (\emph{L}) is a
document. The ``document'' is expanded by considering new, entire
phoneme inventories one language at a time. Any new phonemes are added
to the existing set and the new count recorded (as in Figure 2 above).
If the phoneme type counts predicted by Herdan-Heaps' law match those of
the known sample of inventories, we can then extrapolate the trend to
larger number of inventories. For example, how many phonemes potentially
exist in the full (but only partially documented) set of the world's
7151 spoken languages.\footnote{This estimate is taken from Ethnologue (
  \url{https://www.ethnologue.com/guides/how-many-languages}?).} What
about even more languages? How many phonemes may have existed throughout
time? And what is the upper bound on the number of distinct phonemes? Is
there one? Or does Herdan--Heaps' law suggest that the number of
phonemes in languages is potentially infinite?

\hypertarget{data}{%
\section{Data}\label{data}}

For this study, we use the latest PHOIBLE data, which includes 3020
phonological inventories from 2186 distinct languages, covering 3183
distinct phonemes (\protect\hyperlink{ref-MoranMcCloy2019}{Moran \&
McCloy 2019}). The dataset is a convenience sample of languages, in
which some geographic regions and language families have greater or
lesser coverage. Included in PHOIBLE is UPSID-451
(\protect\hyperlink{ref-Maddieson1984}{Maddieson 1984};
\protect\hyperlink{ref-MaddiesonPrecoda1990}{Maddieson \& Precoda
1990}), a genealogically balanced sample of languages in which each data
point was selected as a representative of its language family.
Therefore, we use both UPSID and PHOIBLE as baselines to evaluate our
models.

Our central question concerns how the number of spoken languages that
are documented relate to the number of unique phonemes that are
observed. This is a question of cumulative growth. It can be rephrased
as: if we have seen \emph{x} languages with \emph{y} unique phonemes,
how many unique phonemes do we expect to find in \(x + i\) languages?

\hypertarget{methods-and-results}{%
\section{Methods and results}\label{methods-and-results}}

\hypertarget{zipfs-law}{%
\subsection{Zipf's law}\label{zipfs-law}}

First, we test whether the segment distribution in the databases follow
\href{https://en.wikipedia.org/wiki/Zipf\%27s_law}{Zipf's law}. Figures
3 \& 4 plot log frequency (y-axis) by log frequency rank (x-axis) for
UPSID and PHOIBLE, respectively. The circles represent the cumulative
distributional counts of each phoneme. The blue dotted line represents
predictions based on Zipf's distribution using the following equation:

\[
\begin{aligned}
  f(rank;a;b;c) & = \log{\frac{c}{(rank+b)^a}} \\ 
  &= \log{(c) + a \cdot\log{(rank + b)}}
\end{aligned}
\]

where \(a\), \(b\), and \(c\) are tunable parameters. In this case, we
determined the optimal parameters using non-linear least squares
regression as implemented in the \emph{nls} function from the
\emph{stats} package (v. 3.6.2) for the programming environment \emph{R}
(\emph{R} Core Team, 2019).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/zipf_distribution_upsid-1}

}

\caption{\label{fig:upsid}Cumulative segments in UPSID-451 follow Zipf's law}\label{fig:upsid_zipf}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/zipf_distribution_phoible-1}

}

\caption{Cumulative segments in PHOIBLE (OIPL) follow Zipf's law}\label{fig:phoible_zipf}
\end{figure}

The close fit between the observed curve and the predictions drawn from
a Zipfian distribution suggests that global phoneme distributions are
Zipf-distributed.\footnote{We note the debates in the literature on what
  the underlying distributions are of phoneme inventory size (e.g.,
  Lehfeldt (\protect\hyperlink{ref-Lehfeldt1975}{1975}), Justeson \&
  Stephens (\protect\hyperlink{ref-JustesonStephens1984}{1984}),
  Maddieson (\protect\hyperlink{ref-Maddieson2008}{2008})) and of
  phonemes frequencies in text
  (\protect\hyperlink{ref-Martindale_etal1996}{Martindale et al. 1996};
  \protect\hyperlink{ref-TambovtsevMartindale2007}{Tambovtsev \&
  Martindale 2007};
  \protect\hyperlink{ref-Macklin-CordesRound2020}{Macklin-Cordes \&
  Round 2020}). But here we are specifically looking at presence of
  phonemes in phonological inventories cross-linguistically.} Note that
this close approximation holds for both the genealogically balanced and
large-scale databases.

\hypertarget{herdanheaps-law}{%
\subsection{Herdan--Heaps' law}\label{herdanheaps-law}}

The distribution of phonemes in both language samples are clearly
Zipfian. We next test whether their distributions likewise follow
Herdan--Heaps' law. Does the number of yet-to-be-seen segment types
decrease as the number of phonological inventories observed increases by
way of a power law relation?

An important feature of the Herdan-Heaps' equation is that it can be
expressed in the form of a standard linear equation by simply taking the
logarithm of each term, or:

\[
\begin{aligned}
  \log d & = \log (K + n^B)\\
  & = \log K + B \cdot \log n
\end{aligned}  
\]

which has the familiar form:

\[y = \beta_0 + \beta_1 \cdot x\]

This gives us the added benefit of being able to fit a log-transformed
linear regression model to the data, such that:

\begin{itemize}
\tightlist
\item
  The resulting intercept is the optimal \(K\) (with some
  back-transformation: \(K = e^k\)).
\item
  The coefficient \(\beta_1\) expresses \(B\).
\end{itemize}

With these values, we can estimate the type count for any sample size
\(n\) using the standard Herdan-Heaps' equation. As proof of concept, we
estimate the number of phonemes for both UPSID-451 and PHOIBLE. Some
languages in PHOIBLE are associated with more than one doculect. To
avoid an overrepresentation bias, we randomly sample one doculect from
every language or take the single doculect where only one exists (one
inventory per language, or OIPL. Results of a preliminary analysis
indicated that the full and OIPL versions of PHOIBLE behaved similarly.
We therefore focus on the results from the OIPL database here (see the
supplementary materials for comparisons between the full and OIPL
versions).

We begin by randomly permuting the order of inventories and computing
the cumulative number of unique segment types for each new language
observed. Then we repeat this process \(k\) times (we set k=10 trials)
for each database. We fit a log-linear regression to the resulting sets
of estimates to generate empirically optimal values for \(B\) and \(K\)
-- our two free parameters in the Herdan-Heaps' equation.

Figures 5, 6 \& 7 show the predicted values against the true values of
the number of segment types (phoneme type count) versus segments
recorded in phonological inventories in the samples (segment token
count). Model fits for the UPSID-451 and PHOIBLE language samples
suggest that the phoneme type versus token count closely follows
Herdan-Heaps' law.\footnote{In the supplementary materials, we also fit
  a ``one inventory per language'' (OIPL) model of PHOIBLE to show that
  the language sample's genealogical and areal biases, e.g., the sample
  is more representative of African languages than other world areas,
  are not affected by simply sampling the whole database. The outcomes
  between a stratified random resampling and testing the segment type
  versus token on the entire convenience sample are nearly the same.}

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/zipf_distribution_upsid-1}

}

\caption{\label{fig:upsid}Cumulative segments in UPSID-451 follow Zipf's law}\label{fig:upsid_heaps}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/zipf_distribution_phoible-1}

}

\caption{\label{fig:upsid}: Cumulative segments in PHOIBLE (OIPL) follow Zipf's law}\label{fig:phoible_heaps}
\end{figure}

Next, we compare the model parameters across the two databases,
UPSID-451 and PHOIBLE. The values for \(B\) and \(K\) are relatively
similar. UPSID-451 and PHOIBLE (all) have nearly identical values for B,
but they differ in the scalar K, which explains the sharper increases in
segment type counts for PHOIBLE.

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
Database & B & K \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
UPSID & 0.4994352 & 8.087208 \\
PHOIBLE (OIPL) & 0.5442040 & 5.870403 \\
\end{longtable}

We illustrate these parameter differences in Figure 7 by combining the
plots on the same scale. The curve describing Herdan-Heaps' estimates
based on UPSID-451 has a shallower trajectory than the curve based on
PHOIBLE. Nevertheless, both curves are highly similar. This similarity
indicates that the known areal and genetic biases present in PHOIBLE do
not have much of an effect on the modeling of phoneme diversity. Based
on this finding, we focus on the larger and more complete database,
PHOIBLE, in the remaining analyses.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/compare_models-1}

}

\caption{\label{fig:upsid}: Cumulative segments in UPSID-451 and PHOIBLE (OIPL) follow Herdan-Heaps' law}\label{fig:compare_models}
\end{figure}

Next, we implement a general and flexible estimator for type counts
called the Chao-Shen estimator
(\protect\hyperlink{ref-Chao_etal2009}{Chao et al. 2009};
\protect\hyperlink{ref-ChaoChiu2016}{Chao \& Chiu 2016}). If the results
from the two estimators (Chao-Shen and Herdan-Heaps' law) converge, then
we have a more solid foundation for extrapolating phoneme counts for
larger, hypothetical sample sizes.

\hypertarget{the-chao-shen-estimator}{%
\subsection{The Chao-Shen estimator}\label{the-chao-shen-estimator}}

The Chao-Shen estimator was developed to determine the diversity of
species within an ecosystem given only limited (spatial and temporal)
sampling (\protect\hyperlink{ref-Chao_etal2009}{Chao et al. 2009};
\protect\hyperlink{ref-ChaoChiu2016}{Chao \& Chiu 2016}). It allows us
to extrapolate the number of types in some population based on an
incomplete sample using the number of rare types and an estimate of the
number of unobserved types.

This estimator takes as input an incidence matrix and returns the
expected number of types in a larger (imaginary) sample. An incidence
matrix is a record of the presence or absence of the types being sampled
from a set of sampling units. For example, imagine that an entomologist
is interested in identifying how many types of beetles are in a forest.
She knows from prior research that there are 1000 documented species of
beetles in the area. In order to get a representative sample of the
diversity of beetle species in the overall ecosystem, she sets up 50
beetle traps, regularly spaced across the forest. In this example, the
traps are the sampling units and the species of beetles are the types.
The entomologist prepares a 50 X 1000-cell spreadsheet in which the rows
and columns correspond to the traps and known beetle species,
respectively. She then visits each trap and for each of the 1000 known
species of beetle she finds, she records a 1 for that species in
appropriate cell. All unseen species are given a value of 0. The result
after all traps have been visited is a 50 X 1000 incidence matrix. The
information contained in this matrix can then be fed into the Chao-Shen
estimator to provide an estimate of the true diversity of beetle species
in the forest.

The same approach can be extended to languages and their phonemes. On
analogy with the beetle example, the forest is the sum of the
human-populated areas of Earth; the beetle traps are all documented
human languages; and the beetles are the phonemes which appear in those
languages. The incidence matrix is of shape \emph{l} X \emph{s}, where
\emph{l} is the number of documented languages and \emph{s} is the
number of types of segments that have been documented as phonemes in at
least one language. By definition, if a segment appears as a phoneme in
a given language, the corresponding cell in the incidence matrix
contains a 1; otherwise, it contains a 0.

Table 1 shows a hypothetical example of an incidence matrix for 150
observed segments from a set of \(S_{obs}\) phonemes in \(R\) languages.
Marginal sums are given per language (\(\sum_{language}\), or
\emph{inventory size}) and per segment (\(\sum_{phoneme}\), or
\emph{cross-linguistic prevalence})

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2273}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1364}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1364}}
  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2273}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1364}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1364}}@{}}
\caption{Hypothetical sample of an incidence matrix}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
Phoneme
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Lang1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Lang2
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
···
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Lang\emph{R}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
∑phoneme
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
Phoneme
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Lang1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Lang2
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
···
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Lang\emph{R}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
∑phoneme
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{Phon1} & 1 & 0 & ··· & 0 & 1 \\
\textbf{Phon2} & 0 & 1 & ··· & 1 & 2 \\
··· & ··· & ··· & ··· & ··· & ··· \\
\textbf{Phon\emph{S}obs} & 1 & 1 & ··· & 1 & 30 \\
\textbf{∑language} & 15 & 23 & ··· & 6 & 150 \\
\end{longtable}

Using the definitions provided above, the Chao-Shen estimator is
expressed formally as follows:

\[ \hat{S}_{\text{Chao-Shen}} = S_{\text{obs}} + \hat{Q}_0 \left[1 - \left(1 - \frac{Q_1}{Q_1 + R \cdot \hat{Q}_0}\right)^r\right] \]

\(S_{obs}\) is the number of segment types that have been observed
overall. \(R\) is the number of languages in the observed sample.
\(Q_1\) is the number of segment types that have only been observed once
(\emph{singletons}), e.g., Phon1 in Table 1). The hyperparameter \(r\)
expresses the number of new languages that we wish to add
(hypothetically) to our existing sample. Finally, \(\hat{Q}_0\) is the
estimated number of unobserved types that were nevertheless ``truly''
present. We do not have empirical access to this figure -- we cannot
directly measure how much we haven't seen. Therefore, it must be
inferred. \(\hat{Q}_0\) can be derived in many ways. Here, we use the
following equation.

\[ \hat{Q}_0 = \begin{cases} S_{\text{obs}} + \left[\frac{R-1}{R}\right] \frac{Q^2_1}{2 \cdot Q_2} & \text{if }Q_2 > 0 \\ S_{\text{obs}} + \left[\frac{R-1}{R} \right] \cdot Q_1 \cdot \frac{Q_1-1}{2} & \text{if }Q_2 = 0\end{cases}\]

In this equation, \(Q_2\) stands for number of segment types which were
observed exactly twice, or \emph{duplicates} (e.g., Phon2 in Table 1).

Similar to the Herdan-Heaps' equation, we can test how well the
estimates of the model match known values based on increasing sample
sizes. However, unlike Herdan-Heaps', the Chao-Shen estimator requires a
hyperparameter that stipulates how many \emph{additional} observations
it should consider in making its estimate. It looks ahead to larger
samples, rather than at the current sample size. We must therefore
approach this problem differently. Our solution is to sample repeatedly
from our set of languages, each time increasing the number of languages
\(R\) by an increment of \(d\). For each value of \(R\), we compute the
Chao-Shen estimates for several values of \(r\), each time increasing
\(r\) by \(c\) additional observations.

The entire procedure is thus:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Pick a random set of phonological inventories of size \(R_s\)
\item
  Compute the frequency vector of the segment types
\item
  Apply the Chao-Shen estimator for \(r\) with \(R = R_s\) 3a. Increase
  \(r\) by \(c\) 3b. Repeat 3a until maximum \(r_{\text{max}}\) is
  reached
\item
  Repeat steps 1-3 \(k\) times 5a. Increase \(R_s\) by \(d\)
\end{enumerate}

Figure 8 shows the results of applying this procedure to PHOIBLE. \(k\)
is arbitrarily set to 10 iterations. The relatively small degree of
variation across iterations suggests that this value is suitably high.
\emph{d} was set to 10. \emph{r} was manually adjusted to exemplify
behavior at different scales (within the limits afforded by the overall
sample size). Grey lines represent the Chao-Shen estimates for each
iteration, and blue lines give the means. Red lines represent the mean
of the true number of observed types across iterations. Red lines become
shorter as \emph{r} increases because we have fewer true values to
compare against when we look further ahead from any position (i.e.,
fewer values of \emph{R} for which we can look \emph{r} steps ahead).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/chao_lang_types-1}

}

\caption{\label{fig:chao_lang_types}Estimated segment types by number of languages}\label{fig:chao_lang_types}
\end{figure}

The first thing to notice about these plots is that they closely
resemble the Herdan-Heaps' curve that we generated from PHOIBLE. These
plots thus also point to slowing but non-asymptotic growth of type
counts.

Regarding the behavior of the parameters, the Chao-Shen estimator shows
near-perfect prediction across values of \emph{R} when \emph{r} is
relatively small. That is, the model is very good at predicting known
segment type counts for samples with relatively fewer additional
languages (small \emph{r}), regardless of the number of observed
languages (all \emph{R}). If we compare the panels against one another,
particularly at lower values of \emph{R}, we see that the differences
between the true values (red lines) and the estimated values (blue
lines) increase with \emph{r}. As \emph{r} increases, so does the degree
of underestimation. In other words, the model is more conservative in
its guesses when data are sparse and the distances of extrapolation are
large. The underestimation appears to dissipate around \emph{R} = 1000.
Figure 9 plots this relationship directly.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/chao_error_estimates-1}

}

\caption{\label{fig:chao_error_estimates}Chao error estimates in PHOIBLE}\label{fig:chao_error_estimates}
\end{figure}

Variability across iterations increases with \emph{r}, particularly for
sample sizes of \textasciitilde1700 languages or fewer. This pattern
indicates that larger values of \emph{r} should be compensated for by
larger values of \emph{R}. Otherwise, the results may not generalize
well to other samples (even if the other samples are drawn from the same
underlying pool, as we have shown here!).

To summarize, estimates converge on the observed values earlier on
average when \emph{R} is larger relative to \emph{r}. Thus, we do better
when we have more data and predict values closer to our observed sample.
In addition, the variability in estimates across iterations (grey lines)
increases with \emph{r} and decreases with \emph{R}. Thus, our estimates
are noisier when we look further ahead and have smaller samples.
Together these findings suggest two things: (a) the ratio of observed
sample size to sample size increase \(\frac{R}{r}\) should be maximized
to reduce sample-induced noise, and (b) \emph{R} should be maximized to
fight the underestimation bias.

Unfortunately, PHOIBLE offers a proportionally small \emph{R} given some
of the values of \emph{r} that we would like to investigate. For
example, to account for the estimated number of current languages, we
would need \emph{r} = 7151 - 2099 = 5052, a value almost 2.5 times as
large as our observed \emph{R} (2099). Extrapolating from our known
sample to the estimated number of world languages therefore requires an
undesirably small proportion \(\frac{R}{r}\) -- around 0.42. That said,
our data suggest that sample sizes of \textasciitilde1000 languages
produce estimates that closely approximate the true segment type counts,
even as \emph{r} increases. In other words, after 1000 languages,
increases in \emph{R} do less to mitigate the underestimation bias.
Given that PHOIBLE contains more than double that number of languages,
we have no reason to suspect significant underestimation for the
relatively large values of \emph{r} that we investigate here. However,
because of the small ratio of \emph{R}:\emph{r}, we should expect our
estimate to come with greater uncertainty.

We have so far described the distributional properties of our databases,
associated their behavior with two linguistic laws, and introduced two
methods for inferring type counts from an observed frequency
distribution. We now explore in more detail how well the two type
estimators -- Herdan-Heaps' and Chao-Shen -- perform when predicting the
total segment type diversity of a database from an incomplete subsample.

\hypertarget{predicting-known-segment-type-diversity}{%
\subsection{Predicting known segment type
diversity}\label{predicting-known-segment-type-diversity}}

To establish whether we can reliably estimate the number of segment
types in a known sample, we evaluate the accuracy of each estimator in
predicting the known type count given a fixed-size random subsample.
From PHOIBLE (\emph{n} = 2099), we randomly select 1500 languages -- an
arbitrary number that includes roughly three-quarters the number of
languages in the total sample. Note that this value was set above the
threshold of \emph{R} = 1000 (reported above for the Chao-Shen
estimator) to help support prediction accuracy. We then extrapolate our
model parameters from this random sample and use them to predict total
segment type diversity in PHOIBLE. We iterate this process \emph{k} =
100 times.

The Herdan-Heaps' estimates are visualized in Figure 10. The mean is
plotted as a solid black line. The 97.5\% confidence intervals (CIs) are
plotted as dotted lines around the mean. The red dotted line indicates
the true number of phoneme types. Blue shading indicates the density
function of the estimated values from the 100 iterations.

For this run of 100 iterations, the Herdan-Heaps' estimator
overestimated the true value by \textasciitilde18 segment types on
average (\textasciitilde2613 estimated vs.~2596 observed).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/herdan_heaps_ci-1}

}

\caption{\label{fig:herdan_heaps_ci}Herdan-Heaps' confidence intervals for a sample of 1500 languages}\label{fig:hh_ci}
\end{figure}

Next we test the Chao-Shen estimator. We again sample 1500 languages 100
times from PHOIBLE. The results are plotted in Figure 11.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/chao_ci-1}

}

\caption{\label{fig:chao_ci}Chao confidence intervals on a sample for 1500 languages}\label{fig:chao_ci}
\end{figure}

The Chao-Shen estimator overestimates the true value by only 1 type (the
estimate is actually closer: 0.56 types). It therefore provides an
almost perfect fit. Note that \(\frac{R}{r} = \frac{1000}{1099} = 0.91\)
(three times the size of what we face with our extrapolation to the
world's 7151 known languages).

The Chao-Shen estimator therefore seems to outperform Herdan-Heaps' law
in predicting type counts in PHOIBLE. However, we stress that
predictions from both models fall remarkably close to the true value.

As a final point of comparison, we plot Herdan-Heaps estimates against
the corresponding Chao-Shen estimates for increasing sample size and
values of \emph{r} (Figure 12).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/comp_plot-1}

}

\caption{\label{fig:comp_plot}Comparing the Herdan-Heaps' and Chao esimtator}\label{fig:comp_plot}
\end{figure}

The black line represents PHOIBLE-based Herdan-Heaps estimates. Grey
points reflect true cumulative type counts from 10 reshuffled versions
of PHOIBLE. The horizontal dotted red line indicates the total known
number of segment types in PHOIBLE. The solid colored lines represent
Chao-Shen estimates for different \emph{r} (indicated in the legend).
Note that these lines have been shifted along the x-axis to reflect that
Chao-Shen estimates look ahead, such that the \(\hat{R}\) being
estimated is actually \emph{R} + \emph{r}.

This plot demonstrates several of the points raised above. The Chao-Shen
estimates at small \emph{R} and large \emph{r} are gross underestimates
of the true value. However, as \emph{R} increases, the estimated values
climb quickly towards the true values. What is striking about this plot
is that the Chao-Shen estimates seem to ``seek out'' the Herdan-Heaps
curve. For these data, the Chao-Shen estimator has decided that segment
type counts grow according to a rate predicted by Herdan-Heaps' law.
This need not have been the case; the Chao-Shen estimator is capable of
predicting values produced by many different kinds of functions.
Finally, and most crucially, neither estimator shows any indication of
leveling out: both appear to be non-asymptotic.

\hypertarget{extrapolating-estimates-to-unseen-numbers-of-observations}{%
\subsection{Extrapolating estimates to unseen numbers of
observations}\label{extrapolating-estimates-to-unseen-numbers-of-observations}}

We have so far established that (a) both the Herdan-Heaps' estimator and
the Chao-Shen estimator predict unseen type counts accurately and (b)
the two ultimately agree on these estimates, though Chao-Shen needs more
data to ``catch up.'' We have also shown that the Chao-Shen estimator
requires two things to perform well: large ratios of \emph{R} to
\emph{r} and generally large \emph{R}. With this in mind, we now apply
both estimators to unseen sample sizes. We start with the number of
known world languages (7151), then proceed to larger values (10k, 50k,
and 100k).

\hypertarget{herdan-heaps-law}{%
\subsubsection{Herdan-Heaps' law}\label{herdan-heaps-law}}

Recall that the Herdan-Heaps' equation requires that we know three
things: two free parameters, \emph{K} and \emph{B}, and the number of
observations \emph{n}. As before, the parameters are optimized using a
log-transformed linear regression over cumulative type counts from
PHOIBLE. However, determining the third variable -- number of
observations -- for an arbitrarily large sample poses a problem. Suppose
you have a sample of 100 languages with 1000 total segments (tokens, not
types), and you want to guess how many total segments you would find in
a sample of 150 languages. Is it simply a linear increase, i.e., 1500?
No, or rather we cannot know. We do not know how many sounds each
language would contribute in an unseen population, so we cannot directly
derive the requisite number of observations from the number of languages
alone. Therefore, we need some way to estimate this figure.

We approach this problem in two ways. First, we sample inventories
randomly with replacement from the observed dataset up to a desired
number of languages. The steps are as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Randomly sample \emph{L} languages from the existing database with
  replacement
\item
  Sum the inventory sizes of the resulting sample of languages.
\item
  Repeat steps 1-2 \emph{k} times
\end{enumerate}

\emph{L} corresponds to the desired number of languages. For example,
setting \emph{L} to 7151 will produce the number of segments (tokens) we
should expect to see if we had access to the inventories of all known
human languages.

Second, we regress the number of total phonemes on sample size using the
observed dataset. The steps are as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Randomly sample \emph{l} languages from the existing database with
  replacement.
\item
  Sum the inventory sizes of all languages in the resulting sample.
\item
  Increase \emph{l} by increment \emph{s}
\item
  Repeat steps 1-3 until desired number of languages (\emph{L}) has been
  sampled.
\item
  Repeat steps 1-4 \emph{k} times
\item
  Fit a regression to the sequences created
\item
  Use the regression to predict token counts for desired \emph{L}
\end{enumerate}

The result of steps 1-5 is \emph{k} sequences of increasing token counts
(by steps of \emph{s}) for independent, random samples \emph{l}
languages up to the desired number of languages \emph{L}. Steps 6-7 fit
a linear regression to the set of \textit{k} sequences (with random
intercept adjustments for the iterations). The resulting model is then
used to predict the number of observations for any number of languages.

We apply both approaches to PHOIBLE for samples sizes of 7151, 10k, 50k,
and 100k. The mean estimates for Approach 1 and 2 are presented in Table
2.

\begin{longtable}[]{@{}rcccc@{}}
\caption{Mean estimates for Approach 1 \& 2.}\tabularnewline
\toprule\noalign{}
& 7151 & 10K & 50K & 100K \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& 7151 & 10K & 50K & 100K \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Approach 1 & 247623 & 346234 & 1731657 & 3465640 \\
Approach 2 & 247620 & 346257 & 1731115 & 3462189 \\
Difference & 3 & -23 & 542 & 3451 \\
\end{longtable}

As sample size increases, Approach 1 produces increasingly larger
estimates relative to Approach 2. At lower sample sizes, the values are
comparable.

The values from Approach 1 and 2 can now be entered into the standard
Herdan-Heaps' equation with \emph{K} and \emph{B} taken from the
full-PHOIBLE regression model. Results are given in Table 3.

\begin{longtable}[]{@{}rcccc@{}}
\caption{Phoneme type counts for different sample sizes.}\tabularnewline
\toprule\noalign{}
& 7151 & 10K & 50K & 100K \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& 7151 & 10K & 50K & 100K \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Approach 1 & 4850 & 5770 & 13282 & 19025 \\
Approach 2 & 4850 & 5770 & 13280 & 19016 \\
Difference & 0.02 & -0.20 & 2.15 & 9.81 \\
\end{longtable}

Both approaches behave similarly, even producing the exact same estimate
for 7151 and 10000 tokens.\footnote{The Supplementary Materials contain
  an analysis which suggests that most languages tend to settle on a
  similar number of phonemes, meaning that the overall number of
  observations can be neatly derived from the number of languages.}

\hypertarget{chao-shen-estimator}{%
\subsubsection{Chao-Shen estimator}\label{chao-shen-estimator}}

Recall that the Chao-Shen estimator only requires a single parameter
\emph{r}, i.e., the number of additional languages that we wish to
sample. To derive \emph{r} for a target sample size, e.g., 7151, we need
only subtract our observed \emph{R}, or 2099, from that number. The
resulting figure is then plugged into the estimator to get the expected
type count \(\hat{R}\). The full set of \emph{r} values and
corresponding type count estimates are given in Table 4.

\begin{longtable}[]{@{}rcccc@{}}
\caption{\emph{r} values and corresponding type count
estimates}\tabularnewline
\toprule\noalign{}
& 7151 & 10K & 50K & 100K \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& 7151 & 10K & 50K & 100K \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(r\) & 5052 & 7901 & 47901 & 97901 \\
\(\hat{R}\) & 4592 & 5037 & 5530.86 & 5530.92 \\
\end{longtable}

Predicted type counts are generally lower than those produced by
Herdan-Heaps' law, and this difference increases with \emph{r}. Unlike
the Herdan-Heaps' estimator, the Chao-Shen estimator does appear to
reach an asymptote. Indeed, increasing \emph{r} beyond 100K always
results in the same value: 5530.92. The presence of this asymptote can
be explained by the function of \emph{r} in the Chao-Shen equation.
\emph{r} serves to exponentiate a term that falls in the range \{0,1\}.
Therefore, as \emph{r} increases, the full term shrinks to 0. The result
is subtracted from 1, meaning that as \emph{r} increases, the compound
term \(1-x^r\) approaches 1. This value controls how much of the
estimated number of unobserved types \(\hat{Q_0}\) is added to the
number of observed types \(S_{\text{obs}}\). Therefore, as the compound
term approaches 1 per unit increase in \emph{r}, the resulting type
count estimates get closer and closer together. Looking too far ahead
slows the growth of type count estimates to a crawl. In turn, this
slowdown produces increasingly more severe underestimates (as outlined
above). The apparent asymptote can therefore be interpreted as a
mechanical feature of the Chao-Shen estimator rather than a substantial
property of global phonological distributions.

Figure 13 summarizes the segment type estimates of the two Herdan-Heaps'
methods and the Chao-Shen estimator.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/diff_plots_ext-1}

}

\caption{\label{fig:diff_plot_ext}Type count estimates from the Chap and Herdan-Heaps' estimators for four sample sizes}\label{fig:diff_plot_ext}
\end{figure}

The two approaches to estimating \emph{n} for Herdan-Heaps are nearly
identical. Furthermore, we see diminishing rates of increase of the
estimates (going from 10K to 50K results in a greater increase than
going from 50K to 100K). For smaller \emph{r}, the Chao-Shen and
Herdan-Heaps' estimators behave similarly. As \emph{r} increases, the
degree of difference between the two estimators increases (as predicted
based on the underestimation bias reported above for Chao-Shen).

To summarize our main findings: (a) Herdan-Heaps' law fits the known
phonological diversity well, (b) a different strategy for estimating
type count performs best when it mimics Herdan-Heaps' law, (c) we
therefore have no reason to believe that there is a limit to the number
of phonemes to be observed in human languages, and (d) even if limited,
the true number is most likely much larger than what we have documented
so far.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The present study asks a seemingly straightforward question: how many
different phonemes are possible in human language? We cannot determine
this number empirically from existing resources because the vast
majority of languages, known or unknown, are undocumented, extinct, or
both. Moreover, differences in documentary practice (theoretical and/or
methodological) mean that even our existing phonological descriptions
are provisional and objects of debate. For example, PHOIBLE contains
approximately 1.4 different inventories per language.\footnote{Different
  inventories per ``language'' mean they do not agree on the number or
  exact repertoire of phonemes. However, this should be expected given
  that phonological descriptions typically represent different
  doculects, i.e., different instances of linguistic behavior at
  different places, times, or with different speakers
  (\protect\hyperlink{ref-GoodCysouw2013}{Good \& Cysouw 2013}) and that
  phoneme analysis is a non-deterministic process
  (\protect\hyperlink{ref-Chao1934}{Chao 1934};
  \protect\hyperlink{ref-Hockett1963}{Hockett 1963}).}

To address these issues, we modeled the distributional properties of
documented phonemes. First, we found that frequencies of phoneme types
in existing cross-linguistic databases are well described by Zipf's law,
regardless of whether samples were controlled for areal and genetic
biases. Systems that obey Zipf's law, under mild assumptions, also
follow Herdan-Heaps' law
(\protect\hyperlink{ref-BaezaNavarro2000}{Baeza--Yates \& Navarro 2000};
\protect\hyperlink{ref-vanLeijenhorst2005}{Leijenhorst \& Van der Weide
2005}), which predicts type count from the number of tokens with only
two tunable parameters. Herdan-Heaps' law was originally intended to
describe how vocabulary size increases as more words are observed in a
document or set of documents. Extending this approach, we demonstrate
that the number of phoneme types in a sample of languages also grows
according to Herdan-Heaps' law. To our knowledge, we are the first to
apply Herdan-Heaps' law to the domain of segmental phonology by
reformulating it as estimator of the number of phonemes in groups of
languages. Finally, we verify the goodness-of-fit of Herdan-Heaps' by
applying an additional and formally unrelated tool borrowed from
Ecology, which we refer to as the Chao-Shen estimator. The predictions
of this estimator confirm that the relationship between sample size and
segment type count is best described by a curve that approximates
Herdan-Heaps' law. Most importantly, the optimal Herdan-Heaps' model did
not converge on any single number of phoneme types as number of
languages increased.

Assuming a target of 7151 known languages,\footnote{\url{https://www.ethnologue.com/guides/how-many-languages}?}
our models predict a range of total segment types from 4592 to 4850
(mean = 4721). Based on PHOIBLE, our current best guess of the diversity
of documented phoneme types is 3183. Therefore, the models argue for the
existence of approximately 1500 living but undocumented phoneme types.
As we probe increasingly larger and theoretically motivated language
sample sizes, i.e., languages that existed \emph{through time} (e.g.,
estimates including 30k-500k (\protect\hyperlink{ref-Pagel2000}{Pagel
2000}) or 64k-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})),
the Herdan-Heaps' model predicted a continual increase in the number of
unique phonemes, reaching values as high as 19K. The Chao-Shen model
produced much more conservative estimates for samples of size 10K and
above, even appearing to settle on an asymptote of roughly 5500 phonemes
for samples of 100K languages or more. However, when we compare the
predictions of Herdan-Heaps and Chao-Shen against known type counts in a
closed dataset, the source of the discrepancy emerged: as the sample
grows, Chao-Shen lags behind Herdan-Heaps in approximating the true
value. Thus, given a larger pool of training data, we expect Chao-Shen
to approximate more closely the estimates of Herdan-Heaps.

We therefore provisionally conclude three things: (1) adding languages
to a sample will always carry a non-zero probability of producing a
novel phoneme; (2) in accordance with Herdan-Heaps' law, the rate at
which new phonemes are discovered per language will always decrease; and
(3) the decrease in the rate of new observations will never reach 0.
Thus, based on the evidence presented here, we have no reason to believe
that there is an upper bound on the number of unique phonemes possible
in spoken languages. Infinite linguistic diversity seems to breed
infinite possible phonemes. But what would it mean if there were indeed
infinitely many potential phonemes?

If one adopts a feature-based perspective on the phonetic targets of
phonemes, then there is an \emph{a priori} upper bound of \(2^F\)
contrastive sounds, where \(F\) is the number of feature categories and
2 reflects the choice between presence (+) or absence (-) of that
feature.\footnote{In some feature sets, some values may be 0, i.e.,
  undefined, underspecified, or ``for which a segment can be said `not
  to care''' (\protect\hyperlink{ref-Hayes2009}{Hayes 2009}, pg. 91).
  For example, speech sounds that are -coronal are underspecified for
  the features +/-anterior and +/-distributed, which capture coronal
  distinctions for places of articulation involving (inter-)dentals and
  alveolars vs palatal alveolars and retroflexes (anterior), and laminal
  vs apical coronals (distributed).} Clements
(\protect\hyperlink{ref-Clements2009}{2009}) coined this calculation
\emph{Feature Bounding}. The expected number of unique phonemes is
controlled by the number of features, where additional features increase
the pool of possible distinctive segments (nearly) exponentially. For
example, given a feature set such as in PHOIBLE with 37 distinctive
features, an upper bound on the number of segments under a strictly
binary formalization is over 137 billion (2\^{}37) feature vectors.

However, most widely accepted feature sets contain features that are
specified at different levels of phonetic detail: some features are more
abstract (i.e., general similarities such as +/- syllabic), while others
are more concrete (i.e., closer to the articulatory reality, e.g.,
advanced tongue root). Some are grounded in pure articulatory or
acoustic phonetics, while others involve phonological properties of
language. And looming in the background of these discussions, the debate
of whether distinctive features are innate and universal or rather
learned and emergent continues. Nevertheless, in general, distinctive
feature theories of different theoretical stances propose around two
dozen features (\protect\hyperlink{ref-MielkeHume2006}{Mielke \& Hume
2006}), which still has the potential to generate over 16 million
distinct feature vectors (\(2^24\)). Current consensus aside, there does
not appear to be a principled limit to the degree of phonetic
specificity that can be encoded by a feature. Making finer and finer
contrasts may always produce more and more features. Thus, the exponent
in the Feature Bounding calculation can itself (by hypothesis) be
infinite, in line with the predictions of the models presented here, and
therefore with the notion of infinite possible phonemes. That said,
contemporary feature sets can capture phonological classes for place and
manner of articulation, which neatly classify the phonological patterns
and processes in and across languages (cf. Mielke
(\protect\hyperlink{ref-Mielke2008}{2008})).

We have so far seen that the feature space is a discrete re-coding of
the continuous variability of the physical acoustic and articulatory
spaces -- and so it should by way of finer and finer specification, be
able to support an infinitely diverse set of phonemes. However, one
objection to this account is that humans are subject to physiological
limits, both in production and perception. In speech production,
articulatory gestures are executed with some degree of imprecision
(\protect\hyperlink{ref-Ladefoged_etal1972}{Ladefoged et al. 1972};
\protect\hyperlink{ref-Lindblom1990}{Lindblom 1990}). If this stochastic
variability is greater than the difference between two or more phonetic
targets, then no reliable distinction can be sustained (cf. Stevens
(\protect\hyperlink{ref-Stevens1972}{1972})). In comparison, if the
range of articulatory gestures is greater than their acoustic (or
perceptual) reception, there is a ``many-to-one'' mapping between
articulation and perception. In perception, the ability to discriminate
acoustic stimuli is also limited both anatomically and cognitively.
Differences in the signal that fall below the various thresholds for
discrimination are thus ignored.

In response to these physical constraints, languages tend to spread
their phonological inventories across the articulatory and acoustic
spaces, thereby maintaining workable discriminability. A stark example
comes from Central Rotokas (North Bougainville; Papua New Guinea), which
has only 6 consonants (9 counting allophones) from three easily
discriminated places of articulation: labials \texttt{{[}p,\ b{]}},
alveolars \texttt{{[}t,\ d{]}}, and velars \texttt{{[}k,\ g{]}}
(\protect\hyperlink{ref-FirchowFirchow1969}{Firchow \& Firchow 1969}).
Compare this to a hypothetical less-optimal language that contrasts only
\texttt{{[}t,\ d{]}}, \texttt{{[}tʰ,\ dʰ{]}}, and \texttt{{[}t̪,\ d̪{]}}.
Given this strategy, it is unlikely that ultra-fine-grained phonetic
contrasts would ever arise in a single language. Rather, the contrasts
are most likely to arise between languages. In other words, the infinity
of phonemes predicted by the analyses reported here plays out in the
realm of cross-linguistic variation of phonetic targets, even if those
differences cannot be reliably articulated by a single tongue or
perceived by a single ear.

Given that all speakers use perceptually contrastive cues (whether
signed, spoken or written), how these continuous signals map to discrete
elements is a core question of language science that goes beyond
whatever the phonetics-phonology interface may be. Thus, we believe the
approach presented here has broad applications to the study of other
areas of grammar and the cross-linguistic study of languages.

\hypertarget{data-accessibility}{%
\section{Data accessibility}\label{data-accessibility}}

Data and code required to reproduce all results presented in this paper
are available as electronic supplementary material at:
\url{https://github.com/bambooforest/heaps}.

\hypertarget{competing-interests}{%
\section{Competing interests}\label{competing-interests}}

The authors declare no competing interests.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

Steven Moran was funded by the Swiss National Science Foundation (Grant
No.~PCEFP1\_186841).

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-BaezaNavarro2000}{}}%
Baeza--Yates, Ricardo \& Gonzalo Navarro. 2000. Block addressing indices
for approximate text retrieval. \emph{Journal of the American Society
for Information Science}. Wiley Online Library 51(1). 69--82.

\leavevmode\vadjust pre{\hypertarget{ref-Boersma2001}{}}%
Boersma, Paul. 2001. Praat, a system for doing phonetics by computer.
\emph{Glot International} 5. 341--345.

\leavevmode\vadjust pre{\hypertarget{ref-BrantsFranz2006}{}}%
Brants, Thorsten \& Alex Franz. 2006. {Web 1T 5-gram Version 1 (2006)}.
\emph{Linguistic Data Consortium, Philadelphia}.

\leavevmode\vadjust pre{\hypertarget{ref-ChaoChiu2016}{}}%
Chao, Anne \& Chun-Huo Chiu. 2016. Species richness: Estimation and
comparison. In \emph{Wiley StatsRef: Statistics reference online},
1--26. John Wiley \& Son.

\leavevmode\vadjust pre{\hypertarget{ref-Chao_etal2009}{}}%
Chao, Anne, Robert K Colwell, Chih-Wei Lin \& Nicholas J Gotelli. 2009.
Sufficient sampling for asymptotic minimum species richness estimators.
\emph{Ecology}. Wiley Online Library 90(4). 1125--1133.

\leavevmode\vadjust pre{\hypertarget{ref-Chao1934}{}}%
Chao, Yuen-ren. 1934. The non-uniqueness of phonemic solutions of
phonetic systems. \emph{Academia Sinica (Bulletin of the Institute of
History and Philology)} 4(4).

\leavevmode\vadjust pre{\hypertarget{ref-Clements2009}{}}%
Clements, G. N. 2009. The role of features in phonological inventories.
In Eric Raimy \& Charles E. Cairns (eds.), \emph{Contemporary views on
architecture and representations in phonology}, 19--68. MIT Press.

\leavevmode\vadjust pre{\hypertarget{ref-Crystal2002}{}}%
Crystal, David. 2002. \emph{Language death}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Egghe2007}{}}%
Egghe, Leo. 2007. Untangling herdan's law and heaps' law: Mathematical
and informetric arguments. \emph{Journal of the American Society for
Information Science and Technology}. Wiley Online Library 58(5).
702--709.

\leavevmode\vadjust pre{\hypertarget{ref-EvansLevinson2009}{}}%
Evans, Nicholas \& Stephen C. Levinson. 2009. The myth of language
universals: Language diversity and its importance for cognitive science.
\emph{Behavioral and Brain Sciences}. Cambridge University Press 32(5).
429--448.

\leavevmode\vadjust pre{\hypertarget{ref-Everett2018}{}}%
Everett, Caleb. 2018. The similar rates of occurrence of consonants
across the world's languages: A quantitative analysis of phonetically
transcribed word lists. \emph{Language Sciences}. Elsevier 69. 125--135.

\leavevmode\vadjust pre{\hypertarget{ref-FirchowFirchow1969}{}}%
Firchow, Irwin \& Jacqueline Firchow. 1969. {An Abbreviated Phoneme
Inventory}. \emph{Anthropological Linguistics} 11. 271--276.

\leavevmode\vadjust pre{\hypertarget{ref-GoodCysouw2013}{}}%
Good, Jeff \& Michael Cysouw. 2013. Languoid, doculect, and glossonym:
Formalizing the notion'language'. \emph{Language Documentation and
Conservation} 7. 331--359.

\leavevmode\vadjust pre{\hypertarget{ref-Hayes2009}{}}%
Hayes, Bruce. 2009. \emph{Introductory phonology}. Blackwell.

\leavevmode\vadjust pre{\hypertarget{ref-Heaps1978}{}}%
Heaps, Harold Stanley. 1978. \emph{Information retrieval, computational
and theoretical aspects}. Academic Press.

\leavevmode\vadjust pre{\hypertarget{ref-Herdan1964}{}}%
Herdan, Gustav. 1964. Quantitative linguistics. Butterworth.

\leavevmode\vadjust pre{\hypertarget{ref-Hockett1963}{}}%
Hockett, Charles. 1963. The problem of universals in language. In Joseph
H. Greeberg (ed.), \emph{Universals of language}. Cambridge, MA: MIT
Press.

\leavevmode\vadjust pre{\hypertarget{ref-Hockett1995-phoneme}{}}%
Hockett, Charles F. 1995. The birth and deaths of the phoneme.
\emph{Id{é}ologies dans le Monde Anglo-Saxon} 8. 5--50.

\leavevmode\vadjust pre{\hypertarget{ref-Joos1948}{}}%
Joos, Martin. 1948. Acoustic phonetics. \emph{Language}. Linguistic
Society of America 24(2). 5--136. (7 December, 2022).

\leavevmode\vadjust pre{\hypertarget{ref-JustesonStephens1984}{}}%
Justeson, John S. \& Laurence D. Stephens. 1984. On the relationship
between the numbers of vowels and consonants in phonological systems.
\emph{Linguistics} 22. 531--545.

\leavevmode\vadjust pre{\hypertarget{ref-Kornai2002}{}}%
Kornai, András. 2002. {How many words are there?} \emph{Glottometrics}
4. 61--86.

\leavevmode\vadjust pre{\hypertarget{ref-Ladefoged_etal1972}{}}%
Ladefoged, Peter, Joseph DeClerk, Mona Lindau \& George Papcun. 1972. An
auditory-motor theory of speech production. \emph{{UCLA} working papers
in phonetics}. UCLA Los Angeles, CA 22(48). 48--76.

\leavevmode\vadjust pre{\hypertarget{ref-Lehfeldt1975}{}}%
Lehfeldt, W. 1975. {Die Verteilung der Phonemanzahl in den nat{ü}rlichen
Sprachen}. \emph{Phonetica} 31. 247--287.

\leavevmode\vadjust pre{\hypertarget{ref-vanLeijenhorst2005}{}}%
Leijenhorst, Dick C van \& Th P Van der Weide. 2005. A formal derivation
of heaps' law. \emph{Information Sciences}. Elsevier 170(2-4). 263--272.

\leavevmode\vadjust pre{\hypertarget{ref-Lindblom1990}{}}%
Lindblom, Björn. 1990. Explaining phonetic variation: A sketch of the
h\&h theory. In W. J. Hardcastle \& A. Marchal (eds.), \emph{Speech
production and speech modelling}, 403--439. Springer.
https://doi.org/\url{https://doi.org/10.1007/978-94-009-2037-8_16}.

\leavevmode\vadjust pre{\hypertarget{ref-Macklin-CordesRound2020}{}}%
Macklin-Cordes, Jayden L \& Erich R Round. 2020. Re-evaluating phoneme
frequencies. \emph{Frontiers in Psychology}. Frontiers Media SA 11.
570895.

\leavevmode\vadjust pre{\hypertarget{ref-Maddieson1984}{}}%
Maddieson, Ian. 1984. \emph{Patterns of sounds}. Cambridge: Cambridge
University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Maddieson2008}{}}%
Maddieson, Ian. 2008. Consonant inventories. In Martin Haspelmath,
Matthew S. Dryer, David Gil \& Bernard Comrie (eds.), \emph{The world
atlas of language structures online}. Munich: Max Planck Digital
Library.

\leavevmode\vadjust pre{\hypertarget{ref-MaddiesonPrecoda1990}{}}%
Maddieson, Ian \& Kristin Precoda. 1990. {Updating UPSID}. In \emph{UCLA
working papers in phonetics}, vol. 74, 104--111. Department of
Linguistics, UCLA.

\leavevmode\vadjust pre{\hypertarget{ref-Martindale_etal1996}{}}%
Martindale, Colin, S. M. Gusein‐Zade, Dean McKenzie \& Mark Yu.
Borodovsky. 1996. Comparison of equations describing the ranked
frequency distributions of graphemes and phonemes. \emph{Journal of
Quantitative Linguistics}. Routledge 3(2). 106--112.

\leavevmode\vadjust pre{\hypertarget{ref-Mielke2008}{}}%
Mielke, Jeff. 2008. \emph{{The Emergence of Distinctive Features}}.
Oxford University Press.

\leavevmode\vadjust pre{\hypertarget{ref-MielkeHume2006}{}}%
Mielke, Jeff \& Elizabeth Hume. 2006. Distinctive features. In
\emph{Encylopedia of langauge}. Elsevier.

\leavevmode\vadjust pre{\hypertarget{ref-Milivcka2009}{}}%
Milička, Jiři. 2009. Type-token \& hapax-token relation: A combinatorial
model. \emph{Glottotheory}. De Gruyter Akademie Forschung 2(1). 99--110.

\leavevmode\vadjust pre{\hypertarget{ref-Moran2012}{}}%
Moran, Steven. 2012. \emph{{Phonetics Information Base and Lexicon}}.
University of Washington PhD thesis.

\leavevmode\vadjust pre{\hypertarget{ref-MoranMcCloy2019}{}}%
Moran, Steven \& Daniel McCloy (eds.). 2019. \emph{{PHOIBLE 2.0}}. Jena:
Max Planck Institute for the Science of Human History.

\leavevmode\vadjust pre{\hypertarget{ref-Pagel2000}{}}%
Pagel, Mark. 2000. The history, rate and pattern of world linguistic
evolution. In Chris Knight, Michael Studdert-Kennedy \& James Hurford
(eds.), \emph{The evolutionary emergence of language: Social function
and the origins of linguistic form}, 391--416. Cambridge University
Press.

\leavevmode\vadjust pre{\hypertarget{ref-Piantadosi2014}{}}%
Piantadosi, Steven T. 2014. Zipf's word frequency law in natural
language: A critical review and future directions. \emph{Psychonomic
bulletin \& review}. Springer 21(5). 1112--1130.

\leavevmode\vadjust pre{\hypertarget{ref-Potter1945}{}}%
Potter, Ralph K. 1945. Visible patterns of sound. \emph{Science}.
American Association for the Advancement of Science 102(2654). 463--470.

\leavevmode\vadjust pre{\hypertarget{ref-Snyman1970}{}}%
Snyman, J. W. 1970. \emph{{An Introduction to the !Xu (!Kung)
Language}}. Capetown: A. A. Balkema.

\leavevmode\vadjust pre{\hypertarget{ref-Snyman1975}{}}%
Snyman, J. W. 1975. \emph{{Zu\textbar'hoasi Fonologie en Woordeboek}}.
Cape Town: Balkema.

\leavevmode\vadjust pre{\hypertarget{ref-Stevens1972}{}}%
Stevens, Kenneth N. 1972. The quantal nature of speech: Evidence from
articulatory-acoustic data. In Edward E David \& Peter B Denes (eds.),
\emph{Human communication: A unified view}. McGraw-Hill.

\leavevmode\vadjust pre{\hypertarget{ref-TambovtsevMartindale2007}{}}%
Tambovtsev, Yuri \& Colin Martindale. 2007. Phoneme frequencies follow a
yule distribution. \emph{SKASE Journal of Theoretical Linguistics}.
Citeseer 4(2). 1--11.

\leavevmode\vadjust pre{\hypertarget{ref-Whalen2019}{}}%
Whalen, Douglas H. 2016. Phonetics. In \emph{Oxford research
encyclopedia of linguistics}. Oxford University Press.

\leavevmode\vadjust pre{\hypertarget{ref-Zipf1936}{}}%
Zipf, George Kingsley. 1936. \emph{The psychobiology of language}.
London: Routledge.

\leavevmode\vadjust pre{\hypertarget{ref-Zipf1949}{}}%
Zipf, George Kingsley. 1949. \emph{Human behavior and the principle of
least effort}. New York: Addison-Wesley.

\end{CSLReferences}

\end{document}

### 2023-06-06 paper.Rmd
edits

@@
 We find that both models produce accurate estimates of the diversity of contrastive speech sounds (phonemes) in existing typological databases.
-Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) documenting more phonological repertoires always yield more phoneme types cross-linguistically.
+Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) documenting more phonological repertoires cross-linguistically always yields more phoneme types.
 As a sanity check, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates align with Herdan-Heaps' law.
@@
 To our knowledge, we are the first to apply Herdan-Heaps' law to the domain of segmental phonology by reformulating it as estimator of the number of phonemes in groups of languages.
-Finally, we verify the goodness-of-fit of Herdan-Heaps' by applying an additional and formally unrelated tool borrowed from Ecology, which we refer to as the Chao-Shen estimator.
+Finally, we verify the goodness-of-fit of Herdan-Heaps' by applying an additional and formally unrelated tool borrowed from ecology, which we refer to as the Chao-Shen estimator.
 The predictions of this estimator confirm that the relationship between sample size and segment type count is best described by a curve that approximates Herdan-Heaps' law.
@@
 Based on PHOIBLE, our current best guess of the diversity of documented phoneme types is 3183.
-Therefore, the models argue for the existence of approximately 1500 living but undocumented phoneme types.
+Therefore, the models argue for the existence of approximately 1500 yet to be documented phoneme types.
 As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30k-500k [@Pagel2000] or 64k-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19K.
@@
 However, when we compare the predictions of Herdan-Heaps and Chao-Shen  against known type counts in a closed dataset, the source of the discrepancy emerged: as the sample grows, Chao-Shen lags behind Herdan-Heaps in approximating the true value.
-Thus, given a larger pool of training data, we expect Chao-Shen to approximate more closely the estimates of Herdan-Heaps.
+Thus, given a larger pool of training data, we expect Chao-Shen to approximate more closely the estimates of Herdan-Heaps' law.
 
@@
 And looming in the background of these discussions, the debate of whether distinctive features are innate and universal or rather learned and emergent continues.
-Nevertheless, in general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors ($2^24$).
+Nevertheless, in general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors (2^24).
 Current consensus aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature.
@@
 
-Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond whatever the phonetics-phonology interface may be.
-Thus, we believe the approach presented here has broad applications to the study of other areas of grammar and the cross-linguistic study of languages.
+Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond whatever the phonetics-phonology interface may or may not be.
+Thus, we believe the approach presented here has broad applications to the study of other areas of grammar and to the cross-linguistic study of languages.
 
@@
 
-Steven Moran was funded by the Swiss National Science Foundation (Grant No. PCEFP1_186841).
+SM was funded by the Swiss National Science Foundation (Grant No. PCEFP1_186841).
+
+# Author contributions
+
+SM conceptualization. SM and NAL designed the study, interpreted the results, and wrote the paper.
 

### 2023-06-06 paper.tex
edits

@@
 phonemes are difficult to detect because they are rare, and (b)
-documenting more phonological repertoires always yield more phoneme
-types cross-linguistically. As a sanity check, we apply an unrelated
+documenting more phonological repertoires cross-linguistically always
+yields more phoneme types. As a sanity check, we apply an unrelated
 estimator of species diversity from biology, Chao-Shen, and show that
@@
 applying an additional and formally unrelated tool borrowed from
-Ecology, which we refer to as the Chao-Shen estimator. The predictions
+ecology, which we refer to as the Chao-Shen estimator. The predictions
 of this estimator confirm that the relationship between sample size and
@@
 of documented phoneme types is 3183. Therefore, the models argue for the
-existence of approximately 1500 living but undocumented phoneme types.
-As we probe increasingly larger and theoretically motivated language
-sample sizes, i.e., languages that existed \emph{through time} (e.g.,
-estimates including 30k-500k (\protect\hyperlink{ref-Pagel2000}{Pagel
-2000}) or 64k-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})),
-the Herdan-Heaps' model predicted a continual increase in the number of
+existence of approximately 1500 yet to be documented phoneme types. As
+we probe increasingly larger and theoretically motivated language sample
+sizes, i.e., languages that existed \emph{through time} (e.g., estimates
+including 30k-500k (\protect\hyperlink{ref-Pagel2000}{Pagel 2000}) or
+64k-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})), the
+Herdan-Heaps' model predicted a continual increase in the number of
 unique phonemes, reaching values as high as 19K. The Chao-Shen model
@@
 value. Thus, given a larger pool of training data, we expect Chao-Shen
-to approximate more closely the estimates of Herdan-Heaps.
+to approximate more closely the estimates of Herdan-Heaps' law.
 
@@
 2006}), which still has the potential to generate over 16 million
-distinct feature vectors (\(2^24\)). Current consensus aside, there does
+distinct feature vectors (2\^{}24). Current consensus aside, there does
 not appear to be a principled limit to the degree of phonetic
@@
 elements is a core question of language science that goes beyond
-whatever the phonetics-phonology interface may be. Thus, we believe the
-approach presented here has broad applications to the study of other
-areas of grammar and the cross-linguistic study of languages.
+whatever the phonetics-phonology interface may or may not be. Thus, we
+believe the approach presented here has broad applications to the study
+of other areas of grammar and to the cross-linguistic study of
+languages.
 
@@
 
-Steven Moran was funded by the Swiss National Science Foundation (Grant
+SM was funded by the Swiss National Science Foundation (Grant
 No.~PCEFP1\_186841).
+
+\hypertarget{author-contributions}{%
+\section{Author contributions}\label{author-contributions}}
+
+SM conceptualization. SM and NAL designed the study, interpreted the
+results, and wrote the paper.
 

### 2023-06-21 paper.Rmd
updates

@@
 Moreover, nearly half of the reported segment types are singletons in the sample, i.e., one-off language-specific occurrences.^[Whether rare segments occur more or less frequently *within* a language is a language-by-language research question.
-See for example [@Everett2018].] While Maddieson did not explicitly test for it, his observations fit a well-attested pattern known as Zipf's law [@Zipf1936;@Zipf1949].
+See for example @Everett2018.] While Maddieson did not explicitly test for it, his observations fit a well-attested pattern known as Zipf's law [@Zipf1936;@Zipf1949].
 Zipf's law states that the frequency distribution of words in a language is inversely proportional to their frequency rank.

### 2023-06-21 paper.tex
adds paper
updates

@@
 \author{Steven Moran \& Nicholas A. Lester}
-\date{06 June, 2023}
+\date{21 June, 2023}
 
@@
   or less frequently \emph{within} a language is a language-by-language
-  research question. See for example
-  (\protect\hyperlink{ref-Everett2018}{Everett 2018}).} While Maddieson
-did not explicitly test for it, his observations fit a well-attested
-pattern known as Zipf's law (\protect\hyperlink{ref-Zipf1936}{Zipf
-1936}; \protect\hyperlink{ref-Zipf1949}{Zipf 1949}). Zipf's law states
-that the frequency distribution of words in a language is inversely
-proportional to their frequency rank. The relationship is log-linear:
-high frequency words are few in number but account for most of the
-tokens in a text, and low frequency words are many, but occur rarely or
-even just once.\footnote{Here is a plot showing Zipf's law in the first
-  10 million words in the Wikipedias from 30 different languages:
+  research question. See for example Everett
+  (\protect\hyperlink{ref-Everett2018}{2018}).} While Maddieson did not
+explicitly test for it, his observations fit a well-attested pattern
+known as Zipf's law (\protect\hyperlink{ref-Zipf1936}{Zipf 1936};
+\protect\hyperlink{ref-Zipf1949}{Zipf 1949}). Zipf's law states that the
+frequency distribution of words in a language is inversely proportional
+to their frequency rank. The relationship is log-linear: high frequency
+words are few in number but account for most of the tokens in a text,
+and low frequency words are many, but occur rarely or even just
+once.\footnote{Here is a plot showing Zipf's law in the first 10 million
+  words in the Wikipedias from 30 different languages:
   \url{https://en.wikipedia.org/wiki/Zipf\%27s_law\#/media/File:Zipf_30wiki_en_labels.png}.}

### 2023-06-22 paper.Rmd
edits

@@
 In Section 2, we motivate our research questions.
-In section 3, we describe the data and in Section 4 we present our methods and results.
+In Section 3, we describe the data and in Section 4 we present our methods and results.
 In Section 5, we discuss our results and avenues for future research.
@@
 Ellis (1814--1890) stated as early as 1848 that the number of different speech sounds, i.e., candidate phonemes, is in principle infinite.
-At that time, teachers and philologists aimed to develop a transcription standard that could discriminate between all manners of speech sounds with orthographic symbols for purposes that included (particularly English) spelling reform and orthography development for newly documented and unwritten languages.
+At that time, teachers and philologists aimed to develop a transcription standard that could discriminate between all manners of speech sounds with orthographic symbols for purposes that included (particularly English) spelling reform and orthography development for newly documented and unwritten languages (e.g., for bible translation).
 But as practitioners increasingly converted the articulations of more and more languages into a universal phonetic alphabet, the need for new symbols also seemingly increased without limit [@Hockett1995-phoneme, pg.
@@
 For speech, it measured fluctuations in air pressure, which capture duration, pitch, and intensity.
-The recorded speech waveforms showed visually that speech is dynamic and continuous, even if listeners perceive it as discrete, e.g., phonemes, syllables, words.^[This relationship between continuous and discrete is at the core of the debate about the interface between phonetics and phonology [@Whalen2019].
+Thus, the ability to record, measure, and visualize speech shows that it is dynamic and continuous, even if listeners perceive it as discrete, e.g., phonemes, syllables, words.^[This relationship between continuous and discrete is at the core of the debate about the interface between phonetics and phonology [@Whalen2019].
 The division goes back to at least the late 19th century when phoneticians and dialectologists including Henry Sweet (1845--1912; 1877) and Jost Winteler (1846--1929; 1876) began making a distinction in language descriptions between sounds that change the meaning of words and those that do not.]
@@
 
-The Zipfian distribution characterizes many aspects of language.
+The Zipfian distribution characterizes many aspects of language (and other natural phenomena).
 Two features make it interesting for our purposes.
@@
 ```{r upsid_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}Cumulative segments in UPSID-451 follow Zipf's law"}
-knitr::include_graphics("README_files/figure-gfm/zipf_distribution_upsid-1.png")
+knitr::include_graphics("README_files/figure-gfm/heaps_upsid-1.png")
 ```
@@
 ```{r phoible_heaps, echo=FALSE, fig.align="center", fig.cap="\\label{fig:upsid}: Cumulative segments in PHOIBLE (OIPL) follow Zipf's law"}
-knitr::include_graphics("README_files/figure-gfm/zipf_distribution_phoible-1.png")
+knitr::include_graphics("README_files/figure-gfm/heaps_phoible_oipl-1.png")
 ```

### 2023-06-22 paper.tex
edits

@@
 \author{Steven Moran \& Nicholas A. Lester}
-\date{21 June, 2023}
+\date{22 June, 2023}
 
@@
 how many phonemes are possible in human language? In Section 2, we
-motivate our research questions. In section 3, we describe the data and
+motivate our research questions. In Section 3, we describe the data and
 in Section 4 we present our methods and results. In Section 5, we
@@
 included (particularly English) spelling reform and orthography
-development for newly documented and unwritten languages. But as
-practitioners increasingly converted the articulations of more and more
-languages into a universal phonetic alphabet, the need for new symbols
-also seemingly increased without limit
-(\protect\hyperlink{ref-Hockett1995-phoneme}{Hockett 1995}, pg.
+development for newly documented and unwritten languages (e.g., for
+bible translation). But as practitioners increasingly converted the
+articulations of more and more languages into a universal phonetic
+alphabet, the need for new symbols also seemingly increased without
+limit (\protect\hyperlink{ref-Hockett1995-phoneme}{Hockett 1995}, pg.
 63).\footnote{This early work later became the basis for the
@@
 measured fluctuations in air pressure, which capture duration, pitch,
-and intensity. The recorded speech waveforms showed visually that speech
-is dynamic and continuous, even if listeners perceive it as discrete,
-e.g., phonemes, syllables, words.\footnote{This relationship between
-  continuous and discrete is at the core of the debate about the
-  interface between phonetics and phonology
+and intensity. Thus, the ability to record, measure, and visualize
+speech shows that it is dynamic and continuous, even if listeners
+perceive it as discrete, e.g., phonemes, syllables, words.\footnote{This
+  relationship between continuous and discrete is at the core of the
+  debate about the interface between phonetics and phonology
   (\protect\hyperlink{ref-Whalen2019}{Whalen 2016}). The division goes
@@
 
-The Zipfian distribution characterizes many aspects of language. Two
-features make it interesting for our purposes. First, it describes
-systems for which most types are rare, hence difficult to observe.
-Second, under certain conditions (\(\alpha \ge 1\)), the Zipf function
-is non-asymptotic: more types can always be added at the rank for
-singletons, and the expected type count computed. The fact that
-Maddieson's observations are consistent with the Zipfian distribution
-suggests that phoneme types are indeed difficult to discover
-(particularly at the fringe) and that there could be some room left for
-additional phonemes.
+The Zipfian distribution characterizes many aspects of language (and
+other natural phenomena). Two features make it interesting for our
+purposes. First, it describes systems for which most types are rare,
+hence difficult to observe. Second, under certain conditions
+(\(\alpha \ge 1\)), the Zipf function is non-asymptotic: more types can
+always be added at the rank for singletons, and the expected type count
+computed. The fact that Maddieson's observations are consistent with the
+Zipfian distribution suggests that phoneme types are indeed difficult to
+discover (particularly at the fringe) and that there could be some room
+left for additional phonemes.
 
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/zipf_distribution_upsid-1}
+{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/heaps_upsid-1}
 
@@
 
-{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/zipf_distribution_phoible-1}
+{\centering \includegraphics[width=0.8\linewidth]{README_files/figure-gfm/heaps_phoible_oipl-1}
 

### 2023-08-04 README.Rmd
updated SI
updates
submitted version

@@
 output:
-  github_document:
-      toc: true
+  pdf_document:
+    number_sections: true
+    latex_engine: xelatex    
 bibliography: 'bibliography.bib'
+biblio-style: "apalike"
+csl: unified-style-sheet-for-linguistics.csl
+link-citations: true
 ---
@@
 ```{r cumulative_plot_upsid}
-qplot(id, results, data = upsid_cumulative, xlab="Number of languages (random order)", ylab="Cumulative number of segment types") + theme_bw()
+qplot(id, results, data = upsid_cumulative, 
+      xlab="Number of languages (random order)", 
+      ylab="Cumulative number of segment types") + 
+  theme_bw()
 ```
@@
 ```{r, cumulative_plot_phoible}
-qplot(id, results, data = phoible_cumulative, xlab="Number of languages (random order)", ylab="Cumulative number of segment types") + theme_bw()
+qplot(id, results, data = phoible_cumulative, 
+      xlab="Number of languages (random order)", 
+      ylab="Cumulative number of segment types") + 
+  theme_bw()
 ```
@@
 ```{r}
-plot(segments_upsid$Rank, segments_upsid$Frequency, xlab = "Rank", ylab = "Frequency", main = "UPSID-451 segment distribution")
-plot(segments_phoible$Rank, segments_phoible$Frequency, xlab = "Rank", ylab = "Frequency", main = "PHOIBLE phoneme distribution")
+plot(segments_upsid$Rank, segments_upsid$Frequency, 
+     xlab = "Rank", 
+     ylab = "Frequency", 
+     main = "UPSID-451 segment distribution")
+
+plot(segments_phoible$Rank, segments_phoible$Frequency, 
+     xlab = "Rank", 
+     ylab = "Frequency", 
+     main = "PHOIBLE phoneme distribution")
 ```
@@
 segments_gt_1_upsid <- segments_upsid %>% filter(Frequency > 1)
-plot(segments_gt_1_upsid$Rank, segments_gt_1_upsid$Frequency, xlab = "Rank", ylab = "Frequency", main = "UPSID segment distribution")
+plot(segments_gt_1_upsid$Rank, segments_gt_1_upsid$Frequency, 
+     xlab = "Rank", 
+     ylab = "Frequency", 
+     main = "UPSID segment distribution")
 
 segments_gt_1_phoible <- segments_phoible %>% filter(Frequency > 1)
-plot(segments_gt_1_phoible$Rank, segments_gt_1_phoible$Frequency, xlab = "Rank", ylab = "Frequency", main = "PHOIBLE segment distribution")
+plot(segments_gt_1_phoible$Rank, segments_gt_1_phoible$Frequency, 
+     xlab = "Rank", 
+     ylab = "Frequency", 
+     main = "PHOIBLE segment distribution")
 ```
@@
 
-# Estimating the parameters -- TODO have set c lower
-fitted <- nls(y ~ zipf(x_rank, a, b, c), data = df, start = list(a = -1.3, b = 18, c = 100000), trace = T)
+# Estimating the parameters
+fitted <- nls(y ~ zipf(x_rank, a, b, c), 
+              data = df, 
+              start = list(a = -1.3, b = 18, c = 100000), 
+              trace = T)
 
@@
 
-plot(x, y, xlab = "Rank (log)", ylab = "Frequency (log)", main = "Zipf's law on PHOIBLE (OIPL)")
+plot(x, y, 
+     xlab = "Rank (log)", 
+     ylab = "Frequency (log)", 
+     main = "Zipf's law on PHOIBLE (OIPL)")
 
@@
 # Estimating the parameters
-fitted <- nls(y ~ zipf(x_rank, a, b, c), data = df, start = list(a = -1.3, b = 18, c = 1000000), trace = T)
+fitted <- nls(y ~ zipf(x_rank, a, b, c), 
+              data = df, 
+              start = list(a = -1.3, b = 18, c = 1000000), 
+              trace = T)
 
@@
 
-# Settings of K and B based on trial and error -- not the optimal values, but the result is reasonably close to the true value.
-Therefore, if we do any model tuning, the values we arrive at should be close to these; otherwise, we most likely have a bad fit of Herdan-Heaps' to the data.
+# Settings of K and B based on trial and error -- not the optimal values, but 
+# the result is reasonably close to the true value. Therefore, if we do any 
+# model tuning, the values we arrive at should be close to these; otherwise, 
+# we most likely have a bad fit of Herdan-Heaps' to the data. 
 heaps(8, n, .5)
@@
 ```{r}
-fitHeaps.upsid <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data = heaps_results_upsid)
+fitHeaps.upsid <- 
+  lmer(log(type_count) ~ log(token_count) + (1|iteration), 
+       data = heaps_results_upsid)
 
@@
 ```{r}
-fitHeaps.phoible <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data = heaps_results_phoible)
+fitHeaps.phoible <- 
+  lmer(log(type_count) ~ log(token_count) + (1|iteration), 
+       data = heaps_results_phoible)
 
@@
 # Fit the model
-fitHeaps.phoible.all <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data = heaps_results_phoible_all)
+fitHeaps.phoible.all <- 
+  lmer(log(type_count) ~ log(token_count) + (1|iteration), 
+       data = heaps_results_phoible_all)
 
@@
 
+\newpage
+
 ```{r}
@@
 
-# Function to estimate the number of unseen types from a proposed assemblage (used to compute our estimate of Q0)
+# Function to estimate the number of unseen types from a proposed assemblage 
+# (used to compute our estimate of Q0)
 S_Chao2 <- function(vector, R) {
@@
 # r = 10
-chao.iter.plot.10 <- choa.iter.plotter(chao.data.10.10, mean.obs.df.10, "r = 10") + ylim(0,4000)
+chao.iter.plot.10 <- 
+  choa.iter.plotter(chao.data.10.10, mean.obs.df.10, "r = 10") + 
+  ylim(0,4000)
 
@@
 # r = 50
-chao.iter.plot.50 <- choa.iter.plotter(chao.data.50.10, mean.obs.df.50, "r = 50") + ylim(0,4000)
+chao.iter.plot.50 <- 
+  choa.iter.plotter(chao.data.50.10, mean.obs.df.50, "r = 50") + 
+  ylim(0,4000)
 
@@
 # r = 100
-chao.iter.plot.100 <- choa.iter.plotter(chao.data.100.10, mean.obs.df.100, "r = 100") + ylim(0,4000)
+chao.iter.plot.100 <- 
+  choa.iter.plotter(chao.data.100.10, mean.obs.df.100, "r = 100") + 
+  ylim(0,4000)
 
@@
 # r = 500
-chao.iter.plot.500 <- choa.iter.plotter(chao.data.500.10, mean.obs.df.500, "r = 500") + ylim(0,4000)
+chao.iter.plot.500 <- 
+  choa.iter.plotter(chao.data.500.10, mean.obs.df.500, "r = 500") + 
+  ylim(0,4000)
 
@@
 # r = 1000
-chao.iter.plot.1000 <- choa.iter.plotter(chao.data.1000.10, mean.obs.df.1000, "r = 1000") + ylim(0,4000)
+chao.iter.plot.1000 <- 
+  choa.iter.plotter(chao.data.1000.10, mean.obs.df.1000, "r = 1000") + 
+  ylim(0,4000)
 
@@
 # r = 1500
-chao.iter.plot.1500 <- choa.iter.plotter(chao.data.1500.10, mean.obs.df.1500, "r = 1500") + ylim(0,4000)
+chao.iter.plot.1500 <- 
+  choa.iter.plotter(chao.data.1500.10, mean.obs.df.1500, "r = 1500") + 
+  ylim(0,4000)
 
@@
 # r = 2000
-chao.iter.plot.2000 <- choa.iter.plotter(chao.data.2000.10, mean.obs.df.2000, "r = 2000") + ylim(0,4000)
+chao.iter.plot.2000 <- 
+  choa.iter.plotter(chao.data.2000.10, mean.obs.df.2000, "r = 2000") + 
+  ylim(0,4000)
+
 chao.iter.plot.2000
@@
 
-annotate_figure(increm.chao.plots, bottom = "Number of languages (R)", left = "Estimated number of types")
+annotate_figure(increm.chao.plots, bottom = "Number of languages (R)", 
+                left = "Estimated number of types")
 ```
@@
 
-chao.iter.diff.plot.10 <- diff.plotter(chao.data.10.10, mean.obs.df.10, "r=10") + ylim(-2500, 400)
+chao.iter.diff.plot.10 <- 
+  diff.plotter(chao.data.10.10, mean.obs.df.10, "r=10") + 
+  ylim(-2500, 400)
 
@@
 
-chao.iter.diff.plot.50 <- diff.plotter(chao.data.50.10, mean.obs.df.50,"r=50") + ylim(-2500, 400)
+chao.iter.diff.plot.50 <- 
+  diff.plotter(chao.data.50.10, mean.obs.df.50,"r=50") + 
+  ylim(-2500, 400)
 
@@
 
-chao.iter.diff.plot.100 <- diff.plotter(chao.data.100.10, mean.obs.df.100,"r=100") + ylim(-2500, 400)
+chao.iter.diff.plot.100 <- 
+  diff.plotter(chao.data.100.10, mean.obs.df.100,"r=100") + 
+  ylim(-2500, 400)
 
@@
 
-chao.iter.diff.plot.500 <- diff.plotter(chao.data.500.10, mean.obs.df.500,"r=500") + ylim(-2500, 400)
+chao.iter.diff.plot.500 <- 
+  diff.plotter(chao.data.500.10, mean.obs.df.500,"r=500") + 
+  ylim(-2500, 400)
 
@@
 
-chao.iter.diff.plot.1000 <- diff.plotter(chao.data.1000.10, mean.obs.df.1000,"r=1000") + ylim(-2500, 400)
+chao.iter.diff.plot.1000 <- 
+  diff.plotter(chao.data.1000.10, mean.obs.df.1000,"r=1000") + 
+  ylim(-2500, 400)
 
@@
 
-chao.iter.diff.plot.1500 <- diff.plotter(chao.data.1500.10, mean.obs.df.1500,"r=1500") + ylim(-2500, 400)
+chao.iter.diff.plot.1500 <- 
+  diff.plotter(chao.data.1500.10, mean.obs.df.1500,"r=1500") + 
+  ylim(-2500, 400)
 
@@
 
-chao.iter.diff.plot.2000 <- diff.plotter(chao.data.2000.10, mean.obs.df.2000,"r=2000")
+chao.iter.diff.plot.2000 <- 
+  diff.plotter(chao.data.2000.10, mean.obs.df.2000,"r=2000")
 
@@
 # Fit the model
-fitHeaps.phoible.rand.1k <- lmer(log(type_count) ~ log(token_count) + (1|iteration), data=rand.pho.1k.results)
+fitHeaps.phoible.rand.1k <- 
+  lmer(log(type_count) ~ log(token_count) + (1|iteration), data=rand.pho.1k.results)
 
@@
 ```{r diff_plots_ext}
-plot.diff.dat <- data.frame(Type.ct.est = c(pho.mean.extrap, predictions, chao.ext), Estimator = factor(rep(c("HH Approach 1", "HH Approach 2", "Chao-Shen"), each = 4)), Sample.Size = factor(rep(c("7151", "10K", "50K", "100K"), 3)))
-
-plot.diff.dat$Sample.Size = factor(plot.diff.dat$Sample.Size, levels = levels(plot.diff.dat$Sample.Size)[c(4, 2, 3, 1)])
+plot.diff.dat <- 
+  data.frame(Type.ct.est = c(pho.mean.extrap, predictions, chao.ext), 
+             Estimator = factor(rep(c("HH Approach 1", "HH Approach 2", "Chao-Shen"), each = 4)), 
+             Sample.Size = factor(rep(c("7151", "10K", "50K", "100K"), 3)))
+
+plot.diff.dat$Sample.Size = 
+  factor(plot.diff.dat$Sample.Size, 
+         levels = levels(plot.diff.dat$Sample.Size)[c(4, 2, 3, 1)])
 

### 2023-08-04 paper.Rmd
submitted version

@@
 title: "How many speech sounds are there in the world's languages?"
-author: "Steven Moran & Nicholas A. Lester\n"
+author: |
+  | Steven Moran^1,2^ & Nicholas A. Lester^3^
+  | ^1^Institute of Biology, University of Neuchâtel, Neuchâtel, Switzerland
+  | ^2^Department of Anthropology, University of Miami, Coral Gables, FL, USA
+  | ^3^Department of Linguistics, University of North Texas, Denton, TX, USA
+  | OrcID, SM, 0000-0002-3969-6549
+  | Corresponding authors: steven.moran@unine.ch; nicholas.a.lester@gmail.com
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 We find that both models produce accurate estimates of the diversity of contrastive speech sounds (phonemes) in existing typological databases.
-Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) documenting more phonological repertoires cross-linguistically always yields more phoneme types.
-As a sanity check, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates align with Herdan-Heaps' law.
-We use both models to extrapolate phoneme counts to undocumented languages, resulting in an estimate of 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k), which is roughly 1.5x more than what is currently documented in the world's languages (~35%).
+Our findings suggest that (a) most phonemes are difficult to detect because they are rare, and (b) increased documentation of phonological inventories always yields more phoneme types.
+To check this, we apply an unrelated estimator of species diversity from biology, Chao-Shen, and show that its estimates also align with Herdan-Heaps' law.
+We use both models to extrapolate phoneme counts to undocumented languages, resulting in an estimate of 4600-4900 distinct phonemes in the full set of languages thought to be spoken today (~7-8k), which is roughly 1.5 time more than what is currently documented in the world's languages (~35%).
 By extrapolating to hypothetical sample sizes, e.g., the number of languages estimated to have been spoken through time, and beyond, we find no indication of convergence: infinite languages mean infinite phonemes.
@@
 Therefore, the models argue for the existence of approximately 1500 yet to be documented phoneme types.
-As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30k-500k [@Pagel2000] or 64k-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19K.
-The Chao-Shen model produced much more conservative estimates for samples of size 10K and above, even appearing to settle on an asymptote of roughly 5500 phonemes for samples of 100K languages or more.
+As we probe increasingly larger and theoretically motivated language sample sizes, i.e., languages that existed *through time* (e.g., estimates including 30-500k [@Pagel2000] or 64-140k [@Crystal2002]), the Herdan-Heaps' model predicted a continual increase in the number of unique phonemes, reaching values as high as 19k.
+The Chao-Shen model produced much more conservative estimates for samples of size 10k and above, even appearing to settle on an asymptote of roughly 5500 phonemes for samples of 100k languages or more.
 However, when we compare the predictions of Herdan-Heaps and Chao-Shen  against known type counts in a closed dataset, the source of the discrepancy emerged: as the sample grows, Chao-Shen lags behind Herdan-Heaps in approximating the true value.
@@
 Thus, based on the evidence presented here, we have no reason to believe that there is an upper bound on the number of unique phonemes possible in spoken languages.
-Infinite linguistic diversity seems to breed infinite possible phonemes.
+Infinite linguistic diversity seems to bring about infinite possible phonemes.
 But what would it mean if there were indeed infinitely many potential phonemes?
 
-If one adopts a feature-based perspective on the phonetic targets of phonemes, then there is an *a priori* upper bound of $2^F$ contrastive sounds, where $F$ is the number of feature categories and 2 reflects the choice between presence (+) or absence (-) of that feature.^[In some feature sets, some values may be 0, i.e., undefined, underspecified, or ``for which a segment can be said 'not to care'" [@Hayes2009, pg.
+If one adopts a distinctive feature-based perspective on the phonetic targets of phonemes, then there is an *a priori* upper bound of $2^F$ contrastive sounds, where $F$ is the number of feature categories and 2 reflects the choice between presence (+) or absence (-) of that feature.^[In some distinctive feature sets, some values may be 0, i.e., undefined, underspecified, or ``for which a segment can be said 'not to care'" [@Hayes2009, pg.
 91].
@@
 For example, given a feature set such as in PHOIBLE with 37 distinctive features, an upper bound on the number of segments under a strictly binary formalization is over 137 billion (2^37) feature vectors.
+In general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors (2^24).
+Crucially, this theoretical bound on phoneme diversity does not in itself preclude the possibility of infinitely many phonemes.
+To get there, one would only need to discover infinitely many distinctive features.
+But, as pointed out by @Lindblom1990a, most proponents of feature-based analysis assume a fixed, small-ish inventory of features.
+The question would then be: how does the feature pool grow empirically with the addition of new languages?
+Does it converge as assumed?
+Unfortunately, this question presents a host of new challenges, not the least of which concerns how to define the features themselves.
 
@@
 
-However, most widely accepted feature sets contain features that are specified at different levels of phonetic detail: some features are more abstract (i.e., general similarities such as +/- syllabic), while others are more concrete (i.e., closer to the articulatory reality, e.g., advanced tongue root).
+Most widely accepted feature sets define features that are specified at different levels of phonetic detail: some features are more abstract (i.e., general similarities such as +/- syllabic), while others are more concrete (i.e., closer to the articulatory reality, e.g., advanced tongue root).
 Some are grounded in pure articulatory or acoustic phonetics, while others involve phonological properties of language.
 And looming in the background of these discussions, the debate of whether distinctive features are innate and universal or rather learned and emergent continues.
-Nevertheless, in general, distinctive feature theories of different theoretical stances propose around two dozen features [@MielkeHume2006], which still has the potential to generate over 16 million distinct feature vectors (2^24).
-Current consensus aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature.
+Thus, the precise limits on what may be included as a feature are underspecified and a matter of some controversy (for more fundamental critiques of features and how they are defined, see @Lindblom1990a).
+Current consensus and critiques aside, there does not appear to be a principled limit to the degree of phonetic specificity that can be encoded by a feature.
 Making finer and finer contrasts may always produce more and more features.
@@
 However, one objection to this account is that humans are subject to physiological limits, both in production and perception.
-In speech production, articulatory gestures are executed with some degree of imprecision [@Ladefoged_etal1972;@Lindblom1990].
+In speech production, articulatory gestures are executed with some degree of imprecision [@Ladefoged_etal1972;@LindblomSundberg1971].
 If this stochastic variability is greater than the difference between two or more phonetic targets, then no reliable distinction can be sustained (cf.
@@
 
-Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond whatever the phonetics-phonology interface may or may not be.
-Thus, we believe the approach presented here has broad applications to the study of other areas of grammar and to the cross-linguistic study of languages.
+<!-- In principle, Lindblom asks whether phonetic alphabets are finite (or non-finite).-->
+
+<!-- Deriving what possible speech sounds are from an ``anthropophonic'' perspective, i.e., to derive them from the ``total sound producing potential'' of the vocal tract also allows us to investigate the evolutionary selection pressures on speech sounds [@Catford1977;@Lindblom1990a].
+-->
+
+Given that all speakers use perceptually contrastive cues (whether signed, spoken or written), how these continuous signals map to discrete elements is a core question of language science that goes beyond what the phonetics-phonology interface may or may not be.
+Thus, we believe the approaches presented here have broad applications to the study of other areas of grammar and to the cross-linguistic study of languages.
+
+<!-- 
+"Anthro"This anthropophonic perspective invites us to look for the constraints that operate in the evolutionary selection of speech sounds."
+
+"without the anthropophonic perspective , phonetics cannot even begin to address issues such as circularity in the definition of speech sounds and the finiteness of phonetic alphabets"
+
+we can mention the circularity and closed-class assumption of DF styles approaches. I think it is a simple elaboration if the discussion where we raise definitional issues with DF.
+
+The anthropophonic stuff could be added as a precedent for emphasizing the closed infinity of the articulatory space. Maybe one sentence or so.
+
+Makes sense from what you say.
+I guess all I could comment at the moment is that it’s striking how sound inventories are organized for (perhaps) capturing the most stable acoustic “spaces” — where the identities of sounds are recoverable irrespective of articulatory “failure”.
+And you seem to have an idea about what gets ignored or can be ignored in service of categorical sound distinctions.
+
+The other point about so-called quantal vowels is that they are non-monotonic, i.e., it's very clear where they start and where they stop -- there's an abrupt perceptual distinction between them and other vowels.
+This is purported to be because you can articulate and [i] or [u] in myriad ways, but the acoustics are relatively stable (again because of the convergence of formants), whereas other non-quantal vowels have to be made much more precise articulatorily.
+
+Nothing besides what I said earlier.
+Could mention the anthropophonemic idea (or whatever it was).
+Circularity of the argument for possible sound from DF approaches.
+Untested assumption of finite inventory of possible sounds.
+
+Could mention the anthrop…. approach when we motivate the notion of infinite phonemes, specifically in conjunction with the infinite articulatory potential but.
+-->
 

### 2023-08-04 paper.tex
submitted version

@@
 \title{How many speech sounds are there in the world's languages?}
-\author{Steven Moran \& Nicholas A. Lester}
-\date{22 June, 2023}
+\author{Steven Moran\textsuperscript{1,2} \& Nicholas A.
+Lester\textsuperscript{3}\\
+\textsuperscript{1}Institute of Biology, University of Neuchâtel,
+Neuchâtel, Switzerland\\
+\textsuperscript{2}Department of Anthropology, University of Miami,
+Coral Gables, FL, USA\\
+\textsuperscript{3}Department of Linguistics, University of North Texas,
+Denton, TX, USA\\
+OrcID, SM, 0000-0002-3969-6549\\
+Corresponding authors:
+\href{mailto:steven.moran@unine.ch}{\nolinkurl{steven.moran@unine.ch}};
+\href{mailto:nicholas.a.lester@gmail.com}{\nolinkurl{nicholas.a.lester@gmail.com}}}
+\date{04 August, 2023}
 
@@
 phonemes are difficult to detect because they are rare, and (b)
-documenting more phonological repertoires cross-linguistically always
-yields more phoneme types. As a sanity check, we apply an unrelated
-estimator of species diversity from biology, Chao-Shen, and show that
-its estimates align with Herdan-Heaps' law. We use both models to
-extrapolate phoneme counts to undocumented languages, resulting in an
-estimate of 4600-4900 distinct phonemes in the full set of languages
-thought to be spoken today (\textasciitilde7-8k), which is roughly 1.5x
-more than what is currently documented in the world's languages
-(\textasciitilde35\%). By extrapolating to hypothetical sample sizes,
-e.g., the number of languages estimated to have been spoken through
-time, and beyond, we find no indication of convergence: infinite
-languages mean infinite phonemes.
+increased documentation of phonological inventories always yields more
+phoneme types. To check this, we apply an unrelated estimator of species
+diversity from biology, Chao-Shen, and show that its estimates also
+align with Herdan-Heaps' law. We use both models to extrapolate phoneme
+counts to undocumented languages, resulting in an estimate of 4600-4900
+distinct phonemes in the full set of languages thought to be spoken
+today (\textasciitilde7-8k), which is roughly 1.5 time more than what is
+currently documented in the world's languages (\textasciitilde35\%). By
+extrapolating to hypothetical sample sizes, e.g., the number of
+languages estimated to have been spoken through time, and beyond, we
+find no indication of convergence: infinite languages mean infinite
+phonemes.
 
@@
 sizes, i.e., languages that existed \emph{through time} (e.g., estimates
-including 30k-500k (\protect\hyperlink{ref-Pagel2000}{Pagel 2000}) or
-64k-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})), the
+including 30-500k (\protect\hyperlink{ref-Pagel2000}{Pagel 2000}) or
+64-140k (\protect\hyperlink{ref-Crystal2002}{Crystal 2002})), the
 Herdan-Heaps' model predicted a continual increase in the number of
-unique phonemes, reaching values as high as 19K. The Chao-Shen model
-produced much more conservative estimates for samples of size 10K and
+unique phonemes, reaching values as high as 19k. The Chao-Shen model
+produced much more conservative estimates for samples of size 10k and
 above, even appearing to settle on an asymptote of roughly 5500 phonemes
-for samples of 100K languages or more. However, when we compare the
+for samples of 100k languages or more. However, when we compare the
 predictions of Herdan-Heaps and Chao-Shen against known type counts in a
@@
 that there is an upper bound on the number of unique phonemes possible
-in spoken languages. Infinite linguistic diversity seems to breed
+in spoken languages. Infinite linguistic diversity seems to bring about
 infinite possible phonemes. But what would it mean if there were indeed
@@
 
-If one adopts a feature-based perspective on the phonetic targets of
-phonemes, then there is an \emph{a priori} upper bound of \(2^F\)
-contrastive sounds, where \(F\) is the number of feature categories and
-2 reflects the choice between presence (+) or absence (-) of that
-feature.\footnote{In some feature sets, some values may be 0, i.e.,
-  undefined, underspecified, or ``for which a segment can be said `not
-  to care''' (\protect\hyperlink{ref-Hayes2009}{Hayes 2009}, pg. 91).
-  For example, speech sounds that are -coronal are underspecified for
-  the features +/-anterior and +/-distributed, which capture coronal
-  distinctions for places of articulation involving (inter-)dentals and
-  alveolars vs palatal alveolars and retroflexes (anterior), and laminal
-  vs apical coronals (distributed).} Clements
+If one adopts a distinctive feature-based perspective on the phonetic
+targets of phonemes, then there is an \emph{a priori} upper bound of
+\(2^F\) contrastive sounds, where \(F\) is the number of feature
+categories and 2 reflects the choice between presence (+) or absence (-)
+of that feature.\footnote{In some distinctive feature sets, some values
+  may be 0, i.e., undefined, underspecified, or ``for which a segment
+  can be said `not to care''' (\protect\hyperlink{ref-Hayes2009}{Hayes
+  2009}, pg. 91). For example, speech sounds that are -coronal are
+  underspecified for the features +/-anterior and +/-distributed, which
+  capture coronal distinctions for places of articulation involving
+  (inter-)dentals and alveolars vs palatal alveolars and retroflexes
+  (anterior), and laminal vs apical coronals (distributed).} Clements
 (\protect\hyperlink{ref-Clements2009}{2009}) coined this calculation
@@
 features, an upper bound on the number of segments under a strictly
-binary formalization is over 137 billion (2\^{}37) feature vectors.
-
-However, most widely accepted feature sets contain features that are
-specified at different levels of phonetic detail: some features are more
-abstract (i.e., general similarities such as +/- syllabic), while others
-are more concrete (i.e., closer to the articulatory reality, e.g.,
-advanced tongue root). Some are grounded in pure articulatory or
-acoustic phonetics, while others involve phonological properties of
-language. And looming in the background of these discussions, the debate
-of whether distinctive features are innate and universal or rather
-learned and emergent continues. Nevertheless, in general, distinctive
-feature theories of different theoretical stances propose around two
-dozen features (\protect\hyperlink{ref-MielkeHume2006}{Mielke \& Hume
-2006}), which still has the potential to generate over 16 million
-distinct feature vectors (2\^{}24). Current consensus aside, there does
-not appear to be a principled limit to the degree of phonetic
-specificity that can be encoded by a feature. Making finer and finer
-contrasts may always produce more and more features. Thus, the exponent
-in the Feature Bounding calculation can itself (by hypothesis) be
-infinite, in line with the predictions of the models presented here, and
-therefore with the notion of infinite possible phonemes. That said,
-contemporary feature sets can capture phonological classes for place and
-manner of articulation, which neatly classify the phonological patterns
-and processes in and across languages (cf. Mielke
+binary formalization is over 137 billion (2\^{}37) feature vectors. In
+general, distinctive feature theories of different theoretical stances
+propose around two dozen features
+(\protect\hyperlink{ref-MielkeHume2006}{Mielke \& Hume 2006}), which
+still has the potential to generate over 16 million distinct feature
+vectors (2\^{}24). Crucially, this theoretical bound on phoneme
+diversity does not in itself preclude the possibility of infinitely many
+phonemes. To get there, one would only need to discover infinitely many
+distinctive features. But, as pointed out by Lindblom
+(\protect\hyperlink{ref-Lindblom1990a}{1990}), most proponents of
+feature-based analysis assume a fixed, small-ish inventory of features.
+The question would then be: how does the feature pool grow empirically
+with the addition of new languages? Does it converge as assumed?
+Unfortunately, this question presents a host of new challenges, not the
+least of which concerns how to define the features themselves.
+
+Most widely accepted feature sets define features that are specified at
+different levels of phonetic detail: some features are more abstract
+(i.e., general similarities such as +/- syllabic), while others are more
+concrete (i.e., closer to the articulatory reality, e.g., advanced
+tongue root). Some are grounded in pure articulatory or acoustic
+phonetics, while others involve phonological properties of language. And
+looming in the background of these discussions, the debate of whether
+distinctive features are innate and universal or rather learned and
+emergent continues. Thus, the precise limits on what may be included as
+a feature are underspecified and a matter of some controversy (for more
+fundamental critiques of features and how they are defined, see Lindblom
+(\protect\hyperlink{ref-Lindblom1990a}{1990})). Current consensus and
+critiques aside, there does not appear to be a principled limit to the
+degree of phonetic specificity that can be encoded by a feature. Making
+finer and finer contrasts may always produce more and more features.
+Thus, the exponent in the Feature Bounding calculation can itself (by
+hypothesis) be infinite, in line with the predictions of the models
+presented here, and therefore with the notion of infinite possible
+phonemes. That said, contemporary feature sets can capture phonological
+classes for place and manner of articulation, which neatly classify the
+phonological patterns and processes in and across languages (cf. Mielke
 (\protect\hyperlink{ref-Mielke2008}{2008})).
@@
 (\protect\hyperlink{ref-Ladefoged_etal1972}{Ladefoged et al. 1972};
-\protect\hyperlink{ref-Lindblom1990}{Lindblom 1990}). If this stochastic
-variability is greater than the difference between two or more phonetic
-targets, then no reliable distinction can be sustained (cf. Stevens
-(\protect\hyperlink{ref-Stevens1972}{1972})). In comparison, if the
-range of articulatory gestures is greater than their acoustic (or
-perceptual) reception, there is a ``many-to-one'' mapping between
-articulation and perception. In perception, the ability to discriminate
-acoustic stimuli is also limited both anatomically and cognitively.
-Differences in the signal that fall below the various thresholds for
-discrimination are thus ignored.
+\protect\hyperlink{ref-LindblomSundberg1971}{Lindblom \& Sundberg
+1971}). If this stochastic variability is greater than the difference
+between two or more phonetic targets, then no reliable distinction can
+be sustained (cf. Stevens (\protect\hyperlink{ref-Stevens1972}{1972})).
+In comparison, if the range of articulatory gestures is greater than
+their acoustic (or perceptual) reception, there is a ``many-to-one''
+mapping between articulation and perception. In perception, the ability
+to discriminate acoustic stimuli is also limited both anatomically and
+cognitively. Differences in the signal that fall below the various
+thresholds for discrimination are thus ignored.
 
@@
 signed, spoken or written), how these continuous signals map to discrete
-elements is a core question of language science that goes beyond
-whatever the phonetics-phonology interface may or may not be. Thus, we
-believe the approach presented here has broad applications to the study
-of other areas of grammar and to the cross-linguistic study of
-languages.
+elements is a core question of language science that goes beyond what
+the phonetics-phonology interface may or may not be. Thus, we believe
+the approaches presented here have broad applications to the study of
+other areas of grammar and to the cross-linguistic study of languages.
 
@@
 for approximate text retrieval. \emph{Journal of the American Society
-for Information Science}. Wiley Online Library 51(1). 69--82.
+for Information Science} 51(1). 69--82.
 
@@
 Chao, Anne \& Chun-Huo Chiu. 2016. Species richness: Estimation and
-comparison. In \emph{Wiley StatsRef: Statistics reference online},
-1--26. John Wiley \& Son.
+comparison. In \emph{{Wiley StatsRef: Statistics Reference Online}},
+1--26. John Wiley \& Sons.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Egghe2007}{}}%
-Egghe, Leo. 2007. Untangling herdan's law and heaps' law: Mathematical
-and informetric arguments. \emph{Journal of the American Society for
-Information Science and Technology}. Wiley Online Library 58(5).
-702--709.
+Egghe, Leo. 2007. Untangling {Herdan's} law and {Heaps'} law:
+Mathematical and informetric arguments. \emph{Journal of the American
+Society for Information Science and Technology} 58(5). 702--709.
 
@@
 universals: Language diversity and its importance for cognitive science.
-\emph{Behavioral and Brain Sciences}. Cambridge University Press 32(5).
-429--448.
+\emph{Behavioral and Brain Sciences} 32(5). 429--448.
 
@@
 across the world's languages: A quantitative analysis of phonetically
-transcribed word lists. \emph{Language Sciences}. Elsevier 69. 125--135.
+transcribed word lists. \emph{Language Sciences} 69. 125--135.
 
 \leavevmode\vadjust pre{\hypertarget{ref-FirchowFirchow1969}{}}%
-Firchow, Irwin \& Jacqueline Firchow. 1969. {An Abbreviated Phoneme
-Inventory}. \emph{Anthropological Linguistics} 11. 271--276.
+Firchow, Irwin \& Jacqueline Firchow. 1969. An abbreviated phoneme
+inventory. \emph{Anthropological Linguistics} 11. 271--276.
 
@@
 Good, Jeff \& Michael Cysouw. 2013. Languoid, doculect, and glossonym:
-Formalizing the notion'language'. \emph{Language Documentation and
+Formalizing the notion {``language.''} \emph{Language Documentation \&
 Conservation} 7. 331--359.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Herdan1964}{}}%
-Herdan, Gustav. 1964. Quantitative linguistics. Butterworth.
+Herdan, Gustav. 1964. \emph{Quantitative linguistics}. London:
+Butterworth.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Joos1948}{}}%
-Joos, Martin. 1948. Acoustic phonetics. \emph{Language}. Linguistic
-Society of America 24(2). 5--136. (7 December, 2022).
+Joos, Martin. 1948. Acoustic phonetics. \emph{Language} 24(2). 5--136.
 
@@
 auditory-motor theory of speech production. \emph{{UCLA} working papers
-in phonetics}. UCLA Los Angeles, CA 22(48). 48--76.
+in phonetics} 22(48). 48--76.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lehfeldt1975}{}}%
-Lehfeldt, W. 1975. {Die Verteilung der Phonemanzahl in den nat{ü}rlichen
-Sprachen}. \emph{Phonetica} 31. 247--287.
+Lehfeldt, Werner. 1975. {Die Verteilung der Phonemanzahl in den
+nat{ü}rlichen Sprachen}. \emph{Phonetica} 31. 247--287.
 
@@
 Leijenhorst, Dick C van \& Th P Van der Weide. 2005. A formal derivation
-of heaps' law. \emph{Information Sciences}. Elsevier 170(2-4). 263--272.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lindblom1990}{}}%
-Lindblom, Björn. 1990. Explaining phonetic variation: A sketch of the
-h\&h theory. In W. J. Hardcastle \& A. Marchal (eds.), \emph{Speech
-production and speech modelling}, 403--439. Springer.
-https://doi.org/\url{https://doi.org/10.1007/978-94-009-2037-8_16}.
+of {Heaps'} law. \emph{Information Sciences} 170(2-4). 263--272.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lindblom1990a}{}}%
+Lindblom, Björn. 1990. On the notion of {``possible speech sound.''}
+\emph{Journal of Phonetics} 18(2). 135--152.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LindblomSundberg1971}{}}%
+Lindblom, Björn \& Johan Sundberg. 1971. Acoustical consequences of lip,
+tongue, jaw, and larynx movement. \emph{The Journal of the Acoustical
+Society of America} 50(4B). 1166--1179.
 
@@
 Macklin-Cordes, Jayden L \& Erich R Round. 2020. Re-evaluating phoneme
-frequencies. \emph{Frontiers in Psychology}. Frontiers Media SA 11.
-570895.
+frequencies. \emph{Frontiers in Psychology} 11. 570895.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Martindale_etal1996}{}}%
-Martindale, Colin, S. M. Gusein‐Zade, Dean McKenzie \& Mark Yu.
+Martindale, Colin, S. M. Gusein-Zade, Dean McKenzie \& Mark Yu.
 Borodovsky. 1996. Comparison of equations describing the ranked
 frequency distributions of graphemes and phonemes. \emph{Journal of
-Quantitative Linguistics}. Routledge 3(2). 106--112.
+Quantitative Linguistics} 3(2). 106--112.
 
@@
 Mielke, Jeff \& Elizabeth Hume. 2006. Distinctive features. In
-\emph{Encylopedia of langauge}. Elsevier.
+\emph{{Encylopedia of Langauge}}. Elsevier.
 
@@
 Milička, Jiři. 2009. Type-token \& hapax-token relation: A combinatorial
-model. \emph{Glottotheory}. De Gruyter Akademie Forschung 2(1). 99--110.
+model. \emph{Glottotheory} 2(1). 99--110.
 
@@
 language: A critical review and future directions. \emph{Psychonomic
-bulletin \& review}. Springer 21(5). 1112--1130.
+Bulletin \& Review} 21(5). 1112--1130.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Potter1945}{}}%
-Potter, Ralph K. 1945. Visible patterns of sound. \emph{Science}.
-American Association for the Advancement of Science 102(2654). 463--470.
+Potter, Ralph K. 1945. Visible patterns of sound. \emph{Science}
+102(2654). 463--470.
 
@@
 Snyman, J. W. 1975. \emph{{Zu\textbar'hoasi Fonologie en Woordeboek}}.
-Cape Town: Balkema.
+Cape Town: A. A. Balkema.
 
@@
 articulatory-acoustic data. In Edward E David \& Peter B Denes (eds.),
-\emph{Human communication: A unified view}. McGraw-Hill.
+\emph{Human communication: A unified view}. New York: McGraw-Hill.
 
@@
 Tambovtsev, Yuri \& Colin Martindale. 2007. Phoneme frequencies follow a
-yule distribution. \emph{SKASE Journal of Theoretical Linguistics}.
-Citeseer 4(2). 1--11.
+{Yule} distribution. \emph{SKASE Journal of Theoretical Linguistics}
+4(2). 1--11.
 

### 2024-07-20 SI2/phoible_randomization_v2.Rmd
adds second SI

---
title: "Random baselines for PHOIBLE"
author: "Nicholas A. Lester"
date: "2024-01-03"
output:
  github_document:
      toc: true
  html_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=F, warnings=F, message=F, cache=T)
```

```{r, include=FALSE}
rm(list=ls(all=T))
```

```{r, include=F}
# Custom library for the SI
source("./SI_functions.R")
```

# 1. Introduction

The supplemental analyses presented below explore several factors that might impact the results reported in the main text. The following questions are addressed:

1. Is the statistical information encoded in PHOIBLE compatible with randomly constructed sets of inventories? ([Section 2](#randbase))

  - If so, we can assume that the particularities of PHOIBLE (which we take to reflect the particularities of human phonological systems) are in fact irrelevant to discovering a pattern consistent with Herdan-Heaps' law.

2. Do the results reported in the main text hold when we train and evaluate the Herdan-Heaps' models on different subsets of the data? ([Section 3](#overfit))

  - If so, then we can assume that the findings reported in the main text are robust (not the result of overfitting), and that PHOIBLE shows a high degree of internal statistical/distributional consistency.

3. Do general differences in the phonological inventories of languages spoken by (very) large and (very) small populations pose a confound for estimating the Herdan-Heaps' curve? ([Section 4](#popmodels))

  - If so, then (a) prediction accuracy from the high/low population group to the other should decrease relative to models that ignore population, and (b) subgroups could differ in whether or not Herdan-Heaps is a good fit. Such findings would undermine the representativeness of the results reported in the main text.

---

# 2. Question 1: Random baselines and ablation of confounding information {#randbase}

We propose four interrelated algorithms designed to random baselines for PHOIBLE. These baselines are generated by ablating different sources of information through structured randomization and resampling. The two sources of information considered here are *inventory size* and *the frequency distribution of phoneme types*.

By removing different kinds of information, we test their contribution to the observed fit of Herdan-Heaps' law to PHOIBLE. If the random samples behave differently from PHOIBLE, then the corresponding (ablated) sources of information can be interpreted as relevant contributors to the Herdan-Heaps' fit.

We proceed as follows:

1. In [Section 2.1](#info), we describe two sources of information that may confound our interpretation of the findings reported in the main text.

2. In [Section 2.2](#data), we load, process, and explore the PHOIBLE data.

3. In [Section 2.3](#alg1), we remove information about the **distribution of phoneme inventory sizes** while keeping information about the relative frequency of sounds.

4. In [Section 2.4](#alg2), we do the opposite: remove information about the **relative frequency of sounds** while keeping information about the distribution of inventory sizes.

5. In [Section 2.5](#alg3), we **ablate both sources of information**: neither inventory size nor relative frequency of sounds is allowed to contribute.

6. In [Section 2.6](#alg4), we **allow both sources of information** during the construction of random inventories.

7. In [Section 2.7](#disc1), we discuss the results of the randomization and ablation studies.

8. In [Section 2.8](#conc1), we offer preliminary conclusions about the internal statistical consistency of PHOIBLE.

Note that all of these randomization techniques remove the internal structure of inventories (for example, regularities in the co-occurrence of certain sounds, how vowels are balanced against consonants, etc.)

## 2.1 Information used to construct random baselines {#info}

## 2.1.1 Consonant inventory size

Consonant inventory size is known to correlate with appearance of rare/novel phonemes. That is, most languages share a stock of very common consonants. Growing beyond that set requires selection from increasingly less common sounds. Thus, knowledge of the inventory size of a language is to some extent knowledge of how many new sounds that language will contribute to a cumulative sample, even when many languages have already been observed. In the extreme case, the observed Herdan-Heaps'-like pattern could derive from the distribution of inventory sizes.

We can also consider this inventory size in relation to the original application of Herdan-Heaps' law to lexical discovery in documents. Generally, Herdan-Heaps' law depends only on the frequencies of the measured units (usually words, or in our case, phonemes). Steps through the document are made uniformly (e.g., one word at a time). In the present study, however, we step through the full document (PHOIBLE) one inventory at a time. The purpose: to mirror the real-life process of accruing data through phonological documentation and discovery. Because of this methodological change, we test to what extent our treating full inventories as steps might support the observed Herdan-Heaps'-like pattern of PHOIBLE.

## 2.1.2 Relative cross-linguistic phoneme frequency

Another source of information contained in PHOIBLE is the relative frequency of each sound across languages. As mentioned above, this information should be critical as it underlies the link between types and tokens -- the same relationship that is explicitly encoded in Herdan-Heaps' law (as well as Zipf's law), with the embellishment of two free parameters.

# 2.2 Preliminaries {#data}

Read in PHOIBLE and extract some basic information to be used in the randomization procedures.

```{r}
# Read in PHOIBLE
phoible_long <- read_csv("./phoible.csv", 
                    col_types = c(InventoryID = "i", 
                                  Marginal = "l", 
                                  .default = "c")) %>%

          # Select a random inventory per language 
          rand.inv() %>%
          # Extract only the necessary columns
          select(InventoryID, Phoneme, ISO6393)

# Randomize
set.seed(0)
invs = unique(phoible_long$InventoryID) %>% sample()
phoible_long = phoible_long %>% arrange(match(InventoryID, invs))

```

## 2.2.1 Find the inventory sizes in PHOIBLE

```{r}
inv_sizes = phoible_long %>%
            group_by(InventoryID) %>%
            summarize(inv_size = n()) %>%
            pull(inv_size)
```

## 2.2.2 Plot the shape of the inventory size distributions

```{r}
ggplot(as.data.frame(inv_sizes), aes(x=inv_sizes)) +
  geom_density(alpha=0.5, fill="dodgerblue") +
  theme_bw() +
  xlab("Inventory size")
```

## 2.2.3 Plot the log-log rank/frequency relationship

```{r}
freqs <- sort(table(inv_sizes), decreasing=T)
names(freqs) <- 1:length(freqs)
freq.df = as.data.frame(freqs)
colnames(freq.df)[1] <- "Rank"
freq.df$`Rank (log)` <- log(as.numeric(freq.df$Rank))
freq.df$`Freq (log)` <- log(as.numeric(freq.df$Freq))

ggplot(freq.df, aes(x=`Rank (log)`, y=`Freq (log)`)) +
  geom_line() +
  theme_bw()
```

This plot shows the log-transformed frequency rank (1 = most frequent; x-axis) vs. the log-transformed frequency (y-axis) of each inventory size in PHOIBLE. The plot suggests that inventory sizes are not Zipf-distributed. Instead, most languages have very similar sizes.

## 2.2.4 Extract the set of phonemes.

```{r}
phon_typs <- unique(phoible_long$Phoneme)
```

## 2.2.5 Extract the number of distinct inventories.

```{r}
pho.inv.num <- length(unique(phoible_long$InventoryID))
```

## 2.2.6 Processing PHOIBLE

### 2.2.6.1 Cumulative counts

```{r}
pho.counts <- get_cumulative_segment_counts_df(phoible_long %>% mutate(Iteration=1))
```

### 2.2.6.2 Model the PHOIBLE frequencies

```{r}
m.pho.1 = lm(log(results) ~ log(InventoryID), pho.counts)
```

Generate model predictions
```{r}
pho.1.preds = heaps.preds(pho.counts, m.pho.1)
```

### 2.2.6.3 Plot the results

```{r}
plot.emp.vs.pred(pho.counts, pho.1.preds, pho.counts) + guides(color="none")
```

Cumulative type count is given on the y-axis. The x-axis gives the order of addition of inventories. The dotted line represents the model fit; the blue line represents the empirical growth curve based on a randomly shuffled version of PHOIBLE.

As in the analysis reported in the main text, the Herdan-Heaps' model predicts the empirical values quite well.

We now turn to the first of the randomization strategies.

# 2.3 Algorithm 1: Remove inventory size information but retain relative frequencies {#alg1}

- For *n* desired databases,

  - For *k* desired inventories,

    - define inventory size *m* as the mode size in PHOIBLE

    - randomly select *m* phonemes without replacement from PHOIBLE according to their crosslinguistic probabilities

## 2.3.1 Generate 10 different versions of the random database using Alg. 1

```{r}
alg.1.10 <- processing.iterator(phoible_long, 10, pho.inv.num, inv.inf = F, freq.inf=T)
```

Inspect the results
```{r}
summary(alg.1.10)
```

## 2.3.2 Compute the hypothetical Heaps' estimates based on the constructed databases

Create the Herdan-Heaps' model object using the randomized data

```{r}
m.alg.1 = lmer(log(results) ~ log(InventoryID) + (1|Iteration), alg.1.10)
```

Use the model to make predictions

```{r}
alg.1.10.preds = heaps.preds(alg.1.10, m.alg.1)
```

## 2.3.3 Graphing {#phoplot}

"Empirical" vs. modeled estimates (how good a fit is Heaps'?)

```{r}
plot.emp.vs.pred(alg.1.10, alg.1.10.preds, pho.counts) + guides(color="none")
```

The order of addition of inventories is given on the x-axis. The estimated number of total unique segment types is given on the y-axis. The dotted black line corresponds to the regression curve from the Herdan-Heaps' model. Smooth colored lines correspond to the predictions of the model per random shuffling iteration. The blue jagged line gives the true growth curve for PHOIBLE.

Removing information about the distribution of inventory size types does not have a strong impact on the quality of fit of the Herdan-Heaps' equation. However, both the model and the observed number of types in the random sample underestimate the true number of types.

# 2.4 Algorithm 2: Preserve inventory size information but ablate relative phoneme frequency {#alg2}

- For *n* desired databases,

  - For *k* desired inventories,

    - randomly select an inventory size *m* according to its crosslinguistic probability

    - randomly select *m* phonemes without replacement from the set of sounds in PHOIBLE (uniform distribution)

## 2.4.1 Generating 10 databases with the same number of inventories as PHOIBLE using Alg. 2

```{r}
alg.2.10 = processing.iterator(phoible_long, 10, pho.inv.num, inv.inf = T, freq.inf = F)
```

Inspect the results

```{r}
head(alg.2.10)
```

## 2.4.2 Compute the hypothetical Heaps' estimates based on the constructed databases

Create the Herdan-Heaps' model object using the randomized data

```{r}
m.alg.2 = lmer(log(results) ~ log(InventoryID) + (1|Iteration), alg.2.10)
```

Use the model to make predictions

```{r}
alg.2.10.preds = heaps.preds(alg.2.10, m.alg.2)
```

## 2.4.3 Graphing

```{r}
plot.emp.vs.pred(alg.2.10, alg.2.10.preds, pho.counts)  + guides(color="none")
```

Plotting conventions are the same as those reported in [Section 2.3.3](#phoplot).

Removing information about relative frequency (by sampling according to a uniform distribution across segments) drastically changes the shape of the growth curve. In other words, when all sounds are equally easy to find, we need many fewer languages to converge on the true number of segments. Notably, Herdan-Heaps' is a poor fit for the randomized data. Thus, as expected, the relative frequencies of sounds in PHOIBLE contribute heavily to the Herdan-Heapsian growth pattern reported in the main text.

# 2.5 Algorithm 3: Ablate both inventory size and relative frequency {#alg3}

- For *n* desired databases,

  - For *k* desired inventories,

    - define inventory size *m* as the mode size in PHOIBLE

    - randomly select *m* phonemes without replacement from the set of sounds in PHOIBLE (uniform distribution)

## 2.5.1 Generating 10 databases with the same number of inventories as PHOIBLE using Alg. 2

```{r}
alg.3.10 = processing.iterator(phoible_long, 10, pho.inv.num, inv.inf = F, freq.inf = F)
```

Inspect the results

```{r}
head(alg.3.10)
```

## 2.5.2 Compute the hypothetical Heaps' estimates based on the constructed databases

Create the Herdan-Heaps' model object using the randomized data

```{r}
m.alg.3 = lmer(log(results) ~ log(InventoryID) + (1|Iteration), alg.3.10)
```

Use the model to make predictions

```{r}
alg.3.10.preds = heaps.preds(alg.3.10, m.alg.3)
```

## 2.5.3 Graphing

```{r}
plot.emp.vs.pred(alg.3.10, alg.3.10.preds, pho.counts) + guides(color="none")
```

Plotting conventions are the same as those reported in [Section 2.3.3](#phoplot).

Neither preserving (Algorithm 2) nor removing (Algorithm 3) information about inventory sizes alters the effect of removing relative frequencies.

# 2.6 Algorithm 4: Preserving both inventory size and relative frequency {#alg4}

This algorithm preserves information about the frequency distributions of individual phonemes and inventory sizes as encoded in PHOIBLE. In other words, it aims to create a "random pseudo-clone" of the true database.

- For *n* desired databases,

  - For *k* desired inventories,

    - randomly select an inventory size *m* according to its crosslinguistic probability

    - randomly select *m* phonemes without replacement from PHOIBLE according to their crosslinguistic probabilities

## 2.6.1 Generating 10 databases with the same number of inventories as PHOIBLE using Alg. 2

```{r}
alg.4.10 = processing.iterator(phoible_long, 10, pho.inv.num, inv.inf = T, freq.inf = T)
```

Inspect the results

```{r}
head(alg.4.10)
```

## 2.6.2 Compute the hypothetical Heaps' estimates based on the constructed databases

Create the Herdan-Heaps' model object using the randomized data

```{r}
m.alg.4 = lmer(log(results) ~ log(InventoryID) + (1|Iteration), alg.4.10)
```

Use the model to make predictions

```{r}
alg.4.10.preds = heaps.preds(alg.4.10, m.alg.4)
```

## 2.6.3 Graphing

```{r}
plot.emp.vs.pred(alg.4.10, alg.4.10.preds, pho.counts) + guides(color="none")
```

Plotting conventions are the same as those reported in [Section 2.3.3](#phoplot).

As with Algorithm 1, preserving information about relative frequencies leads to a much closer fit between the random databases, the expectations based on Herdan-Heaps' law, and the true database. Notably, in contrast to Algorithm 1, the random estimates show a much smaller underestimation bias. Thus, adding information about inventory sizes **does** improve estimates when relative frequencies are preserved, but not when the latter information is ablated.

## 2.6.4 Increasing the number of languages per iteration

The main text presents several extrapolation analyses. Here, we generate random data approximately consistent with the number of known living languages using Algorithm 4 (our statistical "pseudo-clone" of PHOIBLE). We then apply the model trained on ~2000 languages to the expanded random database.

7K languages per iteration

```{r}
alg.4.7000 <- processing.iterator(phoible_long, 10, 7000, inv.inf = T, freq.inf = T)
```

Use the model to make predictions

```{r}
alg.4.7000.preds = heaps.preds(alg.4.7000, m.alg.4)
```

Graphing

```{r}
plot.emp.vs.pred(alg.4.7000, alg.4.7000.preds, pho.counts) + guides(color="none")
```

Plotting conventions are the same as those reported in [Section 2.3.3](#phoplot).

Extrapolating the number of languages in the random sample to somewhere near the number of known, living languages, we see a continued decrease in the slope of the randomly generated values relative to both the model predictions and true PHOIBLE data.

# 2.7 Discussion {#disc1}

We established several random versions of the PHOIBLE database. Randomization was based around two sources of information: inventory size distributions and relative frequency distributions.

Results showed that removing relative frequency information warped the type count growth curve most dramatically, leading to near immediate convergence on the true type with no correspondence to the expectations of Herdan-Heaps' law.

Inventory size distributions were not particularly relevant, though including this information supported a closer approximation to the true growth curve when information of relative segment frequencies was preserved. Despite this improvement, the random data tend towards underestimation relative to both the model expectations and the true behavior of PHOIBLE (as sample size increases).

There are certainly other sources of information that could be explored along similar lines. For example, consonants are roughly twice as frequent as vowels in PHOIBLE, with tones making up only two percent overall. Relative frequency distributions within and across these classes, along with standards for the combination of classes within single inventories, could influence how easy it is to discover rare sounds. We leave these questions to future research.

# 2.8 Conclusion {#conc1}

We successfully removed the correlation between Herdan-Heaps' law and the data in PHOIBLE. There appears to be something critical about the statistical properties of PHOIBLE that suggests infinite growth in sound types.

Inclusion of more and more refined, nested approaches to random sampling could yield increasingly better fits. However, we stress that including more and more information from the source dataset should generally be expected to yield closer approximations. The real question is how these sources of information interact to yield the statistical properties of PHOIBLE.

# 3. Question 2: Testing for overfitting {#overfit}

In this section, we address a potential concern related to overfitting. By training and testing model parameters on the same dataset, we risk overfitting. In this case, overfitting can be interpreted as an artificially tight overlap between Herdan-Heaps' predictions and the training data. Perhaps HH models based on different subsets of languages in the total sample are more or less predictive of other subsets.

We address this concern in two ways:

1. Train an HH model on 75% of PHOIBLE and compare its predictions to the remaining 25%

  - Iterate the train/test cycle over different groups of testing and training languages, randomly sampled

2. Compare the behavior of subgroups of languages defined by their population (encoded as endangerment levels)

  - Look at within-group patterns of growth and compare to HH expectations

  - Predict the behavior of more endangered languages using a model trained on less endangered languages

## 3.1 Modeling and prediction

Using 10-fold cross-validation with 75% training on 25% testing data from PHOIBLE.

```{r}
x.pred.df = cross.pred(phoible_long, 10)
```

## 3.2 Plotting predicted values against empirical values

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=x.pred.df)

cross.pred.plot(x.pred.df, mod, title = "Cross-validation of Herdan-Heaps' estimates in PHOIBLE\n75% training, 25% testing")
```

The x-axis gives the empirical (observed) values within the test set. The y-axis gives the predictions of the model based on the training set. The regression curve of predicted vs. empirical is given in black. The 95% confidence interval of that regression is indicated by dotted lines and blue shading. Observed relations are given in light grey (grouped by iteration). The hypothetical identity relation between empirical and predicted values is plotted in red.

$\beta$ refers to the coefficient relating Empirical to Predicted values. $p$ refers to the significance level of the relation.

## 3.3 Plotting predicted values per inventory

```{r}
cross.pred.plot(x.pred.df, plot.typ="heaps.pred", title = "Predicted values per inventory vs. empirical values") + 
  # Add a line for true PHOIBLE counts based on the entire database
  geom_line(data=pho.1.preds %>% filter(InventoryID <= 755), aes(x=InventoryID, y = results), color="red", lwd=2) +
  xlim(0, max(x.pred.df$InventoryID))
```

The x-axis shows the order of accumulation of inventories. The y-axis shows the predicted/empirical type count. Predicted values are given in solid lines (colors reflect different iterations). Empirical values are given in dashed lines. The thick red line indicates predictions from the full-PHOIBLE model.

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=x.pred.df)

cross.pred.plot(x.pred.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the test data") 
```

The x-axis gives the order of accumulation of inventories within the test set. The y-axis gives the predicted type counts based on the test set alone. The Herdan-Heaps' curve is given in black. The 95% confidence interval of that curve is indicated by dotted lines and blue shading. Observed relations are given in light grey (grouped by iteration).

# 3.4 Discussion

Considering all inventories in PHOIBLE, cross-prediction from 75% to 25% of the sample using Herdan-Heaps' models yields results consistent with those reported in the main text. Predictions of the model and empirical values correspond at a rate of .97 (near-identity).

Therefore, (a) PHOIBLE exhibits a high degree of internal statistical consistency, and (b) the results of the analysis reported in the main text are not due to overfitting by training and testing on the same data.

# 4. Q3: Population-based sampling {#popmodels}

Population size can be treated as a proxy for difficulty of access and/or documentability. In other words, the smaller the population, the more difficult it is to find and document the language. Hence, languages with fewer speakers should be accrued later in the documentary history of linguistics rather than earlier (on average).

Moreover, languages with fewer speakers may behave differently from languages with many speakers. Perhaps cumulative growth models based primarily on more populous languages are not good fits for less populous languages.

We address these issues by analyzing a subset of the PHOIBLE languages for which we have numerical and categorical measures of population size.

## 4.1 Measuring population size

We measure population size using two sources. For smaller languages, we use the language endangerment labels and population estimates from the Endangered Language Project (ELP; Catalog of Endangered Languages, 2024). For large languages, we use the estimates from Ethnologue (Ethnologue: Languages of the world, 2023) as provided freely through Wikipedia (accessed 30. Jun, 2024).

### 4.1.1 ELP data

First, import the ELP data.

```{r}
eth.dat = read.table("./elp_dat.txt", sep="\t", header=T, na.strings=c("","NA")) 
```

Some clean-up and simplification of endangerment status labels and population size estimates

```{r}
# Simplify the labels for endangerment status in ELP
simp.label = gsub("^At .+", "at_risk", eth.dat$Status)
simp.label = gsub("^C.+", "critically_endangered", simp.label)
simp.label = gsub("^D.+", "dormant", simp.label)
simp.label = gsub("^Aw.+", "awakening", simp.label)
simp.label = gsub("^E.+", "endangered", simp.label)
simp.label = gsub("^S.+", "severely_endangered", simp.label)
simp.label = gsub("^T.+", "threatened", simp.label)
simp.label = gsub("^V.+", "vulnerable", simp.label)

# Factorize the new, simplified labels
eth.dat$simp.status = as.factor(simp.label)

# Remove non-numeric characters
eth.dat$Pop.Size.Clean = as.numeric(gsub("[^0-9\\-]+", "", eth.dat$Pop.Size))

# Remove unnecessary info
eth.dat = eth.dat %>%
          select(ISO6393, Pop.Size.Clean, simp.status) %>%
          filter(Pop.Size.Clean > 0)

# Save the cleaned version
write.table(file="./elp_iso_pop.txt", eth.dat, sep="\t", quote=F, row.names=F)
```

### 4.1.2 Ethnologue/Wiki data

Load Ethonologue (via Wiki) data

```{r}
wiki.pops = read.table("./ethnologue_wiki_dat.txt", sep="\t", header=T) %>%
  select(ISO6393, Pop.Size.Clean) %>%
  mutate(simp.status = "robust") %>%
  filter(!is.na(ISO6393))
```

### 4.1.3 Finalize the population table

Join ELP and Ethnologue counts

```{r}
pop.dat = bind_rows(eth.dat, wiki.pops)
```

Merge the population data with PHOIBLE

```{r}
phoible_with_pop = phoible_long %>%
                   left_join(., pop.dat, by="ISO6393") %>%
                   filter(!is.na(Pop.Size.Clean)) %>% 
                   mutate(simp.status = reorder(simp.status, Pop.Size.Clean, mean))

# New sample size (number of inventories)
length(unique(phoible_with_pop$ISO6393))
```

Find a reasonable split point (here, second quartile)

```{r}
quantile(phoible_with_pop$Pop.Size.Clean, na.rm=T)
```

### 4.1.4 Exploratory plotting of population sizes

Explore the distribution of populations across status (in the log space)

```{r}
ggplot(phoible_with_pop %>% filter(!is.na(simp.status)) %>% select(simp.status, Pop.Size.Clean, InventoryID) %>% distinct(), aes(x = simp.status, y = log(Pop.Size.Clean), fill = simp.status)) +
  geom_violin() + 
  theme_bw() +
  ylab("Population (log)") +
  theme(axis.text.x = element_text(angle=45, vjust = 1, hjust =1)) + 
  guides(fill = "none")
```

The ELP data present a clear log-linear increase in means, with some overlap in the actual populations per category. The Ethnologue data fall a little above expectation, but more or less adhere to the log-linear trend found in the ELP data. These data thus represent a fairly consistent step-wise increase in populations.

Next, we turn to the relative frequencies of each status class.

```{r}
ggplot(phoible_with_pop %>% filter(!is.na(simp.status)) %>% select(simp.status, Pop.Size.Clean, InventoryID) %>% distinct(), aes(x=simp.status, fill=simp.status)) +
  geom_bar() +
  theme_bw() + 
  theme(axis.text.x = element_text(angle=45, vjust = 1, hjust=1)) + 
  guides(fill = "none")
```

The population categories have very different frequencies. However, the distribution of frequencies resembles the Gaussian, with a strong peak in number of observations around the central category ("threatened"), attended by decreasing tails in either direction  (increasing: "vulnerable" > "at risk" $\approx$ "robust"; decreasing: "endangered" > "severely endangered") populations. The most prominent deviation from this pattern is the category reserved for the smallest populations, "critically endangered," which stands as the third-most frequent in terms of number of languages while sitting at the lower extreme of the population distribution.

In terms of cross-prediction tasks (e.g., predicting from "robust" to all other categories), this lack of balance could create artificially poor fits.

## 4.2 Modeling

We model the data several ways:

- predicting from 'robust' to other languages ([Section 4.2.1](#rvall))

- predicting from languages from top 25% of population sizes to other languages ([Section 4.2.2](#25v75)

- predicting from languages from the top 50% of population sizes to other languages ([Section 4.2.3](#50v50))

- predicting from languages from the top 75% of population sizes to other languages ([Section 4.2.4](#75v25))

- repeating the above analyses, but downsampling all endangerment categories to include only as many languages as the smallest category ([Sections 4.2.5](#rvallds) - 4.2.8)

- repeating the downsampled analysis, but removing the smallest category ("at_risk") to increase overall sample size per category (Sections [4.2.9](#rvallds_norisk) - 4.2.12)

- walking through the selection process in a tiered manner (class by class according to descending population size), using random ordering within class ([Section 4.2.14](#classXclass))

  - the goal is to mimic the hypothetical trend that languages spoken by fewer people will be encountered later than languages spoken by more people

---

All remaining analyses follow the same basic structure:

- Train a model on a subsection of the data

- Use the model to predict growth curves for the remaining data

- Iterate this process 10 times with random ordering within group

- Statistically analyze the goodness-of-fit of the model to the held-out data

- Visualize the results

Summaries of the results are given at the end of each subsection

### 4.2.1 Category- and population-based cross-prediction

#### 4.2.1.1 Robust vs. other lanugages {#rvall}

- Training data are taken from the "robust" category.

- Test data are taken from the remaining endangerment categories.

---

```{r}
rob.df = pop.cross.pred(phoible_with_pop, 10, "robust")
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=rob.df)

cross.pred.plot(rob.df, mod, title="Predicting from 'robust' to less populous languages\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(rob.df, plot.typ="heaps.pred", title = "Predicting from 'robust' to less populous languages\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=rob.df)

cross.pred.plot(rob.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for non-'robust' languages")
```

#### 4.2.1.2 Top 25% vs. bottom 75% {#25v75}

- Training data are taken from the top 25% most populous languages.

- Test data are taken from the remaining 75% of languages.

---

```{r}
top.25.df = pop.cross.pred(phoible_with_pop, 10, pop.cut=4)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.25.df)

cross.pred.plot(top.25.df, mod, title = "Predicting from the 25% most populous to the 75% least populous\nEmpirical vs. predicted values")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.25.df, plot.typ="heaps.pred", title="Predicting from the 25% most populous to the 75% least populous languages\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.25.df)

cross.pred.plot(top.25.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 75% least populous languages")
```

### 4.2.1.3 Top 50% vs. bottom 50% {#50v50}

- Training data are taken from the top 50% most populous languages.

- Test data are taken from the remaining 50% of languages.

---

```{r}
top.50.df = pop.cross.pred(phoible_with_pop, 10, pop.cut=3)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.50.df)

cross.pred.plot(top.50.df, mod, title = "Predicting from 50% most populous to 50% least populous languages\nEmpirical vs. predicted values")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.50.df, plot.typ="heaps.pred", title="Predicting from 50% most populous to 50% least populous languages\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.50.df)

cross.pred.plot(top.50.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for 75% least populous languages")
```

### 4.2.1.4 Top 75% vs. bottom 25% {#75v25}

- Training data are taken from the top 75% most populous languages.

- Test data are taken from the remaining 25% of languages.

---

```{r}
top.75.df = pop.cross.pred(phoible_with_pop, 10, pop.cut=3)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.75.df)

cross.pred.plot(top.75.df, mod, title = "Predicting from the 75% most populous to the 25% least populous languages\nEmpirical vs. predicted values")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.75.df, plot.typ="heaps.pred", title="Predicting from the 75% most populous to the 25% least populous languages\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.75.df)

cross.pred.plot(top.75.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 25% least populous languages")
```

#### 4.2.1.5 Interim Discussion

Less populous languages indeed behave differently from more populous languages, no matter how we slice the dataset. Models trained on more populous languages tend to overpredict the growth rate segment types in of less populous languages.

However, when we investigate the internal behavior of the less populous language samples, we find uniformly good fits for Herdan-Heaps' law. The major difference between the more and less populous language samples is therefore not whether one follows the expectations of Herdan-Heaps' law and the other doesn't, but rather that they share adherence to the law, albeit with different parameter settings.

### 4.2.5 Robust vs. other (with downsampling over endangerment categories) {#rvallds}

The population categories in our sample are not equally frequent, and the mean populations per category are distributed in a (quasi-)log-linear fashion. These qualities could result in artificially (i.e., sample-dependent) poor fits when predicting between groups.

To address this point, we repeat the analyses above, this time down-sampling each endangerment category to have only as many languages as the smallest category in the sample.

The analyses reported in this and the following sections have the same structure as those reported in [Sections 4.2.1](#rvall) - 4.2.4.

---

- Training data are taken from the "robust" category (downsampled)

- Test data are taken from the remaining endangerment categories (downsampled)

---

```{r}
rob.down.df = pop.cross.pred(phoible_with_pop, 10, "robust", down.sample = T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=rob.down.df)

cross.pred.plot(rob.down.df, mod, title="Predicting from 'robust' to less populous languages (downsampled)\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(rob.down.df, plot.typ="heaps.pred", title = "Predicting from 'robust' to less populous languages (downsampled)\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=rob.down.df)

cross.pred.plot(rob.down.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for non-'robust' languages (downsampled)")
```

### 4.2.6 Top 25% vs. bottom 75% (with downsampling over endangerment categories)

- Training data are taken from the top 25% most populous languages (downsampled)

- Test data are taken from the remaining 75% of languages (downsampled)

---

```{r}
top.25.down.df = pop.cross.pred(phoible_with_pop, 10, pop.cut=4, down.sample = T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.25.down.df)

cross.pred.plot(top.25.down.df, mod, title="Predicting from the 25% most populous to the 75% least populous languages (downsampled)\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.25.down.df, plot.typ="heaps.pred", title = "Predicting from the 25% most populous to the 75% least populous languages (downsampled)\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.25.down.df)

cross.pred.plot(top.25.down.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 75% least populous languages (downsampled)")
```

### 4.2.7 Top 50% vs. bottom 50% (with downsampling over endangerment categories)

- Training data are taken from the top 50% most populous languages (downsampled)

- Test data are taken from the remaining 50% of languages (downsampled)

---

```{r}
top.50.down.df = pop.cross.pred(phoible_with_pop, 10, pop.cut=3, down.sample=T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.50.down.df)

cross.pred.plot(top.50.down.df, mod, title="Predicting from the 50% most populous to the 50% least populous languages (downsampled)\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.50.down.df, plot.typ="heaps.pred", title = "Predicting from the 50% most populous to the 50% least populous languages (downsampled)\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.50.down.df)

cross.pred.plot(top.50.down.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 50% least populous languages (downsampled)")
```

### 4.2.8 Top 75% vs. bottom 25% (with downsampling over endangerment categories)

- Training data are taken from the top 75% most populous languages (downsampled)

- Test data are taken from the remaining 25% of languages (downsampled)

---

```{r}
top.75.down.df = pop.cross.pred(phoible_with_pop, 10, pop.cut=2, down.sample = T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.75.down.df)

cross.pred.plot(top.75.down.df, mod, title="Predicting from the 75% most populous to the 25% least populous languages (downsampled)\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.75.down.df, plot.typ="heaps.pred", title = "Predicting from the 75% most populous to the 25% least populous languages (downsampled)\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.75.down.df)

cross.pred.plot(top.75.down.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 25% least populous languages (downsampled)")
```

### 4.2.9 Robust vs. other (with downsampling over endangerment categories; no "at-risk" languages) {#rvallds_norisk}

We further trim the data by removing the smallest category ("at-risk") languages. Downsampling is then applied based on the next smallest category ("robust").

---

- Training data are taken from the "robust" category (downsampled, no "at-risk" languages)

- Test data are taken from the remaining endangerment categories (downsampled, no "at-risk" languages)

---

```{r}
rob.down.nar.df = pop.cross.pred(phoible_with_pop %>% filter(simp.status != "at-risk"), 10, "robust", down.sample = T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=rob.down.nar.df)

cross.pred.plot(rob.down.nar.df, mod, title="Predicting from 'robust' to less populous languages (downsampled; no 'at-risk')\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(rob.down.nar.df, plot.typ="heaps.pred", title = "Predicting from 'robust' to less populous languages (downsampled; no 'at-risk')\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=rob.down.nar.df)

cross.pred.plot(rob.down.nar.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for non-'robust' languages (downsampled; no 'at-risk')")
```

### 4.2.10 Top 25% vs. bottom 75% (with downsampling over endangerment categories; no "at-risk" languages)

- Training data are taken from the top 25% most populous languages (downsampled, no "at-risk" languages)

- Test data are taken from the remaining 75% of languages (downsampled, no "at-risk" languages)

---

```{r}
top.25.down.nar.df = pop.cross.pred(phoible_with_pop %>% filter(simp.status != "at-risk"), 10, pop.cut=4, down.sample = T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.25.down.nar.df)

cross.pred.plot(top.25.down.nar.df, mod, title="Predicting from the 25% most populous to the 75% least populous languages (downsampled; no 'at-risk')\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.25.down.nar.df, plot.typ="heaps.pred", title = "Predicting from the 25% most populous to the 75% least populous languages (downsampled; no 'at-risk')\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.25.down.nar.df)

cross.pred.plot(top.25.down.nar.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 75% least populous languages (downsampled; no 'at-risk')")
```

### 4.2.11 Top 50% vs. bottom 50% (with downsampling over endangerment categories; no "at-risk" languages)

- Training data are taken from the top 50% most populous languages (downsampled, no "at-risk" languages)

- Test data are taken from the remaining 50% of languages (downsampled, no "at-risk" languages)

---

```{r}
top.50.down.nar.df = pop.cross.pred(phoible_with_pop %>% filter(simp.status != "at-risk"), 10, pop.cut=3, down.sample=T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.50.down.nar.df)

cross.pred.plot(top.50.down.nar.df, mod, title="Predicting from the 50% most populous to the 50% least populous languages (downsampled; no 'at-risk')\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.50.down.nar.df, plot.typ="heaps.pred", title = "Predicting from the 50% most populous to the 50% least populous languages (downsampled; no 'at-risk')\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.50.down.nar.df)

cross.pred.plot(top.50.down.nar.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 50% least populous languages (downsampled; no 'at-risk')")
```

### 4.2.12 Top 75% vs. bottom 25% (with downsampling over endangerment categories; no "at-risk" languages)

- Training data are taken from the top 75% most populous languages (downsampled, no "at-risk" languages)

- Test data are taken from the remaining 25% of languages (downsampled, no "at-risk" languages)

---

```{r}
top.75.down.nar.df = pop.cross.pred(phoible_with_pop %>% filter(simp.status != "at-risk"), 10, pop.cut=2, down.sample = T)
```

Explore performance of cross-prediction models

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=top.75.down.nar.df)

cross.pred.plot(top.75.down.nar.df, mod, title="Predicting from the 75% most populous to the 25% least populous languages (downsampled; no 'at-risk')\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(top.75.down.nar.df, plot.typ="heaps.pred", title = "Predicting from the 75% most populous to the 25% least populous languages (downsampled; no 'at-risk')\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=top.75.down.nar.df)

cross.pred.plot(top.75.down.nar.df, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 25% least populous languages (downsampled; no 'at-risk')")
```

### 4.2.13 Interim discussion

Down-sampling to level out differences in the number of observations per population class had little effect on the over-estimation bias (still in the range of 1.5 times) in the population-based cross-prediction models. Removing the smallest class from consideration also had little effect.

The nternal behavior of the less populous language samples remains consistent with Herdan-Heaps'.

### 4.2.14 Population-tiered walk through sample {#classXclass}

To mirror the progression from larger to smaller languages that (in aggregate) characterizes the history of documentation, we use three approaches based on categories:

---

**Approach 1**:

- randomize languages within population status categories

- order the randomized groups from largest to smallest population categories

- compute the cumulative frequencies and model them

<br><br>

**Approach 2**:

- randomly sample 75% of languages from each population status category and combine these data

- train a Herdan-Heaps' model on this sample

- predict the behavior of the remaining 25% of languages

<br><br>

**Approach 3**:

- Sample a diminishing proportion ($prop = \frac{1}{rank}$) from each category -- sorted from largest to smallest population

- train a Herdan-Heaps' model on this sample

- predict the behavior of the remaining languages

#### 4.2.14.1 Category-based walk

Walk through the full database, one population status category at a time

---

```{r}
cat.walk.df = cat.walk(phoible_with_pop, 10)
```

Explore performance of the full model (no cross-prediction)

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=cat.walk.df)

cross.pred.plot(cat.walk.df, mod, title="Accumulating languages by population status (large to small)\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(cat.walk.df, plot.typ="heaps.pred", title = "Accumulating languages by population status (large to small)\nEmpirical and predicted growth curves")
```

#### 4.2.14.2 Random sampling within population groups (75% train, 25% test)

- Sampling with 75/25 cross-prediction and category walk

---

```{r}
cat.walk.rand.75 = cross.pred.cat(phoible_with_pop, 10, .75)
```

Explore performance of the model

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=cat.walk.rand.75)

cross.pred.plot(cat.walk.rand.75, mod, title="Accumulating languages by population status (large to small) with cross-prediction\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(cat.walk.rand.75, plot.typ="heaps.pred", title = "Predicting from 75% to 25% of data (within categories)\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=cat.walk.rand.75)

cross.pred.plot(cat.walk.rand.75, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the 25% remaining languages")
```

#### 4.2.14.3 Random sampling within population groups, diminishing proportion for training per group

- Sampling $\frac{1}{rank}$ languages per category with cross-prediction and category walk

---

```{r}
cat.walk.dim = cross.pred.cat.dim(phoible_with_pop, 10)
```

Explore performance of the model

```{r}
mod = lmer(Predicted ~ Empirical + (1|Iteration), data=cat.walk.dim)

cross.pred.plot(cat.walk.dim, mod, title="Diminishing training sample by population with cross-prediction\nEmpirical vs. Predicted")
```

Plot the HH curves from training against the empirical curves

```{r}
cross.pred.plot(cat.walk.dim, plot.typ="heaps.pred", title = "Diminishing training sample by population with cross-prediction\nEmpirical and predicted growth curves")
```

Plot the HH fit within the test data

```{r}
mod = lmer(log(Empirical) ~ log(InventoryID) + (1|Iteration), data=cat.walk.dim)

cross.pred.plot(cat.walk.dim, mod, plot.typ="heaps.emp", title = "Herdan-Heaps' fit for the remaining languages")
```

#### 4.2.14.4 Interim discussion

When we allow for some degree of low-population languages from each class to be included in the training data, predictions on the test set are nearly perfect. Note that these models most closely approximate the true situation: documentation will always expand through inclusion or refinement of analyses from all (or most) population classes.

Summing up: while population does play a confounding role in estimating model parameters, all population classes follow Herdan-Heaps' law. The frequency distribution of languages across these classes has little impact on model estimates. Predicting only from higher to lower population languages results in poor fits with a tendency towards overestimation (~1.5 - 2 times the empirical values). However, enriching the training data with information from all population classes, even when shrinking the proportion of smaller population classes relative to their ranks, yields nearly perfect predictions.

# 5. General Discussion

The analyses reported in the main text raise several questions.

- First, what sources of information contribute to the observed Herdan-Heaps'-like behavior of PHOIBLE, and can these be measured against random baselines?

- Second, because the models reported in the main text train and test on the same dataset, they may suffer from overfitting (i.e., modeling sample-specific noise). Do the same results hold when training and testing on non-overlapping subsets of the data?

- Third, languages with very few speakers are hypothesized to differ systematically from languages with very many speakers. Moreover, languages with many speakers are presumably more likely to be documented earlier by linguists. Do differences in population within the languages of PHOIBLE produce different statistical patterns of behavior? Do any such differences hold under different sampling conditions?

Regarding the first point, we found that the frequency distribution of segments across languages in PHOIBLE was critical for shaping the Herdan-Heaps-like trend. The distribution of inventory sizes had virtually no effect on its own, though it did interact with the presence of segment-frequency information. These findings support the idea that the distribution of segments in PHOIBLE carries non-trivial (i.e., non-randomly-generated) information.

Regarding the second point, we found that models trained on a random 75% training set predict the behavior of the remaining 25% of languages with high accuracy. Overfitting is therefore unlikely to be a significant problem for the models reported in the main text.

Regarding the third point, we found that population size does impact the distribution of segment types: languages with larger populations behave differently as a group than languages with smaller populations. However, we also found that all groups follow Herdan-Heaps' law internally. Therefore, the difference in behavior of the groups is one of degree rather than kind -- a difference in parameter settings, all leading to non-asymptotic growth.

Taken together, these findings (a) support the internal statistical consistency of PHOIBLE, (b) eliminate concern about overfitting, (c) illuminate some of the factors that contribute to the observed growth curve, and (d) expose some consistency underlying the surface-level variation among population classes. All of these results strengthen the arguments made in the main text.

# 6. References

Catalogue of Endangered Languages. (2024). University of Hawaii at Manoa. http://www.endangeredlanguages.com

Ethnologue: Languages of the world. (2023). Dallas, Texas: SIL International,

### 2024-07-20 SI2/SI_functions.R

@@ -1,0 +1,733 @@
+# Functions for use with the Supplementary Materials for Moran & Lester (in prep)
+
+library(tidyverse)
+library(lmerTest)
+library(effects)
+
+# Randomly sample one inventory per language
+
+rand.inv = function(df){
+  set.seed(1)
+  
+  nas =  df %>%
+    distinct(InventoryID, ISO6393) %>%
+    filter(is.na(ISO6393)) %>%
+    pull(InventoryID)
+  
+  inventory_ids_sampled_one_per_isocode = df %>%
+    distinct(InventoryID, ISO6393) %>%
+    drop_na(ISO6393) %>%
+    group_by(ISO6393) %>%
+    sample_n(1) %>%
+    pull(InventoryID)
+  
+  
+  inventory_ids_sampled_one_per_isocode <- c(inventory_ids_sampled_one_per_isocode, nas)
+  
+  output = df %>%
+    filter(InventoryID %in% inventory_ids_sampled_one_per_isocode) 
+  
+  return(output)
+  
+}
+
+# Function to find the mode
+Mode <- function(x){
+  mode <- names(sort(table(inv_sizes), decreasing = T))[1]
+  return(as.numeric(mode))
+}
+
+# Function to build the inventories for a single run
+
+# df: (dataframe-like object) a table containing PHOIBLE (e.g., phoible.csv)
+# k: (integer) the number of random inventories to generate 
+# inv.inf: (boolean) should the database be built using information about inventory sizes from df (PHOIBLE)?
+# freq.inf: (boolean) should the database be built using information about relative frequency of phonemes?
+# repl: (boolean, default = FALSE) should the random sample be constructed using replacement?
+
+inventory.builder <- function(k, inv.size.types, pho.typs, inv.inf, freq.inf, repl=F, inv.size.probs=NULL, pho.probs=NULL){
+  
+  # Open a list to store the inventories
+  inv.list <- vector(mode="list", length=k)
+  
+  # Open a list to store the iteration labels
+  lab.list <- vector(mode="list", length=k)
+  
+  # Create k randomized databases
+  for(i in 1:k){
+    
+    # Define inventory size(s):
+    
+    # Sample an inventory size based on its probability
+    if(inv.inf){
+      inv.size <- as.numeric(sample(inv.size.types, 1, prob=inv.size.probs))
+    }
+    
+    # ... or simply take the mode (if inventory size info is ablated)
+    else{
[…]
